# Classify fresh frozen RNA by existing transcriptome classifiers

Let's start the good stuff. Here, we run through previously published BTC transcriptomic classifiers, using fresh frozen raw read counts from htseq, to give each tumor a class for each classifier.

## Import, filter and normalize

First, we import the data and filter for fresh frozen cases with high quality, and normalize with DESEQ.

```{r}
#devtools::install_github("peterawe/CMScaller")

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(CMScaller)
  library(DESeq2)
})

packageVersion("CMScaller")
packageVersion("DESeq2")

`%ni%` <- Negate(`%in%`)

this.seed = 555

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

#### filter for analysis ####
count_data_df <- fread(file.path(secdir,"/results/rna.htseq.all.txt"))

ncol(count_data_df) - 1

key_tumour_features <- fread(file.path( basedir,'/source_data/BTC_clinical.tumours.tsv'), header = T)
ffpe_tumours <- na.omit(key_tumour_features$Sample[key_tumour_features$`Sample Source(FFPE, FF)` == 'FFPE' ])

ffpe_tumours <- intersect(ffpe_tumours, names(count_data_df)[-1])
all_rna.ff <- dplyr::select(count_data_df, -all_of(ffpe_tumours))

ncol(all_rna.ff) - 1

good_sample_list <- fread(file.path( basedir,'/source_data/sample.list.txt'), header=F)
nrow(good_sample_list)

all_rna.ff <- dplyr::select(all_rna.ff, all_of(c('gene_name',intersect(good_sample_list$V2,names(all_rna.ff)))))

ncol(all_rna.ff) - 1

#### convert ensembl to gene symbols ####
convert.ensid = F
if(convert.ensid){
  gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))
  gene_list <- gene_set$ensid 
  names(gene_list) <- gene_set$gene_name
  
  gene_symbols <- gene_list[all_rna.ff$gene_name]
  all_rna.ff$gene_list <- gene_symbols
  
  genes <- all_rna.ff$gene_list
}

#### filter ####

count_data <- as.matrix(all_rna.ff[,-1])
rownames(count_data) <- all_rna.ff$gene_name

#remove genes without expression
count_data <- count_data[rowSums(count_data) != 0,]

saveRDS(object = count_data, file=file.path(secdir,'/results/raw_counts.ff.rds'))

# Create a dummy metadata dataframe with a single "condition" column
metadata <- data.frame(condition = factor(rep("none", ncol(count_data))),
                       row.names = colnames(count_data))

dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData = metadata,
  design = ~ 1  
)
dds <- estimateSizeFactors(dds)

normalized_counts <- counts(dds, normalized = TRUE)

saveRDS(object = normalized_counts, file = file.path(secdir,'/results/normalized_counts.ff.rds'))

scaled_counts = CMScaller::ematAdjust(normalized_counts, normMethod = "scale")

```

## Start Classifying

Classifier by classifier, we run classification using nearest template prediction (ntp) from the CMScaller package, keeping assignment for cases where assignment false discovery rate (FDR) is less than 0.05.

### Andersen et al. 2012
From [Supplementary Table 1](http://dx.doi.org/10.1053/j.gastro.2011.12.005)	

```{r}

andersen.classifier <- fread(file.path(basedir,'/source_data/signatures/andersen_supp.txt'), 
                             col.names = c('order','probe','p','cv','geom_mean_1','geom_mean_2','ratio','loc','class'), header=T)
andersen.res = ntp(scaled_counts, andersen.classifier, doPlot=T, nPerm=1000, seed=this.seed)
andersen.res <- andersen.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(andersen.res) <- paste0('andersen.',names(andersen.res) )

```

### Sia et al. 2013
From [Supplementary Table 2](http://dx.doi.org/10.1053/j.gastro.2013.01.001)
```{r}

sia.classifier <- fread(file.path(basedir,'/source_data/signatures/sia_supp.txt'), col.names = c('probe','class','effectSize'), header=F)
sia.res = ntp(scaled_counts, sia.classifier, doPlot=T, nPerm=1000, seed=this.seed)
sia.res <- sia.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(sia.res) <- paste0('sia.',names(sia.res) )

```

### Ahn et al. 2019
From [Supplementary Table 1](http://dx.doi.org/10.1007/s12072-019-09954-3)
```{r}
ahn.classifier <- fread(file.path(basedir,'/source_data/signatures/ahn_supp.txt'), col.names = c('probe','class'), header=F)
ahn.res = ntp(scaled_counts, ahn.classifier, doPlot=T, nPerm=1000, seed=this.seed)
ahn.res <- ahn.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(ahn.res) <- paste0('ahn.',names(ahn.res) )
```

### Montal et al. 2020
From [Table S11](http://dx.doi.org/10.1016/j.jhep.2020.03.008)	
```{r}
montal.classifier <- fread(file.path(basedir,'/source_data/signatures/montal_supp.txt'), col.names = c('probe','class'), header=F)
montal.res = ntp(scaled_counts, montal.classifier, doPlot=T, nPerm=1000, seed=this.seed)
montal.res <- montal.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(montal.res) <- paste0('montal.',names(montal.res) )
```

### Nepal et al. 2021
From [Table S6](http://dx.doi.org/10.1016/j.jhep.2020.11.033)	
```{r}
nepal.classifier <- fread(file.path(basedir,'/source_data/signatures/nepal_supp.txt'), col.names = c('probe','HR','p','survival','diff_1','diff_3','diff_2','class'), header=T)
nepal.classifier <- nepal.classifier %>% filter(class != '')
nepal.res = ntp(scaled_counts, nepal.classifier, doPlot=T, nPerm=1000, seed=this.seed)
nepal.res <- nepal.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(nepal.res) <- paste0('nepal.',names(nepal.res) )
```

### O’Rourke et al. 2024
From [Table S2](http://dx.doi.org/10.1136/gutjnl-2023-330748)	

```{r}
orourke.rpls.classifier <- fread(file.path(basedir,'/source_data/signatures/orourke_supp.txt'), col.names = c('probe','class'), header=F)
orourke.rpls.res = ntp(scaled_counts, orourke.rpls.classifier, doPlot=T, nPerm=1000, seed=this.seed)
orourke.rpls.res <- orourke.rpls.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(orourke.rpls.res) <- paste0('orourke.',names(orourke.rpls.res) )
```

### Vijay et al. 2025
From [Table S5 | RNA_cluster_signature](http://dx.doi.org/10.1101/2024.07.04.601970) (bioRxiv)

```{r}
vijay.classifier <- fread(file.path(basedir,'/source_data/signatures/vijay_supp.txt'), col.names = c('r','probe','p','effectSize','q','class'), header=T)
vijay.res = ntp(scaled_counts, vijay.classifier, doPlot=T, nPerm=1000, seed=this.seed)
vijay.res <- vijay.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(vijay.res) <- paste0('vijay.',names(vijay.res) )
```

### Martin-Serrano et al. 2023
From [supplementary](http://dx.doi.org/10.1136/gutjnl-2021-326514)

```{r}
martinserrano.classifier <- fread(file.path(basedir,'/source_data/signatures/martinserrano_supp.txt'), col.names = c('probe','class'), header=F)
martinserrano.res = ntp(scaled_counts, martinserrano.classifier, doPlot=T, nPerm=1000, seed=this.seed)
martinserrano.res <- martinserrano.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(martinserrano.res) <- paste0('martinserrano.',names(martinserrano.res) )
```

### Song G.  et al. 2022
From [Supplementary Data 5](http://dx.doi.org/10.1038/s41467-022-29164-0)	

```{r}
song.classifier <- fread(file.path(basedir,'/source_data/signatures/song_supp.txt'), skip = 1)
names(song.classifier)[names(song.classifier) == 'Gene'] <- 'probe'
song.classifier$class <- NA
song.classifier$class[song.classifier$logFC < -1.5 & song.classifier$`P-value` < 0.05] <- 'spp1_pos'
song.classifier$class[song.classifier$logFC > 1.5 & song.classifier$`P-value` < 0.05] <- 's100p_pos'
song.classifier <- song.classifier %>% filter(!is.na(class)) %>% dplyr::select(probe, class)
song.res = ntp(scaled_counts, song.classifier, doPlot=T, nPerm=1000, seed=this.seed)
song.res <- song.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(song.res) <- paste0('song.',names(song.res) )
```

### Fan et al. 2024
From [Supplementary Figure 7](http://dx.doi.org/10.1038/s41467-024-44748-8)	

```{r}
fan.classifier <- fread(file.path(basedir,'/source_data/signatures/fan_supp.txt'), col.names = c('probe','class'), header=F)
fan.res = ntp(scaled_counts, fan.classifier, doPlot=T, nPerm=1000, seed=this.seed)
fan.res <- fan.res %>% mutate(class = ifelse(FDR < 0.05, as.character(prediction), NA))
names(fan.res) <- paste0('fan.',names(fan.res) )

```

## Save Results

Then all results are combined and saved from later.

```{r}
all.res <- cbind.data.frame(
  fan.res,
  sia.res,
  andersen.res,
  nepal.res,
  song.res,
  montal.res,
  martinserrano.res,
  vijay.res,
  orourke.rpls.res,
  ahn.res
)

all.res <- cbind.data.frame("Sample"=rownames(all.res), all.res)

write.table(all.res, file = paste0(file.path(basedir,'/results/','classified.FF.txt')), sep = '\t',row.names = F,col.names = T,quote = F)

```


