# Use eCCS to train gCCS
We quickly noticed that we'd want a genetic based classifier to capture the eCCS, if possible. The main target was panel data so to give us  access to the most different panels, we settle on using average arm-level relative (T/N) copy. 

The problem we ran into was that the model didn't translate well at first to TCGA and ICGC cohorts, and we were able to diagnose the issue that these cohorts had lower tumor content (compared to our LCM cohort) so the weights for each arm in the glm were too small. To deal with this, we decided to directly add in a proxy for tumor content. We first played with tumor content estimated from optimization of the ploidy / cellularity trade-off directly from the CNV data, but most datasets didn't offer this value; a much more accessible proxy was the median variant allele frequency of tumor variants. Although technically not as precise, we found in practice this did a good enough job.

## Start with variance in arm-level copies
To first decide which chromosome arms to include in the model, we run a PCA on arm-level copies and then keep the top components (rather than a model with 44 parameters).

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(tidyr)
  library(dplyr)
  library(stringr)
  library(tibble)
  library(CNTools)
  library(FactoMineR)
  library(pROC)
  library(ggplot2)
  library(gghalves)
})

`%ni%` <- Negate(`%in%`)

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

source( file.path(basedir, "bin/BTC.functions.R"))

# get genes
gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))
gene_set$Chr <- paste0('chr',gene_set$Chr)

genebed <- gene_set
genebed <- genebed %>% filter(biotype=="protein_coding" & Length >= 500 & Chr %ni% c('chr24','chr25')) %>%
  dplyr::select(Chr, Start,   End,ensid, gene_name) 

names(genebed) <- c('chrom',	'start',	'end',	'geneid',	'genename')

all.seg <- fread(file.path(basedir,'/results/all.seg.txt'), col.names = c('ID','chrom','loc.start','loc.end','num.mark','seg.mean'))
all.seg <- all.seg %>% filter(!is.na(seg.mean))
length(unique(all.seg$ID))

segmented.data <- CNSeg(all.seg)
segment.gene <- getRS(segmented.data, by="gene", imput=FALSE, XY=FALSE, geneMap=genebed, what="min")
segment.gene <- rs(segment.gene)

seg.mat <- as.matrix(segment.gene[,-c(1:5)])
row.names(seg.mat) <- segment.gene$genename

predicted.w.score <- fread(file.path(basedir,'/results/all.eCMS.txt') )
predicted.w.score <- predicted.w.score %>% filter(`Sample Source(FFPE, FF)` == 'FF' & !is.na(rna_class))

predicted.w.score$predicted_label <- NA
predicted.w.score$predicted_label[predicted.w.score$rna_class == 'CMS-A'] <- 1
predicted.w.score$predicted_label[predicted.w.score$rna_class == 'CMS-B'] <- 0

## group by cytoband ##

cytoBand <- fread(file.path(basedir,'/source_data/cytoBand.txt'), na.strings = '') %>% filter(!is.na(Chromosome))

cytoBand<- separate(cytoBand, col = Chromosome, into = c('band','period'), sep = '\\.', remove=F, extra = 'drop', fill = 'right')

cytoBand <- cytoBand %>%
  mutate(
    parts = str_match(band, "^(.*?[pq])(.*)$"),
    arm = parts[,2],
    armband = parts[,3]
  ) %>%
  dplyr::select(-parts) %>% filter(!is.na(arm) & arm %ni% c('Xq','Xp','Yp','Yq'))

segment.gene.cyto <- inner_join(cytoBand, segment.gene,  by=c('Hugo_Symbol'='genename'), relationship = "many-to-many")


arm.med <- segment.gene.cyto %>%
  group_by(arm ) %>%  
  summarise( across(11:ncol(segment.gene.cyto)-1, \(x) median(x, na.rm = TRUE)) ) 



arm.pca <- PCA(t(arm.med[,-1]), graph = F)

allOfit_pca_labs <- cbind.data.frame(sample_id=names(arm.med[,-1]),
                                     
                                     arm.pca$ind$coord
)

allOfit_pca_labs <- left_join(allOfit_pca_labs, predicted.w.score, by=c('sample_id'='sample')) %>%
  filter(!is.na(rna_class))


ggplot(allOfit_pca_labs, aes(x=Dim.1, y=Dim.2, color=rna_class)) + 
  stat_ellipse(level=0.95, alpha=0.7)+
  geom_point(size=1) + 
  theme_bw(base_size=10) + 
  
  scale_color_manual(values=c('CMS-A'="#04597C", 'CMS-B'="#EDA71A"), na.value = 'grey80') +
  theme(panel.grid = element_blank())

arm.pca.contrib <- cbind.data.frame(sample_id=arm.med[,1],  arm.pca$var$contrib) %>% arrange(desc(Dim.1))
arm.pca.contrib[1:10,]
```

## Add in variant allele frequency to account for tumor content
We show that 2 * mean VAF (assuming that most tumor variants are heterozygous) still under-estimates cellularity, but it is systematic enough that it works as a proxy.

```{r}

# Transpose while preserving numeric types
transposed.raw <- arm.med %>%
  column_to_rownames(var = names(arm.med)[1]) %>%  # Use first column as rownames
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>%
  mutate(across(-sample, as.numeric)) %>%  # Ensure numeric columns
  as_tibble() 

names(transposed.raw)[-1] <- paste0('arm.',names(transposed.raw)[-1])

#### get median ####
 
pull.median=F
if(pull.median){
  
  sample_list <- read.table(file.path(basedir,'/source_data/sample.list.txt'), header=F)
  
  first.tumor=T
  for(tumour_id in sample_list$V2){
    cat('processing ',tumour_id,'\n')
    donor_id <- sample_list$V1[sample_list$V2 == tumour_id]
    
    evol.file <- paste0('/Volumes/pcsi/users/fbeaudry/more.data/',tumour_id,'/pyclone/',tumour_id,'.snv.depths.txt')
    info.file <- paste0('/Volumes/pcsi/users/fbeaudry/more.data/',tumour_id,'/pyclone/',tumour_id,'.snv.info.txt')
    
    if(all(file.exists(evol.file) & file.exists(info.file) )){
      
      cat('file found\n')
      evol.data <- fread(evol.file)
      info.data <- fread(info.file) 
      evol.data <-  evol.data %>% filter(mutation_id %in% info.data$mutation_id[info.data$ANNOVAR_EXONIC != 'False'])
      
      vaf.df <- cbind.data.frame('tumor'= tumour_id,
                                 'median.vaf' = median( evol.data$alt_counts / (evol.data$alt_counts + evol.data$ref_counts) , na.rm = T),
                                 'mean.vaf' = mean( evol.data$alt_counts / (evol.data$alt_counts + evol.data$ref_counts) , na.rm = T)
      )
      
      if(first.tumor){all.vaf.df <- vaf.df; first.tumor =F }else {all.vaf.df <- bind_rows(all.vaf.df, vaf.df)}
      
    } else {
      cat('!!file not found\n')
      
    }
  }
  
  write.table(
    all.vaf.df,
    file = file.path(basedir,'/results/median.vaf.txt'),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
} else {
  all.vaf.df <- fread(file.path(basedir,'/results/median.vaf.txt') )
}

cellularity <- fread(file.path(basedir, "/results/btc.cellularity.curated.txt"), 
                     col.names = c('tumor','cellularity','ploidy'))
all.vaf.df<- full_join(cellularity, all.vaf.df, by=c('tumor'='tumor'))

cor.test(all.vaf.df$cellularity, 2*all.vaf.df$median.vaf)
cor.test(all.vaf.df$cellularity, 2*all.vaf.df$mean.vaf)

ggplot(all.vaf.df, aes(x=cellularity, y=2*mean.vaf)) + 
  geom_point() + theme_bw() + geom_abline(slope = 1)

transposed.raw <- inner_join(all.vaf.df, transposed.raw, by=c('tumor'='sample'))

```

## train generalized linear model (glm)

```{r}

transposed <- inner_join(predicted.w.score, transposed.raw, by=c('sample'='tumor'))


refind.feature = F
retrain.model = F

if(refind.feature){
  
  
  predictor_vars <- paste0('arm.',arm.pca.contrib$arm[1:10])  # Extract the list of predictors
  
  formula_obj <- reformulate(termlabels = c(predictor_vars, 'cellularity'), response = "predicted_label")
  
  full_model <- glm(formula_obj, data = transposed , family = binomial)
  
  reduced_model <- step(full_model, direction = "both")
  
  reduced_model_summary <- summary(reduced_model)
  print(reduced_model_summary$coefficients)
  
  
}else if(retrain.model){
  
  train.set <- sample(transposed$sample, size=length(transposed$sample) * 0.8, replace = F)
  test.set <- transposed$sample[transposed$sample %ni% train.set]

  full_model <- glm(predicted_label ~ arm.1q + arm.4p + arm.6q + arm.9q + arm.13q + arm.18q + arm.19p + mean.vaf, 
                    data = transposed %>% filter(sample %in% train.set), 
                    family = binomial)
  

  test.data <- transposed %>% filter(sample %in% test.set)
  
  test.data$prob <- predict(full_model, transposed %>% filter(sample %in% test.set), type = "response")
  
  roc_obj <- roc(test.data$predicted_label, test.data$prob)
  
  cutoff <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"))
  
  test.data$glm_labels <- ifelse(test.data$prob > cutoff$threshold, 1, 0)
  table(test.data[,c('predicted_label', 'glm_labels')])
  
  saveRDS(full_model, file = file.path(basedir,'/results/cnv.model.rds'))

  model.date <- format(Sys.time(), "%y%m%d")
  saveRDS(full_model, file = paste0(basedir,'/results/cnv.model.',model.date,'rds'))
  
  cna.training.samples <- list('training_set'=train.set, 'test_set'=test.set)
  saveRDS(cna.training.samples, file.path(basedir,'/results/cna.training.samples.rds'))
  
  write.table(
    data.frame(full_model$coefficients),
    file = file.path(basedir,'/results/gCMS.weights.txt'),
    row.names = T, quote = FALSE, sep = "\t"
  )
  
}else{
  
  full_model <- readRDS(file.path(basedir,'/results/cnv.model.rds'))
   
   
  cna.training.samples <- readRDS( file.path(basedir,'/results/cna.training.samples.rds'))
  test.set <- cna.training.samples[['test_set']]
  
  
  
  test.data <- transposed %>% filter(sample %in% test.set)
  test.data$prob <- predict(full_model, transposed %>% filter(sample %in% test.set), type = "response")
  
  roc_obj <- roc(test.data$predicted_label, test.data$prob)
  
  cutoff <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"))
  
  test.data$glm_labels <- ifelse(test.data$prob > cutoff$threshold, 1, 0)
  table(test.data[,c('predicted_label', 'glm_labels')])
  
}

length(test.set)
summary(full_model)
cutoff
auc(roc_obj)
plot(roc_obj)

cnv_prediction.test_set.plot <- 
      ggplot(test.data %>% filter(!is.na(rna_class)), aes(x=reorder(sample, prob), y=prob)) + 
         geom_hline(yintercept = 0.34, color='grey80', linetype='dotted')+
        
      #  geom_point(aes(color=factor(driver.class)),  shape=16, size=1, stroke=0.5) + 
        geom_point(aes(color=factor(rna_class)),   shape=1, size=2, stroke=1) + 
        
        theme_bw(base_size=10) +
        scale_color_manual(values = c('BAP1+'='#04597C',
                                      'TP53+'='#EDA71A',
                                      'CMS-A'="#04597C",
                                      'CMS-B'="#EDA71A"
        ), 
        na.value = 'grey80')  +
        
        guides(color='none',fill='none')+
         theme(panel.grid = element_blank(),
               panel.background = element_blank(), 
               plot.background = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks.x = element_blank()
               #axis.text.x = element_text(angle = 45, hjust = 1, size=4)
               ) + 
         labs(x='Test Samples (Sorted)', y='CNA-Based Probability of CCS-A',color='') +
         scale_y_continuous(breaks = c(0,0.34,1))
       
  cnv_prediction.test_set.plot

ggsave(filename=paste0(basedir, "/results/fig/FIG2C.cnv_prediction.test_set.svg"), plot = 
         cnv_prediction.test_set.plot
      ,   device = "svg",width = 72, height = 58, units = 'mm', dpi=72 )


```

## validate in external data
We test the model using copy number and expression data generated in [Farshidfar et al. 2017](http://dx.doi.org/10.1016/j.celrep.2017.02.033) (which you can  download from [cbioportal](https://www.cbioportal.org/study/summary?id=chol_tcga_gdc)). We also add in TP53/BAP1 mutations to the plot, although in the paper narrative we haven't established how these segregate differently by CCS yet.

```{r}

far.cn <- fread(file.path(secdir,'/source_data/cbioportal/chol_tcga_gdc/data_cna_hg38.seg'))
far.cn$chrom <- paste0('chr',far.cn$chrom)

far.cn <- far.cn %>% filter(!is.na(seg.mean))

far.segmented.data <- CNSeg(far.cn)
far.segment.gene <- getRS(far.segmented.data, by="gene", imput=FALSE, XY=FALSE, geneMap=genebed, what="min")
far.segment.gene <- rs(far.segment.gene)

far.segment.gene.cyto <- inner_join(cytoBand, far.segment.gene,  by=c('Hugo_Symbol'='genename'), relationship = "many-to-many")

far.arm.med <- far.segment.gene.cyto %>%
  group_by(arm ) %>% 
  summarise(across(11:ncol(far.segment.gene.cyto)-1, \(x) median(x, na.rm = TRUE)))

far.transposed.raw <- far.arm.med %>%
  column_to_rownames(var = names(far.arm.med)[1]) %>%  # Use first column as rownames
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>%
  mutate(across(-sample, as.numeric)) %>%  # Ensure numeric columns
  as_tibble()

names(far.transposed.raw)[-1] <- paste0('arm.',names(far.transposed.raw)[-1])

# add VAF
far.mut <- fread(file.path(secdir,'/source_data/cbioportal/chol_tcga_gdc/data_mutations.txt'))
far.mut$mean.vaf <- far.mut$t_alt_count / (far.mut$t_ref_count + far.mut$t_alt_count)

far.med.vaf <- far.mut %>%
  group_by(Tumor_Sample_Barcode ) %>% 
  summarise(across(mean.vaf, \(x) mean(x, na.rm = TRUE)))

ggplot(far.med.vaf, aes(x=mean.vaf * 2)) + geom_histogram() + theme_bw() + xlim(0,1)

far.transposed.raw <- inner_join(far.med.vaf, far.transposed.raw, by=c('Tumor_Sample_Barcode'='sample'))

# predict
far.transposed.raw$prob <- predict(full_model, far.transposed.raw , type = "response")

## compare to RNA class
far.mrna <- fread(file.path(secdir,'/source_data/cbioportal/chol_tcga_gdc/data_mrna_seq_tpm.txt'))
gene_info <- fread(file.path(secdir,'/source_data/cbioportal/gene_info.txt'))
far.mrna <- inner_join(gene_info, far.mrna, by=c('ENTREZID'='Entrez_Gene_Id'))

far.mrna.mat <- as.matrix(far.mrna[,-c(1:2)])
rownames(far.mrna.mat) <- far.mrna$Hugo_Symbol

classifier <- readRDS(file.path(basedir,'/results/tps.classifier.rds'))
far.predicted.w.score <- predict_TSP_with_confidence(tsp_classifier=classifier, tpm_matrix=far.mrna.mat, confidence_cutoff = 0.8 )

far.predicted.w.score$Tumor_Sample_Barcode <- rownames(far.predicted.w.score)

## pull common muts

far.driver <- far.mut %>% filter(Hugo_Symbol %in% c('BAP1','FGFR2','IDH1','TP53','SMAD4')) %>% 
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol) %>% 
  dcast(Tumor_Sample_Barcode ~ Hugo_Symbol, value.var = 'Hugo_Symbol')

far.driver[is.na(far.driver)] <- 0

far.driver.cms.cnv <- full_join(far.driver,far.predicted.w.score, by=c('Tumor_Sample_Barcode'='Tumor_Sample_Barcode') ) %>% 
  full_join(far.transposed.raw, by=c('Tumor_Sample_Barcode'='Tumor_Sample_Barcode')) 

far.driver.cms.cnv$sample <- factor(far.driver.cms.cnv$Tumor_Sample_Barcode, 
                                    levels=far.driver.cms.cnv$Tumor_Sample_Barcode[order(far.driver.cms.cnv$prob)])

far.roc_obj <- roc(far.driver.cms.cnv$predicted_label, far.driver.cms.cnv$prob)
auc(far.roc_obj)

plot(far.roc_obj)

coords(far.roc_obj, 0.34, ret = c("threshold", "sensitivity", "specificity"))


far.driver.cms.cnv$glm_labels <- ifelse(far.driver.cms.cnv$prob > cutoff$threshold, 'gCMS-A', 'gCMS-B')

far.driver.cms.cnv$driver.class <- NA
far.driver.cms.cnv$driver.class[far.driver.cms.cnv$BAP1 == 'BAP1'] <- 'BAP1+'
far.driver.cms.cnv$driver.class[far.driver.cms.cnv$TP53 == 'TP53'] <- 'TP53+'

cnv_prediction.farshidfar.plot <- 
ggplot(far.driver.cms.cnv   %>% filter(!is.na(prob)), aes(x=reorder(Tumor_Sample_Barcode, prob), y=prob)) + 
  
  geom_hline(yintercept = 0.34, color='grey', linetype='longdash')+
  
  geom_point(aes(size=mean.vaf*0.5, color=factor(driver.class)),  shape=16,  stroke=0.5) + 
  geom_point(aes( size=mean.vaf,color=factor(rna_class)),   shape=1, stroke=1) + 
  scale_size(range = c(0.1,1.5))  +
  
  theme_bw(base_size=10) +
  scale_color_manual(values = c('BAP1+'='#04597C',
                                'TP53+'='#EDA71A',
                                'CMS-A'="#04597C",
                                'CMS-B'="#EDA71A"
  ), 
  na.value = 'grey80')  +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
        legend.key.size = unit(0.4, "cm"),   # Shrinks the legend keys (symbols)
        legend.text = element_text(size = 8), # Shrinks the text for legend items
        legend.title = element_text(size = 9) ,# Shrinks the legend title
        
        panel.background = element_blank(), 
        plot.background = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) + 
  ylim(0,1)+
  labs(x='', y='CNA-Based Probability of CMS-A',color='', size='mean VAF' )
cnv_prediction.farshidfar.plot


write.table(
  far.driver.cms.cnv %>% dplyr::select(Tumor_Sample_Barcode, rna_class, confidence, glm_labels, prob, driver.class),
  file = file.path(basedir,'/results/tcga.cms.txt'),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

```
