# Evaluate OS difference between CCS 
We use three external cohorts

## Load Functions and Packages

```{r}

suppressPackageStartupMessages({
  library(readxl)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(ggplot2)
  library(ggrepel)
  library(PNWColors)
  library(scales)
  library(survival)
  library(survminer)
  library(survMisc)
  library(vegan)
  library(lumi)
  library(minfi)
  library(limma)
  library(FactoMineR)
  library(sva)
  library(illuminaHumanv4.db)
  library(CNTools)
  library(pROC)
})

`%ni%` <- Negate(`%in%`)

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

source( file.path(basedir, "bin/BTC.functions.R"))
classifier <- readRDS(file.path(basedir,'/results/tps.classifier.rds'))

```

## Data from Dong et al 2022
Data can be downloaded in supplementary materials from [Dong et al. 2022](https://doi.org/10.1016/j.ccell.2021.12.006). We also look here at the effect of accounting for TP53+ in OS.

```{r}

dong.patient.info <- read_excel(file.path(secdir,"/source_data/dong_2022_S1.xlsx"), sheet = 2, skip = 1)
dong.exome.info <- read_excel(file.path(secdir,"/source_data/dong_2022_S1.xlsx"), sheet = 3, skip = 1)
dong.rna.info <- read_excel(file.path(secdir,"/source_data/dong_2022_S1.xlsx"), sheet = 4, skip = 1)

colnames(dong.exome.info)   <- gsub(" ", "_", gsub("[^[:alnum:] ]", "", colnames(dong.exome.info)))                
dong.exome.info$PatientID <- paste0('DONG_',dong.exome.info$SampleID)

colnames(dong.patient.info)   <- gsub(" ", "_", gsub("[^[:alnum:] ]", "", colnames(dong.patient.info)))                
dong.patient.info$PatientID <- paste0('DONG_',dong.patient.info$PatientID)

names(dong.rna.info)[1] <- 'genes'
names(dong.rna.info)[-1] <- paste0('DONG_',names(dong.rna.info[,-1]))
dong.rna.mat <- as.matrix(dong.rna.info[,-1])
row.names(dong.rna.mat) <- dong.rna.info$genes

dong.predicted.w.score <- 
  predict_TSP_with_confidence(tpm_matrix=dong.rna.mat, tsp_classifier=classifier, confidence_cutoff = 0.75  )

dong.predicted.w.score$PatientID <- rownames(dong.predicted.w.score)

dong.vaf <- dong.exome.info %>% group_by(PatientID) %>% summarise(mean = mean(Variant_allele_frequency_VAF))

dong.patient.info <- full_join(dong.patient.info, dong.predicted.w.score, by=c('PatientID'='PatientID'))
dong.patient.info <- full_join(dong.patient.info, dong.vaf, by=c('PatientID'='PatientID'))

dong.unresectable <- dong.patient.info %>% filter(!is.na(OSevent) ) 
nrow(dong.unresectable)

unique(dong.unresectable$TNM_stage)
dong.unresectable$TNM_stage.bin <- dong.unresectable$TNM_stage
dong.unresectable$TNM_stage.bin[dong.unresectable$TNM_stage %in% c('IA','IB')] <- 'I'
dong.unresectable$TNM_stage.bin[dong.unresectable$TNM_stage %in% c('IIIA','IIIB')] <- 'III'

write.table(
  dong.predicted.w.score,
  file = file.path(basedir, "/results/dong.eCMS.txt"),
  row.names = T, quote = FALSE, sep = "\t"
)

dong.os.cms <- survfit(Surv(OS_overall_survival_day/30, OSevent) ~ rna_class , 
                             data = dong.unresectable )

cox.os.cms.dong <- coxph(Surv(OS_overall_survival_day, OSevent) ~ rna_class + TNM_stage.bin, 
                         data =  dong.unresectable )

cox.os.cms.dong

os.cms.dong.plot <- ggsurvplot(dong.os.cms, data = dong.unresectable, 
           pval = F, 
           conf.int = T,
           #xlim = c(0,100),
           legend="none", xlab="Months after surgery",  
                                     risk.table = F,  # Adds number at risk table
                                    # surv.median.line = "hv", 
                                     palette =  c("#04597C","#EDA71A")
)

os.cms.dong.plot.g <- os.cms.dong.plot$plot
os.cms.dong.plot.g <- 
  os.cms.dong.plot.g + 
  theme_bw(base_size=10) + guides(color='none',fill='none') +
  theme(panel.grid = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()) +
  scale_x_continuous(breaks = seq(0, max(dong.unresectable$OS_overall_survival_day/30), by = 12))

os.cms.dong.plot.g

plot.it = T
if(plot.it){
  
  ggsave(paste0(basedir, "/results/fig/FIGS2F.cms.os.dong.svg"),plot =os.cms.dong.plot.g
         , device = "svg",width = 67, height = 65, units = 'mm' )
}

tp53_cases <- dong.exome.info$PatientID[ dong.exome.info$Gene == 'TP53']

dong.tp53wt <- dong.unresectable %>% filter(PatientID %ni% tp53_cases)

dong.os.cms.tp53wt <- survfit(Surv(OS_overall_survival_day/30, OSevent) ~ rna_class , 
                       data = dong.tp53wt )

dong.unresectable$tp53 <- 'wt'
dong.unresectable$tp53[dong.unresectable$PatientID %in% tp53_cases] <- 'TP53+'

cox.os.cms.dong.tp53 <- coxph(Surv(OS_overall_survival_day, OSevent) ~ rna_class + TNM_stage.bin + tp53, 
                         data =  dong.unresectable )

dong.os.cms.tp53 <- survfit(Surv(OS_overall_survival_day/30, OSevent) ~ rna_class + tp53, 
                              data = dong.unresectable )

p <- ggsurvplot(
  dong.os.cms.tp53,
  data = dong.unresectable,
  conf.int = F,
  risk.table = T,
  pval = FALSE,
  censor = F,
  color      = "rna_class", 
  linetype = "tp53"  
)

p$plot <- p$plot +
  scale_color_manual(values = c(
    "CMS-A" = "#04597C",
    "CMS-B" = "#EDA71A"
  )) +
  scale_linetype_manual(values = c("wt" = "dashed", "TP53+" = "solid"))

p.plot <- p$plot
p.plot.g <- 
  p.plot + 
  theme_bw(base_size=10) + 
  guides(color='none',fill='none', linetype='none') +
  theme(panel.grid = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()) +
  scale_x_continuous(breaks = seq(0, max(dong.unresectable$OS_overall_survival_day/30), by = 12))

p.plot.g
ggsave(paste0(basedir, "/results/fig/FIGS4F.cms.os.dong.tp53.svg"),plot =p.plot.g
       , device = "svg",width = 67, height = 75, units = 'mm' )

```

## Data from Jusakul et al 2017
Clinical data from [cbioportal](https://www.cbioportal.org/study/summary?id=chol_icgc_2017). 
Illumina HumanHT-12 V4.0 expression beadchip data from [geo](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE89747).


```{r}

jusakul_sample <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_sample.txt'), skip=4)
jusakul_patient <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_patient.txt'), skip=4)

jusakul_patient <- full_join(jusakul_patient, jusakul_sample, by=c('PATIENT_ID'='PATIENT_ID'))

jusakul_expression_1 <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/GSE89747_CCA_batch01_illumina_Gene_expression_noNorm_noBKGD.txt'))
jusakul_expression_2 <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/GSE89748_CCA_batch02_illumina_Gene_expression_noNorm_noBKGD.txt'))
  
clean_illumina_array <- function(rna.raw){
  expr_cols <- seq(2, ncol(rna.raw), by = 2)
  pval_cols <- seq(3, ncol(rna.raw), by = 2)
  
  pval_data <- rna.raw[, ..pval_cols]
  pval_data$probes <- rna.raw$ID_REF
  bad_probes <- pval_data$probes[rowSums(pval_data > 0.05)  >= (ncol(pval_data) * 0.8 ) ]
  
  rna.info <- rna.raw[, ..expr_cols]
  rna.info$probes <- rna.raw$ID_REF
  
  rna.info <- rna.info %>% filter(probes %ni% bad_probes)
  
  rna.mat <- apply(as.matrix(rna.info[,-c('probes')]), 2, as.numeric)
  rownames(rna.mat) <- rna.info$probes
  
  # Convert your matrix to an ExpressionSet first
  rna.eset <- new("ExpressionSet", exprs = rna.mat)
  
  # Then convert to LumiBatch
  rna.lumi <- as(rna.eset, "LumiBatch")
  
  lumi_norm <- lumiExpresso(rna.lumi,
                            normalize.param = list(method = "quantile"),
                            varianceStabilize.param = list(method = "vst"))
  
  return(lumi_norm)
}

jusakul_expression_1.clean <- clean_illumina_array(rna.raw=jusakul_expression_1)
jusakul_expression_2.clean <- clean_illumina_array(rna.raw=jusakul_expression_2)

common_probes <- intersect(rownames(exprs(jusakul_expression_1.clean)), rownames(exprs(jusakul_expression_2.clean)))

exprs1_common <- exprs(jusakul_expression_1.clean)[common_probes, ]
exprs2_common <- exprs(jusakul_expression_2.clean)[common_probes, ]

combined_exprs <- cbind(exprs1_common, exprs2_common)

batch <- c(rep(1, ncol(exprs1_common)), rep(2, ncol(exprs2_common)))
corrected_exprs <- ComBat(dat = combined_exprs, batch = batch)

jusakul.pca <- PCA(t(corrected_exprs), graph = F)
jusakul.sample <- jusakul.pca$ind$coord

corrected_exprs_gened <- data.frame(corrected_exprs)
gene_symbols <- unlist(mget(rownames(corrected_exprs), illuminaHumanv4SYMBOL, ifnotfound = NA))
corrected_exprs_gened$gene_symbols <- gene_symbols

jusakul.collapsed_df <- corrected_exprs_gened %>%
  group_by(gene_symbols) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) %>%
  ungroup()

jusakul.collapsed.mat <- as.matrix(jusakul.collapsed_df[,-1])
rownames(jusakul.collapsed.mat) <- jusakul.collapsed_df$gene_symbols

jusakul.predicted.w.score <- 
  predict_TSP_with_confidence(tpm_matrix=jusakul.collapsed.mat, 
                              tsp_classifier=classifier, confidence_cutoff = 0.75 )


write.table(
  jusakul.predicted.w.score,
  file = file.path(basedir, "/results/jusakul.eCMS.txt"),
  row.names = T, quote = FALSE, sep = "\t"
)


jusakul.sample.scored <- cbind.data.frame(jusakul.sample, jusakul.predicted.w.score)


jusakul.sample.scored$PATIENT_ID <- rownames(jusakul.sample.scored)

jusakul.sample.scored <- left_join(jusakul.sample.scored, jusakul_patient, by=c('PATIENT_ID'='PATIENT_ID'))


ggplot(jusakul.sample.scored,aes(x=Dim.1, y=Dim.3)) +
  
  stat_ellipse(aes(color=rna_class), alpha=0.5, level = 0.75)+
  
  geom_point(size=1, aes(color=rna_class) ) + theme_bw() +

  scale_color_manual(values=c("#04597C", "#EDA71A","#04597C", "#EDA71A"), na.value = 'grey90') + 
  theme(panel.grid = element_blank())


jusakul_cms <- fread(file.path(basedir,'results/jusakul.eCMS.txt'))
jusakul_patient <- full_join(jusakul_patient, jusakul_cms, by=c('PATIENT_ID'='V1'))


jusakul_muts <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_mutations.txt'))


jusakul_drivers <- jusakul_muts %>% filter(Hugo_Symbol %in% c('BAP1','TP53','FGFR2','IDH1','SMAD4')) %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, Consequence)

jusakul_dcast <- dcast(jusakul_drivers,Tumor_Sample_Barcode~ Hugo_Symbol, value.var = 'Consequence', fun.aggregate = length)
jusakul_dcast$gCMS <- NA
jusakul_dcast$gCMS[(jusakul_dcast$BAP1 == 1 | jusakul_dcast$IDH1  == 1) & (jusakul_dcast$SMAD4 == 0 & jusakul_dcast$TP53 == 0 )] <- 'gCMS-A'
jusakul_dcast$gCMS[(jusakul_dcast$SMAD4 == 1 | jusakul_dcast$TP53  == 1) & (jusakul_dcast$BAP1 == 0 & jusakul_dcast$IDH1 == 0 )] <- 'gCMS-B'


jusakul.sample.scored <- jusakul_patient


jusakul.sample.scored$OS_STATUS[jusakul.sample.scored$OS_STATUS == "0:LIVING"] <- 0
jusakul.sample.scored$OS_STATUS[jusakul.sample.scored$OS_STATUS == "1:DECEASED"] <- 1

jusakul.sample.scored$OS_MONTHS <- as.numeric(jusakul.sample.scored$OS_MONTHS)
jusakul.sample.scored$OS_STATUS <- as.numeric(jusakul.sample.scored$OS_STATUS)


jusakul.sample.scored$STAGE.BIN <- jusakul.sample.scored$STAGE
jusakul.sample.scored$STAGE.BIN[jusakul.sample.scored$STAGE %in% c('0','TBA') ] <- NA
jusakul.sample.scored$STAGE.BIN[jusakul.sample.scored$STAGE %in% c('I','IA') ] <- 'I'
jusakul.sample.scored$STAGE.BIN[jusakul.sample.scored$STAGE %in% c('II','IIA','IIB') ] <- 'II'
jusakul.sample.scored$STAGE.BIN[jusakul.sample.scored$STAGE %in% c('III','IIIA','IIIB') ] <- 'III'
jusakul.sample.scored$STAGE.BIN[jusakul.sample.scored$STAGE %in% c('IV','IVA','IVB','4') ] <- 'IV'

jusakul.sample.scored <- jusakul.sample.scored %>% filter(ANATOMICAL_SUBTYPE != 'N/A')

jusakul.os.cms <- survfit(Surv(OS_MONTHS, OS_STATUS) ~ rna_class , 
                          data = jusakul.sample.scored )


jusakul.os.cms.ggsurv <- 
  ggsurvplot(jusakul.os.cms, data = jusakul.sample.scored , 
             pval = F, conf.int = T,
             xlab="Months after diagnosis",
             legend="none", 
             xlim = c(0,100),
             risk.table = F,  # Adds number at risk table
             #  surv.median.line = "hv", 
             palette =  c("#04597C","#EDA71A")
  )


os.cms.jusakul.plot.g <- jusakul.os.cms.ggsurv$plot
os.cms.jusakul.plot.g <- 
  os.cms.jusakul.plot.g + 
  theme_bw(base_size=10) + guides(color='none',fill='none') +
  theme(panel.grid = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()) +
  scale_x_continuous(breaks = seq(0, max(jusakul.sample.scored$OS_MONTHS, na.rm = T), by = 12))

os.cms.jusakul.plot.g

ggsave(paste0(basedir, "/results/fig/FIGS2G.cms.os.jusakul.svg"),plot =os.cms.jusakul.plot.g
       , device = "svg",width = 67, height = 65, units = 'mm' )

cox.os.cms.jusakul <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ rna_class + STAGE.BIN, 
                            data =  jusakul.sample.scored  )

ggforest(cox.os.cms.jusakul, data =jusakul.sample.scored)

idh.cox <-  coxph(Surv(OS_MONTHS, OS_STATUS) ~ STAGE.BIN + rna_class + ANATOMICAL_SUBTYPE, 
                 data =  jusakul.sample.scored  )

ggforest(idh.cox, data = jusakul.sample.scored)

```


## genetic panel data from Memorial Sloan Kettering (MSK)
Combined datasets from [Jolissaint et al. 2020](https://pubmed.ncbi.nlm.nih.gov/33963001/) ([download on cbioportal](https://www.cbioportal.org/study/summary?id=ihch_mskcc_2020)), [Boerner et al. 2021](https://pubmed.ncbi.nlm.nih.gov/33765338/) ([download on cbioportal](https://www.cbioportal.org/study/summary?id=ihch_msk_2021)) and [Song Y et al. 2024](https://pubmed.ncbi.nlm.nih.gov/38864854/) ([download on cbioportal](https://www.cbioportal.org/study/summary?id=hcc_msk_2024)).

```{r}


hcc_2024_patients = fread(file.path(secdir, '/source_data/cbioportal/hcc_msk_2024', 'data_clinical_patient.txt'), skip = "PATIENT_ID")
names(hcc_2024_patients) <- paste0(names(hcc_2024_patients),'_hcc24')


icca_2021_patients = fread(file.path(secdir, '/source_data/cbioportal/ihch_msk_2021', 'data_clinical_patient.txt'), skip = "PATIENT_ID")
names(icca_2021_patients) <- paste0(names(icca_2021_patients),'_icc21')

icca_cc2020_patients = fread(file.path(secdir, '/source_data/cbioportal/ihch_mskcc_2020', 'data_clinical_patient.txt'), skip = "PATIENT_ID")
names(icca_cc2020_patients) <- paste0(names(icca_cc2020_patients),'_cc20')

these_patients <- full_join(hcc_2024_patients, icca_2021_patients, by=c('PATIENT_ID_hcc24'='PATIENT_ID_icc21'))  %>%
  full_join(icca_cc2020_patients, by=c('PATIENT_ID_hcc24'='PATIENT_ID_cc20')) 

these_samples = fread(file.path(secdir, '/source_data/cbioportal/hcc_msk_2024', 'data_clinical_sample.txt'), skip = "SAMPLE_ID")
names(these_samples) <- paste0(names(these_samples),'_hcc24')

these_patients <- left_join(these_patients, these_samples, by=c('PATIENT_ID_hcc24'='PATIENT_ID_hcc24'))

these_patients$SAMPLE_ID_hcc24[is.na(these_patients$SAMPLE_ID_hcc24)] <- paste0(gsub('p_','s_',these_patients$PATIENT_ID_hcc24[is.na(these_patients$SAMPLE_ID_hcc24)]),'_T') 
these_patients$SAMPLE_ID_hcc24 <- ifelse(grepl("^s_C_", these_patients$SAMPLE_ID_hcc24), paste0(these_patients$SAMPLE_ID_hcc24, "001_d"), these_patients$SAMPLE_ID_hcc24)

these_patients$ONCOTREE_CODE_hcc24[is.na(these_patients$ONCOTREE_CODE_hcc24) & !is.na(these_patients$DUCT_TYPE_cc20)] <- 'IHCH'

these_patients <- these_patients %>% filter(ONCOTREE_CODE_hcc24 != 'HCC')

primary_location_tally <- these_patients %>% group_by(ONCOTREE_CODE_hcc24) %>% tally()

ggplot(primary_location_tally, aes(x=reorder(ONCOTREE_CODE_hcc24, -n), y=n)) + 
  geom_bar(stat='identity',aes(fill=ONCOTREE_CODE_hcc24)) +
  geom_point(size=7, color = 'white', shape=16) +
  geom_point(size=7,  shape=1) +
  geom_text(aes(label=n)) +
  scale_fill_manual(values = pnw_palette(name="Bay", n=length(unique(primary_location_tally$ONCOTREE_CODE_hcc24)) ), na.value = 'grey90') + 
  guides(fill='none')+
  theme_bw(base_size=16) +
  theme(axis.text.x = element_text(hjust = -0, angle = -45), 
        axis.title.x = element_blank(), 
        panel.grid = element_blank() 
        
  )

these_patients %>% group_by(DUCT_TYPE_cc20, ONCOTREE_CODE_hcc24) %>% tally()

write.table( these_patients,  file = file.path(basedir,"/results/msk.patients.txt"), row.names = FALSE, quote = FALSE, sep = "\t")


make.hg19.bed = T
if(make.hg19.bed){
  library(rtracklayer)
  hg19.refGene.gtf <- fread(file.path(secdir, '/source_data/hg19.refGene.gtf')) %>% filter(V3 == 'transcript')
  names(hg19.refGene.gtf) <- c('seqname','source','feature','start','end','score','strand','frame','attribute')
  
  gr <- as(hg19.refGene.gtf, "GRanges")
  
  hg19.bed <- data.frame(
    chrom = as.character(seqnames(gr)),
    start = start(gr),
    end = end(gr)
  )
  
  # Add to GRanges metadata
  hg19.bed$gene_name <- sub('.*gene_name "([^"]+)".*', '\\1', mcols(gr)$attribute)
  hg19.bed$transcript_id <- sub('.*transcript_id "([^"]+)".*', '\\1', mcols(gr)$attribute)
  
  
  hg19.bed$length <- hg19.bed$end - hg19.bed$start + 1
  
  
  hg19.bed <- hg19.bed %>%
    group_by(gene_name) %>%
    slice_max(order_by = length, n = 1, with_ties = FALSE) %>%
    ungroup()
  
  
}

msk.mut <- fread(file.path(secdir, '/source_data/cbioportal/hcc_msk_2024/data_mutations.txt'),skip=2)
msk.mut$mean.vaf <- msk.mut$t_alt_count / (msk.mut$t_ref_count + msk.mut$t_alt_count)

msk.med.vaf <- msk.mut %>%
  group_by(Tumor_Sample_Barcode ) %>% 
  summarise(across(mean.vaf, \(x) mean(x, na.rm = TRUE)))

msk21.mut <- fread(file.path(secdir, '/source_data/cbioportal/ihch_msk_2021/data_mutations.txt'))
msk21.mut$mean.vaf <- msk21.mut$t_alt_count / (msk21.mut$t_ref_count + msk21.mut$t_alt_count)
msk21.mut <-  msk21.mut %>% filter(Tumor_Sample_Barcode %ni% msk.med.vaf$Tumor_Sample_Barcode)

msk21.med.vaf <- msk21.mut %>%
  group_by(Tumor_Sample_Barcode ) %>% 
  summarise(across(mean.vaf, mean, na.rm = TRUE))

msk.med.vaf <- rbind.data.frame(msk.med.vaf, msk21.med.vaf)

msk.driver <- msk.mut %>% filter(Hugo_Symbol %in% c('BAP1','FGFR2','IDH1','TP53','SMAD4')) %>% 
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol) 

msk21.driver <- msk21.mut %>% filter(Hugo_Symbol %in% c('BAP1','FGFR2','IDH1','TP53','SMAD4')) %>% 
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol) 

msk.driver <- rbind.data.frame(msk.driver, msk21.driver) %>% dcast(Tumor_Sample_Barcode ~ Hugo_Symbol, value.var = 'Hugo_Symbol', fun.aggregate = length)


msk.cn <- fread(file.path(secdir, '/source_data/cbioportal/hcc_msk_2024/data_cna_hg19.seg'))
msk.cn$chrom <- paste0('chr',msk.cn$chrom)
msk.cn <- msk.cn %>% filter(!is.na(seg.mean))

msk21.cn <- fread(file.path(secdir, '/source_data/cbioportal/ihch_msk_2021/data_cna_hg19.seg'))
msk21.cn$chrom <- paste0('chr',msk21.cn$chrom)
msk21.cn <- msk21.cn %>% filter(!is.na(seg.mean) & ID %ni% msk.cn$ID)
length(unique(msk21.cn$ID))

msk.cn <- rbind.data.frame(msk.cn, msk21.cn)

msk.segmented.data <- CNSeg(msk.cn)
msk.segment.gene <- getRS(msk.segmented.data, by="gene", imput=FALSE, XY=FALSE, geneMap=as.data.frame(hg19.bed), what="min")
msk.segment.gene <- rs(msk.segment.gene)

cytoBand <- fread(file.path(basedir,'/source_data/cytoBand.txt'), na.strings = '') %>% filter(!is.na(Chromosome))

cytoBand<- separate(cytoBand, col = Chromosome, into = c('band','period'), sep = '\\.', remove=F, extra = 'drop', fill = 'right')

cytoBand <- cytoBand %>%
  mutate(
    parts = str_match(band, "^(.*?[pq])(.*)$"),
    arm = parts[,2],
    armband = parts[,3]
  ) %>%
  dplyr::select(-parts) %>% filter(!is.na(arm) & arm %ni% c('Xq','Xp','Yp','Yq'))

msk.segment.gene.cyto <- inner_join(cytoBand, msk.segment.gene,  by=c('Hugo_Symbol'='gene_name'))

msk.arm.med <- msk.segment.gene.cyto %>%
  group_by(arm ) %>% 
  summarise(across(12:ncol(msk.segment.gene.cyto)-1, median, na.rm = TRUE))

msk.transposed.raw <- msk.arm.med %>%
  column_to_rownames(var = names(msk.arm.med)[1]) %>%  # Use first column as rownames
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>%
  mutate(across(-sample, as.numeric)) %>%  # Ensure numeric columns
  as_tibble()

names(msk.transposed.raw)[-1] <- paste0('arm.',names(msk.transposed.raw)[-1])

msk.transposed.raw <- inner_join(msk.med.vaf, msk.transposed.raw, by=c('Tumor_Sample_Barcode'='sample'))

#### PREDICT gCMS ####

full_model <- readRDS(file.path(basedir,'/results/cnv.model.rds'))
THRESHOLD = 0.34

msk.transposed.raw$prob <- predict(full_model, msk.transposed.raw , type = "response")

msk.driver.cnv <- 
  full_join(msk.driver, msk.transposed.raw, by=c('Tumor_Sample_Barcode'='Tumor_Sample_Barcode')) 

msk.driver.cnv$sample <- factor(msk.driver.cnv$Tumor_Sample_Barcode, levels=msk.driver.cnv$Tumor_Sample_Barcode[order(msk.driver.cnv$prob)])

msk.BAP1_tumours <- msk.driver.cnv %>% filter(BAP1 >0 & TP53 == 0) %>% dplyr::select(Tumor_Sample_Barcode) %>% unlist()
msk.tp53_tumours <- msk.driver.cnv %>% filter(TP53 > 0 & BAP1 == 0) %>% dplyr::select(Tumor_Sample_Barcode) %>% unlist()

msk.driver.cnv$driver.class <- NA
msk.driver.cnv$driver.class[msk.driver.cnv$Tumor_Sample_Barcode %in% msk.BAP1_tumours] <- 'BAP1+'
msk.driver.cnv$driver.class[msk.driver.cnv$Tumor_Sample_Barcode %in% msk.tp53_tumours] <- 'TP53+'

msk.driver.cnv <- full_join(msk.driver.cnv, these_patients, by=c('Tumor_Sample_Barcode'='SAMPLE_ID_hcc24'))

msk.subset <- msk.driver.cnv %>% dplyr::filter(!is.na(driver.class) & CANCER_TYPE_DETAILED_hcc24 != 'Hepatocellular Carcinoma' & mean.vaf > 0.15)

nrow(msk.subset)
roc_obj <- roc(msk.subset$driver.class, msk.subset$prob)
plot(roc_obj)
auc(roc_obj)

coords(roc_obj, THRESHOLD, ret = c("threshold", "sensitivity", "specificity"))

ggplot(msk.subset , aes(x=reorder(Tumor_Sample_Barcode, prob), y=prob)) + 
 
 geom_hline(yintercept = THRESHOLD, color='grey90')+
 
 geom_point(aes(color=factor(driver.class)),  shape=16, size=0.5, stroke=0.5) + 
 
 theme_bw(base_size=10) +
 scale_color_manual(values = c('BAP1+'='#04597C',
                               'TP53+'='#EDA71A'
 ), 
 na.value = 'grey80')  +
 theme(panel.grid = element_blank(),
       panel.background = element_blank(), 
       
       plot.background = element_blank(),
       axis.text.x = element_blank()
 ) + 
 ylim(0,1)+
 labs(x='', y='CNV-Based Probability of CMS-B',color='')

       
msk.subset <- msk.driver.cnv %>% dplyr::filter( (CANCER_TYPE_DETAILED_hcc24 != 'Hepatocellular Carcinoma' | is.na(CANCER_TYPE_DETAILED_hcc24)) & 
                                                  (mean.vaf > 0.15 | is.na(mean.vaf)) )

msk.subset$glm_labels <- ifelse(msk.subset$prob < THRESHOLD, 'gCMS-B', 'gCMS-A')
table(msk.subset[,c('glm_labels','driver.class')])


write.table(
  msk.subset,
  file = paste0(basedir, "/results/msk.gCMS.txt"),
  row.names = T, quote = FALSE, sep = "\t"
)


msk.btc <- fread(paste0(basedir, "/results/msk.gCMS.txt"))
msk.btc$OS_STATUS_hcc24[msk.btc$OS_STATUS_hcc24 == "0:LIVING"] <- 0
msk.btc$OS_STATUS_hcc24[msk.btc$OS_STATUS_hcc24 == "1:DECEASED"] <- 1
msk.btc$OS_STATUS_hcc24 <- as.numeric(msk.btc$OS_STATUS_hcc24)

msk.btc$OS_STATUS_icc21[msk.btc$OS_STATUS_icc21 == "0:LIVING"] <- 0
msk.btc$OS_STATUS_icc21[msk.btc$OS_STATUS_icc21 == "1:DECEASED"] <- 1
msk.btc$OS_STATUS_icc21 <- as.numeric(msk.btc$OS_STATUS_icc21)

msk.btc.surv <- msk.btc %>% filter(!is.na(OS_STATUS_hcc24) & !is.na(OS_MONTHS_hcc24) & !is.na(glm_labels)  )
msk.btc.surv$glm_labels <- as.factor(msk.btc.surv$glm_labels)
nrow(msk.btc.surv)

os <- survfit(Surv(OS_MONTHS_hcc24, OS_STATUS_hcc24) ~ glm_labels  , 
              data = msk.btc.surv )

cox.os.cms.msk <- coxph(Surv(OS_MONTHS_hcc24, OS_STATUS_hcc24) ~ glm_labels  , 
                        data =  msk.btc )

ggforest(cox.os.cms.msk, data =msk.btc)


msk.surv.plt <- 
  ggsurvplot(os, data = msk.btc, 
             pval = F, 
             conf.int = T, 
             censor = TRUE,
             risk.table = F,  
             legend="none", 
             xlab="Months after diagnosis",
             palette =  c("#04597C","#EDA71A")
  ) 

msk.surv.plt.g <- msk.surv.plt$plot
msk.surv.plt.g <- 
  msk.surv.plt.g + 
  theme_bw(base_size=10) + guides(color='none',fill='none') +
  theme(panel.grid = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()) +
  scale_x_continuous(breaks = seq(0, max(msk.btc$OS_MONTHS_icc21, na.rm = T), by = 24), limits = c(0, 180)) 

msk.surv.plt.g
ggsave(paste0(basedir, "/results/fig/FIGS2H.msk.gCMS.OS.svg"), plot = msk.surv.plt.g
       , device = "svg",width = 67, height = 65, units = 'mm', dpi=72 )


msk.btc.complete <- msk.btc %>% filter(TUMOR_STAGE_hcc24 != '')

cox.os.cms.msk <- coxph(Surv(OS_MONTHS_hcc24, OS_STATUS_hcc24) ~ glm_labels + ONCOTREE_CODE_hcc24 + TUMOR_STAGE_hcc24 , 
                        data =  msk.btc.complete )

full.forest <- ggforest(cox.os.cms.msk, data = msk.btc.complete,
                        fontsize = 0.5
)

cox.os.cms.msk

cox.os.cms.msk.summary <-   summary(cox.os.cms.msk)

full.forest.df <- cbind.data.frame(
  cox.os.cms.msk.summary$conf.int,
  'p'=cox.os.cms.msk.summary$coefficients[,5]
) %>% as.data.frame() %>% dplyr::select(`exp(coef)`, `lower .95`, `upper .95`, p)


names(full.forest.df) <- c('HZ','ci.low','ci.hi','p')
full.forest.df$name <- rownames(full.forest.df)

msk.btc.n <- msk.btc.complete %>% dplyr::select(glm_labels, ONCOTREE_CODE_hcc24, TUMOR_STAGE_hcc24 ) %>% 
  melt(measure.vars=c('glm_labels', 'ONCOTREE_CODE_hcc24', 'TUMOR_STAGE_hcc24'))

msk.btc.n$name <- paste0(msk.btc.n$variable, msk.btc.n$value )
msk.btc.n <- msk.btc.n %>% group_by(name, variable, value)  %>% filter(!is.na(value)) %>% tally()

full.forest.df <- left_join(msk.btc.n, full.forest.df, by=c('name'='name'))
full.forest.df$HZ[is.na(full.forest.df$HZ)] <- 1

full.forest.df$p <- round(full.forest.df$p,3)


full.forest.df <- full.forest.df %>%
  mutate(p = if_else(!is.na(p) & p < 0.05,
                     paste0(p, "*"),
                     as.character(p)))

full.forest.df$p[is.na(full.forest.df$p)] <- '(reference)'

full.forest.df$variable <- as.character(full.forest.df$variable)
full.forest.df$variable[full.forest.df$variable == 'ONCOTREE_CODE_hcc24'] <- 'Location'
full.forest.df$variable[full.forest.df$variable == 'TUMOR_STAGE_hcc24'] <- 'Stage'
full.forest.df$variable[full.forest.df$variable == 'glm_labels'] <- 'gCMS'
full.forest.df$category <- factor(full.forest.df$variable, levels = c('gCMS','Location','Stage'))

full.forest.df$value[full.forest.df$value == 'EHCH'] <- 'pCCA/dCCA'
full.forest.df$value[full.forest.df$value == 'HCCIHCH'] <- 'HCC-CCA'
full.forest.df$value[full.forest.df$value == 'IHCH'] <- 'iCCA'

full.forest.df$value.n <- paste0(full.forest.df$value,'\nn=',full.forest.df$n)
full.forest.df$type <- factor(full.forest.df$value.n, levels = c('GBC\nn=120','pCCA/dCCA\nn=83','iCCA\nn=361','HCC-CCA\nn=32',
                                                                 'Well differentiated\nn=23','Moderately differentiated\nn=323','Poorly differentiated\nn=250',
                                                                 'gCMS-B\nn=319','gCMS-A\nn=215'))


msk.gCMS.HZ.plot <- 
ggplot(full.forest.df, aes(x=-log(HZ), y=type)) +
         facet_grid(category~., scales='free',space='free',switch='y')+
         geom_vline(xintercept = 0, color='grey80')+
         geom_errorbar( aes(xmin =  -log(ci.low), xmax = -log(ci.hi)), width=0) +
         geom_point(shape=16, aes(color=type))+ 
         geom_text(aes(label=p), x=min(-log(full.forest.df$ci.hi), na.rm = T), hjust=1.5, size=2.5) +
         #coord_fixed()+
         theme_bw(base_size=10) +
         # scale_color_manual(values = pnw_palette(name="Bay", n=30)[c(3,20)], na.value = 'grey90') +
         
         theme(panel.grid = element_blank(), 
               panel.border = element_blank(),
               panel.spacing = unit(0.2, "lines"),
               strip.background = element_blank(),
               plot.margin = unit(c(0, 0, 0, 0), "mm"),
               legend.position = 'bottom',
               
               panel.background = element_blank(), 
               plot.background = element_blank() ,
               axis.ticks.y = element_blank() ,
               axis.line.x = element_line(color = "grey80"),
               axis.text.y  = element_text(size = 7, margin = margin(r = unit(55, "mm")))  ,
               strip.placement = "outside",        # Place strip outside axis
               strip.text.y.left = element_text(   # Style facet text
                 angle = 0,                        # Horizontal text
                 face = "bold",                    # Bold
                 hjust = 1                         # Align to left of axis labels
               )
               
         ) +scale_color_manual(  values=c(
           "gCMS-B\nn=319"    = "#EDA71A" , 
           "gCMS-A\nn=215" = "#04597C" 
         ), na.value = 'black') +
         guides(color='none', size='none')+
         labs(x='Hazard Ratio',y='')+
         scale_y_discrete(position = "left")+ 
         coord_cartesian(clip = "off") 

msk.gCMS.HZ.plot

ggsave(paste0(basedir, "/results/fig/FIG2G.msk.gCMS.HZ.svg"), plot = 
         
     msk.gCMS.HZ.plot
       
       ,
       
       device = "svg", width = 100, height = 75, 
       units = 'mm', dpi=72 )


```

## anatomical location, still using MSK

```{r}

msk.subset[msk.subset == ''] <- NA

msk.subset %>%  dplyr::select(glm_labels, DUCT_TYPE_cc20) %>%  table()

msk.subset$LD <- NA
msk.subset$LD[!is.na(msk.subset$DUCT_TYPE_cc20) ] <- 0
msk.subset$LD[msk.subset$DUCT_TYPE_cc20 == 'Large bile duct type' ] <- 1

large.model <- msk.subset %>% dplyr::select(glm_labels, LD) %>% table() %>% fisher.test()

msk.subset$SD <- NA
msk.subset$SD[!is.na(msk.subset$DUCT_TYPE_cc20) ] <- 0
msk.subset$SD[msk.subset$DUCT_TYPE_cc20 == 'Small bile duct type' ] <- 1

small.model <- msk.subset %>% dplyr::select(glm_labels, SD) %>% table() %>% fisher.test()

msk.btc.complete <- msk.subset %>% filter(TUMOR_STAGE_hcc24 != '')


msk.btc.complete$eCCA <- NA
msk.btc.complete$eCCA[!is.na(msk.btc.complete$ONCOTREE_CODE_hcc24) ] <- 0
msk.btc.complete$eCCA[msk.btc.complete$ONCOTREE_CODE_hcc24 == 'EHCH' ] <- 1

e.model <- msk.btc.complete %>% dplyr::select(glm_labels, eCCA) %>% table() %>% fisher.test()

msk.btc.complete$GBC <- NA
msk.btc.complete$GBC[!is.na(msk.btc.complete$ONCOTREE_CODE_hcc24) ] <- 0
msk.btc.complete$GBC[msk.btc.complete$ONCOTREE_CODE_hcc24 == 'GBC' ] <- 1

g.model <- msk.btc.complete %>% dplyr::select(glm_labels, GBC) %>% table() %>% fisher.test()

msk.btc.complete$iCCA <- NA
msk.btc.complete$iCCA[!is.na(msk.btc.complete$ONCOTREE_CODE_hcc24) ] <- 0
msk.btc.complete$iCCA[msk.btc.complete$ONCOTREE_CODE_hcc24 == 'IHCH' ] <- 1

i.model <- msk.btc.complete %>% dplyr::select(glm_labels, iCCA) %>% table() %>% fisher.test()


msk.loc.lods <- rbind.data.frame(
  cbind.data.frame('type'='large duct','pvalue'=large.model$p.value, 'cilow'=large.model$conf.int[1], 'cihi'=large.model$conf.int[2], 'odds'=large.model$estimate, 'group'='iCCA', 'n'= 12),
  cbind.data.frame('type'='small duct','pvalue'=small.model$p.value, 'cilow'=small.model$conf.int[1], 'cihi'=small.model$conf.int[2], 'odds'=small.model$estimate , 'group'='iCCA', 'n'=117),
  
  cbind.data.frame('type'='eCCA','pvalue'=e.model$p.value, 'cilow'=e.model$conf.int[1], 'cihi'=e.model$conf.int[2], 'odds'=e.model$estimate , 'group'='BTC', 'n'=72),
  cbind.data.frame('type'='GBC','pvalue'=g.model$p.value, 'cilow'=g.model$conf.int[1], 'cihi'=g.model$conf.int[2], 'odds'=large.model$estimate , 'group'='BTC', 'n'=114),
  cbind.data.frame('type'='iCCA','pvalue'=i.model$p.value, 'cilow'=i.model$conf.int[1], 'cihi'=i.model$conf.int[2], 'odds'=i.model$estimate , 'group'='BTC', 'n'=322)
)

msk.loc.lods$sig <- NA
msk.loc.lods$sig[msk.loc.lods$pvalue < 0.05 & msk.loc.lods$odds < 1] <- 'CMS-A'
msk.loc.lods$sig[msk.loc.lods$pvalue < 0.05 & msk.loc.lods$odds > 1] <- 'CMS-B'


msk.loc.lods$type.n <- paste0(msk.loc.lods$type,'\nn=',msk.loc.lods$n)

msk.loc.lods$loc <- factor(msk.loc.lods$type.n, levels = rev(c('iCCA\nn=322','eCCA\nn=72','GBC\nn=114','small duct\nn=117','large duct\nn=12')))


cms.location.msk.plot <- 
     ggplot(msk.loc.lods %>% filter(group=='iCCA'), aes(x=-log(odds), y=loc)) +
         facet_grid(group~., scales='free',space='free')+
         geom_vline(xintercept = 0, color='grey80')+
         geom_errorbar( aes(xmin =  -log(cilow), xmax = -log(cihi)), width=0) +
         geom_point(aes(color=sig))+ 
         #coord_fixed()+
         theme_bw(base_size=10) +
         scale_color_manual(values = pnw_palette(name="Bay", n=30)[c(3,20)], na.value = 'grey90') +
         
         theme(panel.grid = element_blank(), 
               panel.border = element_blank(),
               panel.spacing = unit(1, "lines"),
               strip.background = element_blank(),
               strip.text = element_blank(),
               axis.ticks.y = element_blank(),
               # axis.text.y =element_blank(),
               legend.position = 'bottom',
               panel.background = element_blank(), 
               plot.background = element_blank() ,
               axis.line.x = element_line(color = "grey80")
               
         ) +
         guides(color='none', size='none')+
         labs(x='gCCS-A LOD',y='')+
         # scale_x_discrete(position = "top") +
         scale_y_discrete(position = "left")

cms.location.msk.plot     
  
ggsave(file.path(basedir, "/results/fig/FIG2E.cms.location.msk.svg"),plot =
         cms.location.msk.plot
       , device = "svg", width = 55, height = 30, units = 'mm' )

```

