# Differential methylation and protein expression 
Here, we return to our external datasets from [Dong et al. 2022](https://doi.org/10.1016/j.ccell.2021.12.006) and [Jusakul et al. 2017](https://pubmed.ncbi.nlm.nih.gov/28667006/) to explore other data modalities.

## Data intake

```{r}

#BiocManager::install("lumi")

suppressPackageStartupMessages({

  library(data.table)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(stringr)
  
  library(lumi)
  library(limma)
  library(minfi)
  library(IlluminaHumanMethylation450kmanifest)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  
  library(FactoMineR)
  library(Rtsne)
  
  library(ggplot2)
  library(ggrepel)
  library(PNWColors)
  
})

packageVersion("lumi")
packageVersion('minfi')
packageVersion('limma')

`%ni%` <- Negate(`%in%`)

basedir = '~/Documents/scripts/github/btc-ccs-old/'
secdir = '~/Documents/scripts/github/evolution/'

source( file.path(basedir, "bin/BTC.functions.R"))

classifier <- readRDS(file.path(basedir,'/results/tps.classifier.rds'))

CMSA_MARKERS = c('HHEX','PROX1','CFTR','BICC1','SPP1', 'DCDC2', 'SCTR','ONECUT1','ALB','NCAM1', 
                 'HNF1B','CD24','TNFSF12') 

CMSB_MARKERS = c('S100P', 'MKI67','SSTR1','MUC2','MUC5AC','CDX2','KRT20','PTGS2','CEACAM5', 
                 'MUC17','KRT19','KRT7','CDK1','CD44') 

```

## Methylation variance (from Jusakul et al.)
We take the methylation array data from Jusakul et al and identify differentially methylated genes between samples classified by eCCS (in step 1.3.1). We also add in a little analysis of IDH1+ samples and find patterns suggestive that within CCS-A, these samples might be doing their own thing (as expected from methylation of IDH1+ samples).

```{r}

jusakul_methylation.meta <-  fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW/jusakul_methylation.txt'))

jusakul_methylation.meta$barcode <- 
  paste(
    jusakul_methylation.meta$V2,
    jusakul_methylation.meta$V3,
    jusakul_methylation.meta$V4,
    sep='_'
  )

jusakul_methylation.meta$icgc_id <- 
  paste(
    jusakul_methylation.meta$V8,
    jusakul_methylation.meta$V9,
    jusakul_methylation.meta$V10,
    sep='_'
  )

jusakul.predicted.w.score <-  fread(file.path(basedir, "/results/jusakul.eCMS.txt"))
names(jusakul.predicted.w.score)[1] <- 'icgc_id'

jusakul_methylation.meta <- full_join(jusakul_methylation.meta, jusakul.predicted.w.score, by=c('icgc_id'='icgc_id'))


# Set the path to your .idat files
idat_dir <- file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW')

unzip.files =F
if(unzip.files){
  gz_files <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  
  # Unzip each file
  for (f in gz_files) {
    R.utils::gunzip(f, remove = TRUE)  # remove = TRUE deletes the .gz file after extraction
  }
  
}


RGSet <- read.metharray.exp(base = idat_dir, recursive = TRUE)

MSet <- preprocessRaw(RGSet)  

normSet <- preprocessQuantile(RGSet)

betaVals <- getBeta(normSet)

betaVals.df <- cbind.data.frame( betaVals)

jusakul.methyl.pca <- PCA(t(betaVals), graph = F)
jusakul.methyl.sites <- jusakul.methyl.pca$var$coord
jusakul.methyl.sample <- cbind.data.frame('barcode'=rownames(jusakul.methyl.pca$ind$coord),
                                          jusakul.methyl.pca$ind$coord)

jusakul.methyl.sample <- left_join(jusakul.methyl.sample, jusakul_methylation.meta, by=c('barcode'='barcode'))

ggplot(jusakul.methyl.sample, aes(x=Dim.1, y=Dim.2, color=rna_class)) + 
         geom_point(size=2) + theme_bw(base_size=10) +
         
         stat_ellipse(data = ~ subset(., !is.na(rna_class) ), linetype='dashed')+
         scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + 
         theme(    plot.margin = unit(c(0, 0, 0, 0), "mm"),
                   panel.background = element_blank(),
                   plot.background = element_blank(),
                   
                   panel.grid = element_blank()) + coord_fixed() + 
         labs(x='Jusakul et al. Methylation PC1',y='Jusakul et al. Methylation PC2') + guides(color='none')
       

tsne_result <- Rtsne(t(betaVals), dims = 2, perplexity = 30, verbose = TRUE, max_iter = 500)
tsne.dims <- as.data.frame(tsne_result$Y)
names(tsne.dims) <- c('tsne.dim.1','tsne.dim.2')
tsne.dims$barcode <- colnames(betaVals)

tsne.dims <- left_join(tsne.dims, jusakul.methyl.sample, by=c('barcode'='barcode'))

jusakul_muts <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_mutations.txt'))
jusakul_drivers <- jusakul_muts %>% filter(Hugo_Symbol %in% c('BAP1','TP53','FGFR2','IDH1','SMAD4')) %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, Consequence)

tsne.dims$IDH1 <- 'wt'
tsne.dims$IDH1[tsne.dims$icgc_id %in% jusakul_drivers$Tumor_Sample_Barcode[jusakul_drivers$Hugo_Symbol == 'IDH1']] <- 'IDH1+'

ggplot(tsne.dims, aes(x=tsne.dim.1, y=tsne.dim.2, color=rna_class)) + 
         geom_point(size=2, aes(shape=IDH1)) + theme_bw() +
         scale_shape_manual(values = c('wt'=1,'IDH1+'=16)) +
         
         scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + theme(panel.grid = element_blank())

## explain variance

jusakul_sample <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_sample.txt'), skip=4)
jusakul_patient <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_patient.txt'), skip=4)

jusakul_patient <- full_join(jusakul_patient, jusakul_sample, by=c('PATIENT_ID'='PATIENT_ID'))
jusakul_patient <- jusakul_patient %>% filter(ANATOMICAL_SUBTYPE != 'N/A' & SAMPLE_ID %in% tsne.dims$icgc_id[!is.na(tsne.dims$rna_class)] )
table(jusakul_patient$ANATOMICAL_SUBTYPE)

jusakul_methylation.pat <- left_join(jusakul_patient, jusakul_methylation.meta, by=c('SAMPLE_ID'='icgc_id'))


#TODO: subset to patients with all info (CCS, location and methylation)

locations <- jusakul_methylation.pat$ANATOMICAL_SUBTYPE
names(locations) <- jusakul_methylation.pat$SAMPLE_ID

rna_class <- jusakul_methylation.pat$rna_class
names(rna_class) <- jusakul_methylation.pat$SAMPLE_ID

betas.t <- t(betaVals)
betas.t.filt <- betas.t[jusakul_methylation.pat$barcode,]
betas_rda <- rda(betas.t.filt, scale = TRUE)


# Test fit on top 2 axes
r.tab = NULL
for(i in c(1:5)){

  fit2_f1 <- envfit(betas_rda, locations, permutations = 999, choices = i:i)
  fit2_f2 <- envfit(betas_rda, rna_class, permutations = 999, choices = i:i)

  this.r.tab <- rbind.data.frame ( cbind('i'=as.numeric(i),
               'f'='location',
               'r'=unlist(fit2_f1$factors$r),
               'p'=unlist(fit2_f1$factors$pvals)
  ),
  cbind('i'=as.numeric(i),
        'f'='eCCs',
        'r'=unlist(fit2_f2$factors$r),
        'p'=unlist(fit2_f2$factors$pvals)
  )
  )
  if(is.null(r.tab)){
    r.tab <- this.r.tab
  } else{
    r.tab <- rbind.data.frame(r.tab,this.r.tab)
  }
  
}

r.tab$p <- as.numeric(r.tab$p)
r.tab$r <- as.numeric(r.tab$r)

r.tab$p.sig <- NA
r.tab$p.sig[r.tab$p < 0.05] <- r.tab$p[r.tab$p < 0.05]

r.tab$pc.i <- paste0('PC',r.tab$i)
r.tab$pc.i <- factor(r.tab$pc.i, levels=rev(paste0('PC',c(1:5)) ))
r.tab

methyl.rda.plot <- 
ggplot(r.tab %>% filter(i %in% c(2:4)), aes(y=pc.i, x=r, shape=f)) + 
  geom_point(aes( size=-log(p.sig)), fill='white') + 
  geom_line(aes(group=pc.i), color='white',
            arrow = arrow(length = unit(1, "mm"), type = "closed"), 
  )+
  geom_line(aes(group=pc.i) )+
  scale_shape_manual(values = c(16,21)) +
  scale_size(range = c(5, 10)) +
  theme_bw(base_size=30)+
  theme(
    plot.margin = unit(c(0, 0, 0, 5), "mm"),
    panel.background = element_blank(),
    #panel.border = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    legend.position = 'bottom'
  ) + xlim(0,0.3) + labs(x=expression(r^2~ "of methylation variance")) +
  guides(shape='none',size='none')

methyl.rda.plot

ggsave(file.path(basedir,"/results/fig/CCF26.methyl_rda.svg"),plot =
  methyl.rda.plot       
, device = "svg", height = 56,width = 200 , units='mm')

```

## Differential Methylation

```{r}

pheno <- jusakul_methylation.meta %>% filter(!is.na(barcode) ) %>% 
  dplyr::select(barcode, icgc_id, V6, rna_class, confidence.polarized) %>% DataFrame()
rownames(pheno) <- pheno[,1]

pData(normSet) <- pheno

normSet_filtered <- normSet[ , !is.na(pData(normSet)$rna_class) ]

betaVals <- getBeta(normSet_filtered)

group <- factor(pData(normSet_filtered)$rna_class)  
design <- model.matrix(~ group)

fit <- lmFit(betaVals, design)
fit <- eBayes(fit)

topTable <- topTable(fit, coef = 2, number = Inf, adjust = "BH")
topTable$IlmnID <- rownames(topTable)

Gene_features <- fread(file.path(secdir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW/GPL16304_Gene_features_PlatformTable.txt'))

topTable.ft <- full_join(topTable, Gene_features, by=c('IlmnID'='IlmnID'))

top.gene.mthyl <- topTable.ft %>% filter(IlluminaGene != '')  %>%
  group_by(IlluminaGene) %>%
  slice_max(order_by = abs(logFC), n = 1, with_ties = FALSE) %>%
  ungroup()

multiple.testing <- nrow(top.gene.mthyl)



top.gene.mthyl$eccs <- NA
top.gene.mthyl$eccs[top.gene.mthyl$IlluminaGene %in% classifier$TSPs[,1]] <- 'CMS-A'
top.gene.mthyl$eccs[top.gene.mthyl$IlluminaGene %in% classifier$TSPs[,2]] <- 'CMS-B'

jusakul.diffmethyl.plot <- 
ggplot(top.gene.mthyl , aes(x=logFC, y=-log(adj.P.Val ))) + 
  geom_hline(yintercept = -log(0.05 ), color='grey',linetype='dotted')+
  
  geom_point( data = ~ subset(., adj.P.Val >= 0.05  ),   color='grey50') + 
  
  geom_point( data = ~ subset(.,  adj.P.Val < 0.05   & logFC > 0 ),   
              color=pnw_palette(name="Bay", n=19 )[c(1)]) + 
  geom_point( data = ~ subset(.,  adj.P.Val < 0.05  & logFC < 0 ),   
              color=pnw_palette(name="Bay", n=19 )[c(10)]) + 
  
  geom_label_repel(  data = ~ subset(., IlluminaGene %in% c(CMSA_MARKERS,
                                                         CMSB_MARKERS) & adj.P.Val < 0.05 ), 
                     aes(label=IlluminaGene, fill=eccs, color=eccs), 
                     size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  
  geom_point( data = ~ subset(., IlluminaGene %in% CMSA_MARKERS & adj.P.Val < 0.05 ),   
              color='black', fill=pnw_palette(name="Bay", n=19 )[c(1)], shape=21) + 
  geom_point( data = ~ subset(., IlluminaGene %in% CMSB_MARKERS & adj.P.Val < 0.05 ),   
              color='black', fill=pnw_palette(name="Bay", n=19 )[c(10)], shape=21) + 
  scale_fill_manual(values=c("CMS-A"="#04597C", "CMS-B"="#EDA71A"), na.value = 'white') +
  scale_color_manual(values=c("CMS-A"="white", "CMS-B"="white"), na.value = 'black') +
  labs(x='Differential methylation',y='adj. p-value (-log)')+
  guides(color='none')+
  guides(color='none',fill='none')+
  theme_bw(base_size=10) +
  theme(plot.background = element_blank(), 
        panel.background = element_blank(), 
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
        
        panel.grid =  element_blank()) + xlim(-max(abs(top.gene.mthyl$logFC)),max(abs(top.gene.mthyl$logFC)))

jusakul.diffmethyl.plot

ggsave(file.path(basedir, "/results/fig/FIG3C.jusakul.diffmethyl.svg"),plot =
jusakul.diffmethyl.plot
   , device = "svg",width = 80, height = 63, units = 'mm' )




```

## Protein expression from Dong et al (2022)

```{r}

dong.predicted.w.score <-  fread( file.path(basedir, "/results/dong.eCMS.txt"), )
rownames(dong.predicted.w.score) <- dong.predicted.w.score[[1]]
dong.predicted.w.score[[1]] <- NULL
dong.predicted.w.score <- dong.predicted.w.score %>% filter(!is.na(rna_class))

dong.prot.info <- read_excel(file.path(secdir,"/source_data/dong_2022_S1.xlsx"), sheet = 5, skip = 2)
names(dong.prot.info)[c(1:2)] <- c('genes','protein')
names(dong.prot.info)[-c(1:2)] <- paste0('DONG_',names(dong.prot.info[,-c(1:2)]))
dong.prot.info[dong.prot.info == 'NA'] <- NA

dong.prot.mat <- as.matrix(dong.prot.info[,-c(1:2)])
dong.prot.mat <- matrix(as.numeric(dong.prot.mat), nrow = nrow(dong.prot.mat), ncol = ncol(dong.prot.mat))

dong.prot.pca <- PCA(t(dong.prot.mat), graph = F)
dong.prot.sample <- data.frame(dong.prot.pca$ind$coord)
dong.prot.sample$PatientID <- names(dong.prot.info[,-c(1:2)])

dong.prot.sample <- left_join(dong.prot.sample, dong.predicted.w.score, by=c('PatientID'='PatientID'))

ggplot(dong.prot.sample ,aes(x=Dim.1, y=Dim.2, color=rna_class)) + 
  geom_point( size=2) + 
  theme_bw(base_size=10) +
  stat_ellipse(data = ~ subset(., !is.na(rna_class) ), linetype='dashed')+
  scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + 
  theme(    plot.margin = unit(c(0, 0, 0, 0), "mm"),
            panel.background = element_blank(),
            plot.background = element_blank(),
            panel.grid = element_blank()) + 
  guides(color='none') + coord_fixed() + labs(x='Dong et al. Protein PC1', y='Dong et al. Protein PC2')

cms.prot.dong.ids <- unique(intersect(names(dong.prot.info), dong.predicted.w.score$PatientID))

normalized_data.p <- dong.prot.info %>% dplyr::select(all_of(cms.prot.dong.ids))

dong.predicted.filt.p <- dong.predicted.w.score %>% 
  filter(PatientID %in% cms.prot.dong.ids) %>%
  arrange(match(PatientID, cms.prot.dong.ids))

group <- factor(gsub('-','',dong.predicted.filt.p$rna_class))
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

normalized_data.complete <- as.matrix(normalized_data.p)
normalized_data.complete <- apply(normalized_data.complete, 2, as.numeric)
rownames(normalized_data.complete) <- dong.prot.info$genes

fit <- lmFit(normalized_data.complete, design)
ncol(normalized_data.complete)
nrow(design)

contrast.matrix <- makeContrasts(
  TreatmentVsControl = CMSA - CMSB,  
  levels = design
)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

results <- topTable(fit2, coef = 1, number = Inf, adjust.method = "BH")

results$gene_name <- results$ID
multiple.testing <- nrow(results) 

results$eccs <- NA
results$eccs[results$gene_name %in% classifier$TSPs[,1]] <- 'CMS-A'
results$eccs[results$gene_name %in% classifier$TSPs[,2]] <- 'CMS-B'

dong.diffprot.plot <- 
         ggplot(results , aes(x=logFC, y=-log(adj.P.Val ))) + 
         geom_hline(yintercept = -log(0.05 ), color='grey',linetype='dotted')+
         
         geom_point( data = ~ subset(., adj.P.Val >= 0.05  ),   color='grey50') + 
         
         geom_point( data = ~ subset(.,  adj.P.Val < 0.05   & logFC > 0 ),   
                     color=pnw_palette(name="Bay", n=19 )[c(1)]) + 
         geom_point( data = ~ subset(.,  adj.P.Val < 0.05  & logFC < 0 ),   
                     color=pnw_palette(name="Bay", n=19 )[c(10)]) + 
         
         geom_label_repel(  data = ~ subset(., gene_name %in% c(CMSA_MARKERS,
                                                                CMSB_MARKERS) & adj.P.Val < 0.05 ), 
                            aes(label=gene_name, fill=eccs, color=eccs), 
                            size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
         
         geom_point( data = ~ subset(., gene_name %in% CMSA_MARKERS & adj.P.Val < 0.05 ),   
                     color='black', fill=pnw_palette(name="Bay", n=19 )[c(1)], shape=21) + 
         geom_point( data = ~ subset(., gene_name %in% CMSB_MARKERS & adj.P.Val < 0.05 ),   
                     color='black', fill=pnw_palette(name="Bay", n=19 )[c(10)], shape=21) + 
           scale_fill_manual(values=c("CMS-A"="#04597C", "CMS-B"="#EDA71A"), na.value = 'white') +
           scale_color_manual(values=c("CMS-A"="white", "CMS-B"="white"), na.value = 'black') +
         guides(color='none')+
         labs(x='Differential methylation',y='adj. p-value (-log)')+
         guides(color='none',fill='none')+
         theme_bw(base_size=10) +
         theme(plot.background = element_blank(), 
               panel.background = element_blank(), 
               plot.margin = unit(c(0, 0, 0, 0), "mm"),
               
               panel.grid =  element_blank()) + xlim(-max(abs(results$logFC)),max(abs(results$logFC)))

dong.diffprot.plot

ggsave(file.path(basedir, "/results/fig/FIG3d.dong.diffprot.svg"),plot =
         dong.diffprot.plot
      , device = "svg",width = 80, height = 63, units = 'mm' )


```


