# Differential methylation and protein expression 
Here, we return to our external datasets from [Dong et al. 2022](https://doi.org/10.1016/j.ccell.2021.12.006) and [Jusakul et al. 2017](https://pubmed.ncbi.nlm.nih.gov/28667006/) to explore other data modalities.

## Data intake

```{r}

#BiocManager::install("lumi")

suppressPackageStartupMessages({

  library(data.table)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(stringr)
  
  library(lumi)
  library(limma)
  library(minfi)
  library(IlluminaHumanMethylation450kmanifest)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  
  library(FactoMineR)
  library(Rtsne)
  library(vegan)
  
  library(ggplot2)
  library(ggrepel)
  library(PNWColors)
  
})

packageVersion("lumi")
packageVersion('minfi')
packageVersion('limma')

`%ni%` <- Negate(`%in%`)

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'

source( file.path(code.dir, "bin/BTC.functions.R"))

classifier <- readRDS(file.path(data.dir,'/results/tps.classifier.rds'))

CMSA_MARKERS = c( 'DCDC2','BICC1','ONECUT1',
                  'PROX1','HNF1B','SPP1', 
                  'CFTR','NCAM1')
  
CMSB_MARKERS = c('PTGS2','S100P', 'MKI67', 
                  'CDK1', 'CDX2', 'CD44',
                 'KRT7', 'KRT19', 'KRT20',
                 'MUC2','MUC5AC','MUC17')

```

## Methylation variance (from Jusakul et al.)
We take the methylation array data from Jusakul et al and identify differentially methylated genes between samples classified by eCCS (in step 1.3.1). We also add in a little analysis of IDH1+ samples and find patterns suggestive that within CCS-A, these samples might be doing their own thing (as expected from methylation of IDH1+ samples).

```{r}

jusakul_methylation.meta <-  fread(file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW/jusakul_methylation.txt'))

jusakul_methylation.meta$barcode <- 
  paste(
    jusakul_methylation.meta$V2,
    jusakul_methylation.meta$V3,
    jusakul_methylation.meta$V4,
    sep='_'
  )

jusakul_methylation.meta$icgc_id <- 
  paste(
    jusakul_methylation.meta$V8,
    jusakul_methylation.meta$V9,
    jusakul_methylation.meta$V10,
    sep='_'
  )

jusakul.predicted.w.score <-  fread(file.path(data.dir, "/results/jusakul.eCCS.txt"))
names(jusakul.predicted.w.score)[1] <- 'icgc_id'

jusakul_methylation.meta <- full_join(jusakul_methylation.meta, 
                                      jusakul.predicted.w.score, by=c('icgc_id'='icgc_id')) %>% 
  filter(!is.na(barcode) & !is.na(rna_class) & V6 != 'normal' ) 


# Set the path to your .idat files
idat_dir <- file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW')

unzip.files =F
if(unzip.files){
  gz_files <- list.files(idat_dir, pattern = "\\.idat\\.gz$", full.names = TRUE)
  
  # Unzip each file
  for (f in gz_files) {
    R.utils::gunzip(f, remove = TRUE)  # remove = TRUE deletes the .gz file after extraction
  }
  
}


RGSet <- read.metharray.exp(base = idat_dir, recursive = TRUE)
RGSet <- RGSet[,jusakul_methylation.meta$barcode]
dim(RGSet)

#MSet <- preprocessRaw(RGSet)  

normSet <- preprocessQuantile(RGSet)

betaVals <- getBeta(normSet)

betaVals.df <- cbind.data.frame( betaVals)

jusakul.methyl.pca <- PCA(t(betaVals), graph = F)
jusakul.methyl.sites <- jusakul.methyl.pca$var$coord
jusakul.methyl.sample <- cbind.data.frame('barcode'=rownames(jusakul.methyl.pca$ind$coord),
                                          jusakul.methyl.pca$ind$coord)

jusakul.methyl.sample <- left_join(jusakul.methyl.sample, 
                                   jusakul_methylation.meta, by=c('barcode'='barcode'))

ggplot(jusakul.methyl.sample, aes(x=Dim.1, y=Dim.2, color=rna_class)) + 
         geom_point(size=2) + theme_bw(base_size=10) +
         
         stat_ellipse(data = ~ subset(., !is.na(rna_class) ), linetype='dashed')+
         scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + 
         theme(    plot.margin = unit(c(0, 0, 0, 0), "mm"),
                   panel.background = element_blank(),
                   plot.background = element_blank(),
                   
                   panel.grid = element_blank()) + coord_fixed() + 
         labs(x='Jusakul et al. Methylation PC1',y='Jusakul et al. Methylation PC2') + guides(color='none')

```

## tSNE

```{r}

tsne_result <- Rtsne(t(betaVals), dims = 2, perplexity = 30, verbose = TRUE, max_iter = 500)
tsne.dims <- as.data.frame(tsne_result$Y)
names(tsne.dims) <- c('tsne.dim.1','tsne.dim.2')
tsne.dims$barcode <- colnames(betaVals)

tsne.dims <- left_join(tsne.dims, jusakul.methyl.sample, by=c('barcode'='barcode'))

jusakul_muts <- fread(file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/data_mutations.txt'))
jusakul_drivers <- jusakul_muts %>% filter(Hugo_Symbol %in% c('BAP1','TP53','FGFR2','IDH1','SMAD4')) %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, Consequence)

tsne.dims$IDH1 <- 'wt'
tsne.dims$IDH1[tsne.dims$icgc_id %in% jusakul_drivers$Tumor_Sample_Barcode[jusakul_drivers$Hugo_Symbol == 'IDH1']] <- 'IDH1+'

ggplot(tsne.dims, aes(x=tsne.dim.1, y=tsne.dim.2, color=rna_class)) + 
         geom_point(size=2, aes(shape=IDH1)) + theme_bw() +
         scale_shape_manual(values = c('wt'=1,'IDH1+'=16)) +
         
         scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + theme(panel.grid = element_blank())

```

## explain variance

```{r}

jusakul_sample <- fread(file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_sample.txt'), skip=4)
jusakul_patient <- fread(file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/data_clinical_patient.txt'), skip=4)

jusakul_patient <- full_join(jusakul_patient, jusakul_sample, by=c('PATIENT_ID'='PATIENT_ID'))
jusakul_patient <- jusakul_patient %>% filter(ANATOMICAL_SUBTYPE != 'N/A' & SAMPLE_ID %in% tsne.dims$icgc_id[!is.na(tsne.dims$rna_class)] )
table(jusakul_patient$ANATOMICAL_SUBTYPE)

jusakul_methylation.pat <- left_join(jusakul_patient, jusakul_methylation.meta, by=c('SAMPLE_ID'='icgc_id'))


#TODO: subset to patients with all info (CCS, location and methylation)

locations <- jusakul_methylation.pat$ANATOMICAL_SUBTYPE
names(locations) <- jusakul_methylation.pat$SAMPLE_ID

rna_class <- jusakul_methylation.pat$rna_class
names(rna_class) <- jusakul_methylation.pat$SAMPLE_ID

betas.t <- t(betaVals)
betas.t.filt <- betas.t[jusakul_methylation.pat$barcode,]

betas_rda <- rda(betas.t.filt, scale = TRUE)

eig.val <- jusakul.methyl.pca$eig

# Test fit on top 2 axes
r.tab = NULL
for(i in c(1:5)){

  fit2_f1 <- envfit(betas_rda, locations, permutations = 999, choices = i:i)
  fit2_f2 <- envfit(betas_rda, rna_class, permutations = 999, choices = i:i)

  this.r.tab <- rbind.data.frame ( cbind('i'=as.numeric(i),
               'f'='anatomy',
               'e'=as.numeric(eig.val[i,2]),
               'r'=unlist(fit2_f1$factors$r),
               'p'=unlist(fit2_f1$factors$pvals)
  ),
  cbind('i'=as.numeric(i),
        'f'='CCS',
        'e'=as.numeric(eig.val[i,2]),
        'r'=unlist(fit2_f2$factors$r),
        'p'=unlist(fit2_f2$factors$pvals)
  )
  )
  
  if(is.null(r.tab)){
    r.tab <- this.r.tab
  } else{
    r.tab <- rbind.data.frame(r.tab,this.r.tab)
  }
  
}

r.tab$p <- as.numeric(r.tab$p)
r.tab$r <- as.numeric(r.tab$r)
r.tab$e <- as.numeric(r.tab$e)

r.tab$p.sig <- NA
r.tab$p.sig[r.tab$p < 0.05] <- r.tab$p[r.tab$p < 0.05]

r.tab$pc.i <- paste0('PC',r.tab$i)
r.tab$pc.i <- factor(r.tab$pc.i, levels=rev(paste0('PC',c(1:5)) ))
r.tab


methyl.rda.stack.plot <- 
ggplot(r.tab %>% filter(p < 0.05), aes(y=f , x=r*e)) + 

  geom_bar(aes(fill=pc.i), stat='identity')+
    geom_text(
    aes(label = pc.i), color='white',
    position = position_stack(vjust = 0.5), size=2,
  ) +
  theme_bw(base_size=10)+
  theme(
    plot.margin = unit(c(0, 2, 0, 2), "mm"),
    panel.background = element_blank(),
    #panel.border = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    legend.position = 'bottom',
    
  ) + labs(x='Variance explained (%)')+

    scale_fill_manual(  values=c(
      'PC1'='black',
      'PC2'='grey20',
      'PC3'='grey40',
      'PC4'='grey60',
      'PC5'='grey80'
    
  ), na.value = 'cornflowerblue') +
  expand_limits(x = max(r.tab$r * r.tab$e, na.rm = TRUE) * 1.1) + 
  guides(fill='none')

methyl.rda.stack.plot

ggsave(file.path(data.dir,"/results/fig/FIG3B.methyl_rda.stack.svg"),plot =
  methyl.rda.stack.plot       
, device = "svg", height = 18, width = 50 , units='mm')

```

### compare to Jusakul classes

```{r}

library(readxl)
library(ggalluvial)

jusakul.S3A <- read_excel(file.path(data.dir,
                                    '/source_data/cbioportal/chol_icgc_2017/21598290cd170368-sup-181437_2_supp_4118307_1ryl1l.xlsx'), 
                          sheet = 1, skip = 1)

jusakul.mVAF <- jusakul.S3A %>% group_by(Tumor) %>% 
  summarise(mean_vaf = mean(`Variant frequency`))

content.cutoff <- median(jusakul.mVAF$mean_vaf, na.rm=T)

jusakul.mVAF$content.bin <- NA
jusakul.mVAF$content.bin[jusakul.mVAF$mean_vaf >= content.cutoff] <- '>0.2'
jusakul.mVAF$content.bin[jusakul.mVAF$mean_vaf < content.cutoff] <- '<0.2'

ggplot(jusakul.mVAF, aes(x=mean_vaf, fill=content.bin)) + 
  geom_histogram(position=position_dodge()) + theme_bw() + 
  geom_vline(xintercept = content.cutoff)+
   scale_fill_manual(  
     values =  pnw_palette(name="Sailboat", n=4 )
     , na.value = 'white') 

jusakul.S1A <- read_excel(file.path(data.dir,
                                    '/source_data/cbioportal/chol_icgc_2017/21598290cd170368-sup-181437_2_supp_4118310_krylkl.xlsx'), 
                          sheet = 1, skip = 1)

jusakul.S1A[jusakul.S1A == 'N/A'] <- NA

jusakul.S1A <- full_join(jusakul.S1A, tsne.dims, by=c('Sample ID'='icgc_id')) %>% 
  full_join(jusakul.mVAF, by=c('Sample ID'='Tumor'))

transition_df.1 <- jusakul.S1A %>% 
  filter(!is.na(`Expanded cluster`) & !is.na(rna_class)  ) %>% 
  group_by(`Expanded cluster`, rna_class) %>% tally()

colnames(transition_df.1) <- c("Jusakul", "CCS","Freq")

transition.max <- transition_df.1 %>% group_by(Jusakul) %>% 
  filter(Freq == max(Freq) ) 

transition_df.1$Jusakul.f <- factor(transition_df.1$Jusakul, 
                                  levels=transition.max$Jusakul[order(transition.max$CCS)])


ggplot(transition_df.1,
          aes(axis1 = CCS, axis2 = Jusakul.f, y = Freq)) +
   geom_alluvium(width=0.8,  aes(fill=Jusakul.f)) +
   geom_stratum(width=0.8, color = "black", aes(fill=Jusakul.f), alpha=0.2) +
   geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=3) +
   scale_x_discrete(limits = c("CCS", "Jusakul"), expand = c(.05, .05)) +
   scale_fill_manual(  
     values =  pnw_palette(name="Shuksan", n=4 )
     , na.value = 'white') +
   theme_bw(base_size=10) + 
   guides(fill='none')+
   labs(fill='STIM', y='Sample Count')+
   theme(
     plot.margin = unit(c(0, 0, 0, 0.5), "mm"),
     panel.spacing = unit(0, "mm"),
     panel.background = element_blank(),
     panel.border = element_blank(),
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     strip.background = element_blank(),
   )

```

## Differential Methylation

```{r}

pheno <- jusakul_methylation.meta %>% filter(!is.na(barcode) ) %>% 
  dplyr::select(barcode, icgc_id, V6, rna_class, confidence.polarized) %>% DataFrame()
rownames(pheno) <- pheno[,1]

pData(normSet) <- pheno

normSet_filtered <- normSet[ , !is.na(pData(normSet)$rna_class) ]

betaVals <- getBeta(normSet_filtered)

group <- factor(pData(normSet_filtered)$rna_class)  
design <- model.matrix(~ group)

fit <- lmFit(betaVals, design)
fit <- eBayes(fit)

topTable <- topTable(fit, coef = 2, number = Inf, adjust = "BH")
topTable$IlmnID <- rownames(topTable)

Gene_features <- fread(file.path(data.dir,'/source_data/cbioportal/chol_icgc_2017/GSE89803_RAW/GPL16304_Gene_features_PlatformTable.txt'))

topTable.ft <- full_join(topTable, Gene_features, by=c('IlmnID'='IlmnID'))

top.gene.mthyl <- topTable.ft %>% filter(IlluminaGene != '')  %>%
  group_by(IlluminaGene) %>%
  slice_max(order_by = abs(logFC), n = 1, with_ties = FALSE) %>%
  ungroup()

multiple.testing <- nrow(top.gene.mthyl)

top.gene.mthyl$eccs <- NA
top.gene.mthyl$eccs[top.gene.mthyl$IlluminaGene %in% classifier$TSPs[,1]] <- 'CCS-A'
top.gene.mthyl$eccs[top.gene.mthyl$IlluminaGene %in% classifier$TSPs[,2]] <- 'CCS-B'

jusakul.diffmethyl.plot <- 
ggplot(top.gene.mthyl , aes(x=logFC, y=-log(adj.P.Val ))) + 
  geom_hline(yintercept = -log(0.05 ), color='grey',linetype='dotted')+
  
     geom_point( data = ~ subset(., adj.P.Val >= (0.05 ) & IlluminaGene %in% sample(top.gene.mthyl$IlluminaGene, round(nrow(top.gene.mthyl) * 0.1), replace = F )),  color='grey50') + 

     geom_point( data = ~ subset(.,  adj.P.Val < (0.05 ) & logFC > 0 & IlluminaGene %in% sample(top.gene.mthyl$IlluminaGene, round(nrow(top.gene.mthyl) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(1)]) + 
     geom_point( data = ~ subset(.,  adj.P.Val < (0.05 ) & logFC < 0 & IlluminaGene %in% sample(top.gene.mthyl$IlluminaGene, round(nrow(top.gene.mthyl) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(10)]) + 
         
  geom_label_repel(  data = ~ subset(., IlluminaGene %in% c(CMSA_MARKERS,
                                                         CMSB_MARKERS) & adj.P.Val < 0.05 ), 
                     aes(label=IlluminaGene, fill=eccs, color=eccs), 
                     size = 1.8,   force = 0.5, label.padding = 0.1,
                     max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  
  geom_point( data = ~ subset(., IlluminaGene %in% CMSA_MARKERS & adj.P.Val < 0.05 ),   
              color='black', fill=pnw_palette(name="Bay", n=19 )[c(1)], shape=21) + 
  geom_point( data = ~ subset(., IlluminaGene %in% CMSB_MARKERS & adj.P.Val < 0.05 ),   
              color='black', fill=pnw_palette(name="Bay", n=19 )[c(10)], shape=21) + 
  scale_fill_manual(values=c("CCS-A"="#04597C", "CCS-B"="#EDA71A"), na.value = 'white') +
  scale_color_manual(values=c("CCS-A"="white", "CCS-B"="white"), na.value = 'black') +
  labs(x='Differential methylation',y='adj. p-value (-log)')+
  guides(color='none',fill='none')+
  theme_bw(base_size=10) +
  theme(plot.background = element_blank(), 
        panel.background = element_blank(), 
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
        panel.grid =  element_blank()) + 
  xlim(-max(abs(top.gene.mthyl$logFC)),max(abs(top.gene.mthyl$logFC)))

jusakul.diffmethyl.plot

overwrite.plot = T
if(overwrite.plot){
  ggsave(file.path(data.dir, "/results/fig/FIG3D.jusakul.diffmethyl.svg"),plot =
    jusakul.diffmethyl.plot
  , device = "svg",width = 65, height = 45, units = 'mm' )
}

```

## Protein expression from Dong et al (2022)

```{r}

dong.predicted.w.score <-  fread( file.path(data.dir, "/results/dong.eCCS.txt"), )
rownames(dong.predicted.w.score) <- dong.predicted.w.score[[1]]
dong.predicted.w.score[[1]] <- NULL
dong.predicted.w.score <- dong.predicted.w.score %>% filter(!is.na(rna_class))

dong.prot.info <- read_excel(file.path(data.dir,"/source_data/dong_2022_S1.xlsx"), sheet = 5, skip = 2)
names(dong.prot.info)[c(1:2)] <- c('genes','protein')
names(dong.prot.info)[-c(1:2)] <- paste0('DONG_',names(dong.prot.info[,-c(1:2)]))
dong.prot.info[dong.prot.info == 'NA'] <- NA

dong.prot.mat <- as.matrix(dong.prot.info[,-c(1:2)])
dong.prot.mat <- matrix(as.numeric(dong.prot.mat), nrow = nrow(dong.prot.mat), ncol = ncol(dong.prot.mat))

dong.prot.pca <- PCA(t(dong.prot.mat), graph = F)
dong.prot.sample <- data.frame(dong.prot.pca$ind$coord)
dong.prot.sample$PatientID <- names(dong.prot.info[,-c(1:2)])

dong.prot.sample <- left_join(dong.prot.sample, dong.predicted.w.score, by=c('PatientID'='PatientID'))

ggplot(dong.prot.sample ,aes(x=Dim.1, y=Dim.2, color=rna_class)) + 
  geom_point( size=2) + 
  theme_bw(base_size=10) +
  stat_ellipse(data = ~ subset(., !is.na(rna_class) ), linetype='dashed')+
  scale_color_manual(values=c("#04597C", "#EDA71A"), na.value = 'grey80') + 
  theme(    plot.margin = unit(c(0, 0, 0, 0), "mm"),
            panel.background = element_blank(),
            plot.background = element_blank(),
            panel.grid = element_blank()) + 
  guides(color='none') + coord_fixed() + labs(x='Dong et al. Protein PC1', y='Dong et al. Protein PC2')

cms.prot.dong.ids <- unique(intersect(names(dong.prot.info), dong.predicted.w.score$PatientID))

normalized_data.p <- dong.prot.info %>% dplyr::select(all_of(cms.prot.dong.ids))

dong.predicted.filt.p <- dong.predicted.w.score %>% 
  filter(PatientID %in% cms.prot.dong.ids) %>%
  arrange(match(PatientID, cms.prot.dong.ids))

group <- factor(gsub('-','',dong.predicted.filt.p$rna_class))
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

normalized_data.complete <- as.matrix(normalized_data.p)
normalized_data.complete <- apply(normalized_data.complete, 2, as.numeric)
rownames(normalized_data.complete) <- dong.prot.info$genes

fit <- lmFit(normalized_data.complete, design)
ncol(normalized_data.complete)
nrow(design)

contrast.matrix <- makeContrasts(
  TreatmentVsControl = CCSA - CCSB,  
  levels = design
)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

results <- topTable(fit2, coef = 1, number = Inf, adjust.method = "BH")

results$gene_name <- results$ID
multiple.testing <- nrow(results) 

results$eccs <- NA
results$eccs[results$gene_name %in% classifier$TSPs[,1]] <- 'CCS-A'
results$eccs[results$gene_name %in% classifier$TSPs[,2]] <- 'CCS-B'

dong.diffprot.plot <- 
     ggplot(results , aes(x=logFC, y=-log(adj.P.Val ))) + 
         geom_hline(yintercept = -log(0.05 ), color='grey',linetype='dotted')+
         
     geom_point( data = ~ subset(., adj.P.Val >= (0.05 ) & gene_name %in% sample(results$gene_name, round(nrow(results) * 0.1), replace = F )),  color='grey50') + 

     geom_point( data = ~ subset(.,  adj.P.Val < (0.05 ) & logFC > 0 & gene_name %in% sample(results$gene_name, round(nrow(results) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(1)]) + 
     geom_point( data = ~ subset(.,  adj.P.Val < (0.05 ) & logFC < 0 & gene_name %in% sample(results$gene_name, round(nrow(results) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(10)]) + 
         
         
         geom_label_repel(  data = ~ subset(., gene_name %in% c(CMSA_MARKERS,
                                                                CMSB_MARKERS) & adj.P.Val < 0.05 ), 
                            aes(label=gene_name, fill=eccs, color=eccs), 
                            size = 1.8,   force = 0.5, label.padding = 0.1,
                            max.time = 4,  min.segment.length = 0, max.overlaps =15) +
         
         geom_point( data = ~ subset(., gene_name %in% CMSA_MARKERS & adj.P.Val < 0.05 ),   
                     color='black', fill=pnw_palette(name="Bay", n=19 )[c(1)], shape=21) + 
         geom_point( data = ~ subset(., gene_name %in% CMSB_MARKERS & adj.P.Val < 0.05 ),   
                     color='black', fill=pnw_palette(name="Bay", n=19 )[c(10)], shape=21) + 
           scale_fill_manual(values=c("CCS-A"="#04597C", "CCS-B"="#EDA71A"), na.value = 'white') +
           scale_color_manual(values=c("CCS-A"="white", "CCS-B"="white"), na.value = 'black') +
         guides(color='none')+
         labs(x='Differential protein expression',y='adj. p-value (-log)')+
         guides(color='none',fill='none')+
         theme_bw(base_size=10) +
         theme(plot.background = element_blank(), 
               panel.background = element_blank(), 
               plot.margin = unit(c(0, 0, 0, 0), "mm"),
               
               panel.grid =  element_blank()) + 
  xlim(-max(abs(results$logFC)),max(abs(results$logFC)))

dong.diffprot.plot

ggsave(file.path(data.dir, "/results/fig/FIG3e.dong.diffprot.svg"),plot =
         dong.diffprot.plot
   , device = "svg",width = 65, height = 45, units = 'mm' )


```


