# Variance and bias in gene expression
We've established that CCS plays a role in survival across multiple datasets, and has something to do with anatomical location of the primary tumor. Let's dive further into the whole genome expression data to find out more about this.

## Quantify variance by location and CCS
Here, we analytically compare the variance explained by anatomical location and compare it to variance explained by CCS.

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(DESeq2)
  library(FactoMineR)
  library(vegan)
  library(cowplot)
  library(ggrepel)
  library(ggplot2)
  library(PNWColors)
})

`%ni%` <- Negate(`%in%`)

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'
source( file.path(code.dir, "bin/BTC.functions.R"))

gene_set <- fread(file.path(data.dir,'/source_data/fgs.txt'))
classifier <- readRDS(file.path(data.dir,'/results/tps.classifier.rds'))

predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt') )
predicted.w.score$rna_class[is.na(predicted.w.score$rna_class)] <- 'Other'

sample.info <- fread(file.path(data.dir,"/results/samples.info.txt"))

sample.typed <- left_join(sample.info, predicted.w.score)

this_tpm <- fread(file.path(data.dir,"/results/rna.tpm.ff.txt")) 

tpm.t <- t(this_tpm[,-c(1)])

```

## Run PCA

```{r}
pca <- PCA(tpm.t, graph = F, ncp = 10)

pca_labs <- cbind.data.frame('Sample'=names(this_tpm[,-c(1)]),
                             pca$ind$coord
)

pca_labs <- left_join(pca_labs, sample.typed)

exp.pca.plot <- 
ggplot(pca_labs  ,aes(x=Dim.1,y=Dim.3)) +
  
  stat_ellipse( geom = "polygon", level=0.95, aes(fill = rna_class), alpha=0.1) +
  
  geom_point(size=1, shape=21, stroke = 1, aes(color=location,fill=rna_class)) + 
     

  labs(
    x=paste("PC1 (",round(pca$eig[1,2],digits=2),"%)",sep=""),
    y=paste("PC3 (",round(pca$eig[3,2],digits=2),"%)",sep="")) +
  theme_bw(base_size=10) + 

  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    panel.background = element_blank(),
    panel.border = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = 'top'
  ) +

  scale_fill_manual(  values=c(
    "CCS-B"    = "#EDA71A" , 
    "CCS-A" = "#04597C" ,
    "Other" ="white"
    
  ), na.value = 'black') +
  scale_color_manual(  values=c(
    "CCS-B"    = "#EDA71A" , 
    "CCS-A" = "#04597C" ,
     "Other" ="white",
    'HCC-CCA' = "#b4b4b4ff",
    'Perihilar' = "#e59d54ff",
    'Intrahepatic' = "#fbcc7aff",
    'Distal' = "#ba7a44ff",
    'Gallbladder' = "#855b40ff"
    
  ), na.value = 'black')

ggsave(file.path(data.dir,"/results/fig/FIGS3A.exp.pca.plot.svg"),plot =
  exp.pca.plot       
, device = "svg", height = 80, width = 70 , units='mm')

exp.pca.plot   + geom_label_repel(  data = ~ subset(., rna_class %in% c('Other') ), 
                    aes(label=Sample), 
                    size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  
                    min.segment.length = 0, max.overlaps =15) 


```

## Run RDA

```{r}

sample.filt <- sample.typed %>% filter(!is.na(location) & 
                                          !is.na(rna_class) & 
                                          !is.na(Site_of_tissue))

location <- sample.filt$location
names(location)  <- sample.filt$Sample
rna_class <- sample.filt$rna_class
names(rna_class)  <- sample.filt$Sample
site <- sample.filt$Site_of_tissue
names(site)  <- sample.filt$Sample

these.samples <- intersect(names(this_tpm), sample.filt$Sample)
 
this_tpm <- this_tpm %>% 
  dplyr::select(all_of(c('gene_name', 
                        these.samples)))

location <- location[these.samples]
rna_class <- rna_class[these.samples]
site <- site[these.samples]

# RDA identical ordination to PCA when centered/scaled
tpm.t <- t(this_tpm[,-c(1)])

r.tab <- test_rda(tpm.t, location, rna_class, site)

var.sum <- r.tab %>% filter(!is.na(p.sig)) %>% group_by(f) %>% summarise(var_sum = sum(var))
var.sum[1,2] / var.sum[2,2]

exp.rda.stack.plot <- 
ggplot(
  r.tab %>% filter(p < 0.05 & factor != "site"),
  aes(y = factor, x = r * e)
) +
  geom_bar(aes(fill = pc.i), stat = "identity") +
  geom_text(
    aes(label = pc.i), color='white',
    position = position_stack(vjust = 0.5), size=2,
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.margin = unit(c(0, 2, 0, 2), "mm"),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) + labs(x='Variance explained (%)')+
  scale_fill_manual(
    values = c(
      PC1 = "black",
      PC2 = "grey20",
      PC3 = "grey40",
      PC4 = "grey60",
      PC5 = "grey80"
    ),
    na.value = "cornflowerblue"
  ) +
  expand_limits(x = max(r.tab$r * r.tab$e, na.rm = TRUE) * 1.1) + 
  guides(fill='none')

exp.rda.stack.plot

ggsave(file.path(data.dir,"/results/fig/FIG3A.exp_rda.stack.svg"),plot =
  exp.rda.stack.plot       
, device = "svg", height = 18, width = 50 , units='mm')

## group by eCCA

location[location == "Perihilar" | location == "Distal" ] <- "Extrahepatic"

r.tab.e <- test_rda(tpm.t, location, rna_class, site)

var.sum.e <- r.tab.e %>% filter(!is.na(p.sig)) %>% group_by(f) %>% summarise(var_sum = sum(var))
var.sum.e[1,2] / var.sum.e[2,2]

```


## Differential Gene Expression
Markers assembled from manual review of the literature.

```{r}

good.genes <- gene_set$gene_name[gene_set$Chr %in% c(1:22) & 
                                   gene_set$biotype == 'protein_coding' & 
                                   gene_set$Length > 500]

dup.genes <- table(good.genes)[table(good.genes) > 1]
good.genes <- good.genes[good.genes %ni% names(dup.genes)]

rna_raw <-  fread(file.path(data.dir,"/results/rna.htseq.all.txt"))
this_rna <- as.matrix(rna_raw[,-1])
rownames(this_rna) <- rna_raw$gene_name
this_rna <- this_rna[good.genes,]

confident.samples <- intersect(colnames(this_rna), sample.typed$Sample)
confident.rna.ff <- this_rna[,confident.samples]

run.deseq = F
if(run.deseq){
  
  cms_labels <- gsub('-','',sample.typed$rna_class)
  names(cms_labels) <- sample.typed$Sample
  cms_labels <- cms_labels[confident.samples]

  expdesign <-  as.matrix(as.factor(cms_labels))
  colnames(expdesign) <- 'btc'

  cds = DESeqDataSetFromMatrix(countData = confident.rna.ff, 
                               colData = expdesign, 
                               design = ~ btc)
  
  cds <- DESeq(cds)
  
  res <- results(cds, contrast = c("btc", c("CCSA","CCSB")))
  
  res.df <- as.data.frame(res)
  res.df$gene_name <- rownames(res.df)
  
  res.df <- left_join(res.df, gene_set, by=c('gene_name'='gene_name'))
  
  write.table( res.df,  
               file = file.path(data.dir,'/results/cms.deseq.txt'), 
               row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
} else {
  res.df <- fread(file.path(data.dir,'/results/cms.deseq.txt'))
}

multiple.testing <- nrow(res.df) 

CMSA_MARKERS = c( 'DCDC2','BICC1','ONECUT1',
                  'PROX1','HNF1B','SPP1', 
                  'CFTR','NCAM1')
  
               #  'HNF4A','ELF3','IGFBP2','SLC3A1','CD24','PROM1','TNFSF12',
               #  'TNFRSF12A','GPC3','ARG1','CPS1','TTR','APOH',  'HHEX', 'SCTR','ALB' 

CMSB_MARKERS = c('PTGS2','S100P', 'MKI67', 
                  'CDK1', 'CDX2','CD44',
                 'KRT7', 'KRT19', 'KRT20',
                 'MUC2','MUC5AC','MUC17')
                 
                 #'SALL4','SSTR1','CEACAM5', 
                 

res.df$eccs <- NA
res.df$eccs[res.df$gene_name %in% classifier$TSPs[,1]] <- 'CCS-A'
res.df$eccs[res.df$gene_name %in% classifier$TSPs[,2]] <- 'CCS-B'

diffexp.plot <- 
   ggplot(res.df %>% filter(biotype == 'protein_coding' ), aes(x=log2FoldChange, y=-log(padj ))) + 
      geom_hline(yintercept = -log(0.05 ), color='grey',linetype='dotted')+
     
     geom_point( data = ~ subset(., padj >= (0.05 ) & gene_name %in% sample(res.df$gene_name, round(nrow(res.df) * 0.1), replace = F )),  color='grey50') + 

     geom_point( data = ~ subset(.,  padj < (0.05 ) & log2FoldChange > 0 & gene_name %in% sample(res.df$gene_name, round(nrow(res.df) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(1)]) + 
     geom_point( data = ~ subset(.,  padj < (0.05 ) & log2FoldChange < 0 & gene_name %in% sample(res.df$gene_name, round(nrow(res.df) * 0.75), replace = F )),   
                 color=pnw_palette(name="Bay", n=19 )[c(10)]) + 
         
     geom_label_repel(  data = ~ subset(., gene_name %in% c(CMSA_MARKERS,
                                                            CMSB_MARKERS) & padj < (0.001 )), 
                        aes(label=gene_name, fill=eccs, color=eccs), 
                        size = 1.8,   force = 0.5, label.padding = 0.1,
                        max.time = 4,  min.segment.length = 0, max.overlaps =15) +
     
     geom_point( data = ~ subset(., gene_name %in% CMSA_MARKERS & padj < (0.001 )),   
                 color='black', fill=pnw_palette(name="Bay", n=19 )[c(1)], shape=21) + 
     geom_point( data = ~ subset(., gene_name %in% CMSB_MARKERS & padj < (0.001 )),   
                 color='black', fill=pnw_palette(name="Bay", n=19 )[c(10)], shape=21) + 
     
     guides(color='none', fill='none')+
     theme_bw(base_size=10) +
     theme(plot.background = element_blank(), 
           panel.background = element_blank(), 
           plot.margin = unit(c(0, 0, 0, 0), "mm"),
           
           panel.grid =  element_blank()) +
    scale_fill_manual(values=c("CCS-A"="#04597C", "CCS-B"="#EDA71A"), na.value = 'white') +
    scale_color_manual(values=c("CCS-A"="white", "CCS-B"="white"), na.value = 'black') +
    labs(x='Differential RNA expression', y='adj. p-value (-log)')+ 
    xlim(-max(abs(res.df$log2FoldChange), na.rm = T), max(abs(res.df$log2FoldChange), na.rm = T))+ 
    ylim(0, max(sort(-log(res.df$padj), decreasing = T)[-1], na.rm = T) )
   

ggsave(file.path(data.dir, "/results/fig/FIG3C.diff_exp.svg"),plot =
  diffexp.plot
   , device = "svg",width = 65, height = 45, units = 'mm' )

diffexp.plot


```

