# Explore Immune signatures

## NMF Immune signatures
First, let's look at the correlation between our immune signatures and their relationship with CCS. We'll find the strongest signal is from variance in tumor content in the sample.

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(reshape2)
  library(ggplot2)
  library(PNWColors)
  library(ggalluvial)
})


basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

H_matrix <- read.csv(file.path(basedir,'/results/H_matrix.txt'), row.names = 1)

H_matrix_ordered_norm <-   data.frame(t(apply(H_matrix, 1, scale)))
names(H_matrix_ordered_norm) <- names(H_matrix)

H_matrix_ordered_norm.df <- data.table(H_matrix_ordered_norm)
H_matrix_ordered_norm.df$sig.id <- paste0('sig', row.names(H_matrix_ordered_norm))

H_matrix_ordered_norm.melt <- melt(H_matrix_ordered_norm.df, id.vars='sig.id') 

y_order.convert <- fread(file.path(basedir,'/results/NMF.conversion.txt'))

h.new <- left_join(H_matrix_ordered_norm.melt, y_order.convert, by=c('sig.id'='og')) %>%
  reshape2::dcast( variable ~ new, value.var = 'value')

cohort.info <- fread(file.path(basedir,'/results/btc.cellularity.curated.txt'), col.names = c('tumour','cellularity','ploidy'))

h.new <- left_join(h.new, cohort.info, by=c('variable'='tumour'))
h.new$content.bin <-NA
h.new$content.bin[h.new$cellularity <= 1 ] <- 'High (0.6-1)'
h.new$content.bin[h.new$cellularity < 0.6 ] <- 'Low (<0.6)'

predicted.w.score <- fread(file.path(basedir,"/results/all.eCMS.txt")) 
H_t_labs <- left_join(predicted.w.score, h.new,  by=c('sample'='variable'))

immune_nmf.tumour_content.plot <- 
         ggplot(H_t_labs %>% filter(!is.na(content.bin)) , aes(x=sig.2, y=sig.6, color=rna_class, shape=content.bin)) + 
  facet_grid(content.bin~.)+
  geom_vline(xintercept = 0, color='grey80')+
  geom_hline(yintercept = 0, color='grey80')+
  
  geom_point(aes(size=cellularity))  + 
  coord_fixed() + 
  theme_bw(base_size=10)+
   
  scale_shape_manual(values=c(16,1)) + 
  #scale_color_manual(values = pnw_palette(name="Sailboat", n=5 ))+
  theme(
    plot.margin = unit(c(0, 0, 0, 1), "mm"),
    panel.spacing = unit(0, "mm"),
    
    panel.background = element_blank(),
    
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.key.height = unit(6, "pt"),
    legend.key.width  = unit(6, "pt"),
    legend.spacing.x  = unit(4, "pt"),
    legend.spacing.y  = unit(1, "pt"),
    legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0)
  ) +
  scale_color_manual(  values=c(
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" 
  ), na.value = 'grey90') +
    guides(color='none')+
    scale_size(range = c(0.25, 2.5)) + labs(x='Signature 2', y='Signature 6',shape='Cellularity Bin',size='Cellularity',color='CMS')

immune_nmf.tumour_content.plot

ggsave(paste0(basedir, "/results/fig/fig3f.immune_nmf.tumour_content.svg"),plot =
  immune_nmf.tumour_content.plot
, device = "svg", height = 70, width = 70, units = 'mm' )

```

## Stroma-Tumor-Immune classification
Next, we compare our immune signatures with the Stroma-Tumor-Immune (STIM) classification of [Martin-Serrano et al. 2023](http://dx.doi.org/10.1136/gutjnl-2021-326514). This classifier was trained to predict iCCAs so we subset to this population only. When we noticed the strong segregation specific STIM classes with CCS, we also enquired into whether tumor content might impact inference.

```{r}

labels_df.raw <- fread(file.path(basedir,"/results/classified.FF.txt"))
stim.class <- labels_df.raw  %>% dplyr::select( Sample | starts_with("martinserrano") ) 
stim.class[stim.class == 'Hepati Stem-like'] <- 'Hepatic Stem-like'
  
stim.prob <- left_join(H_t_labs, stim.class, by=c('sample'='Sample')) %>% filter( !is.na(rna_class))

timeline.outcome <- fread(file.path(secdir,"/results/timeline.outcome.txt"))
timeline.outcome$BTCID <- gsub('_','',timeline.outcome$BTC_ID)
iccas <- timeline.outcome$BTCID[timeline.outcome$Location == 'Intrahepatic_CCA_iCCA']

stim.prob$content.bin <-NA
stim.prob$content.bin[stim.prob$cellularity <= 1 ] <- 'High (>=85%)'
stim.prob$content.bin[stim.prob$cellularity < 0.85 ] <- 'Mid (65-85%)'
stim.prob$content.bin[stim.prob$cellularity <= 0.65 ] <- 'Low (<=65%)'


transition_df.1 <- stim.prob %>% filter(!is.na(martinserrano.class) &  !is.na(rna_class) & BTCID %in% iccas) %>% group_by(martinserrano.class, rna_class) %>% tally()

colnames(transition_df.1) <- c("STIM", "CMS", "Freq")

transition.max <- transition_df.1 %>% group_by(STIM) %>% filter(Freq == max(Freq) ) 

transition_df.1$STIM.f <- factor(transition_df.1$STIM, 
                                  levels=transition.max$STIM[order(transition.max$CMS)])


ggplot(transition_df.1,
          aes(axis1 = CMS, axis2 = STIM.f, y = Freq)) +
   geom_alluvium(width=0.8,  aes(fill=STIM.f)) +
   geom_stratum(width=0.8, color = "black", aes(fill=STIM.f), alpha=0.2) +
   geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=1.5) +
   scale_x_discrete(limits = c("CCS", "STIM"), expand = c(.05, .05)) +
   scale_fill_manual(  
     values =  pnw_palette(name="Sailboat", n=5 )
     , na.value = 'white') +
   theme_bw(base_size=10) + 
   guides(fill='none')+
   labs(fill='STIM', y='Sample Count')+
   theme(
     plot.margin = unit(c(0, 0, 0, 0.5), "mm"),
     panel.spacing = unit(0, "mm"),
     panel.background = element_blank(),
     panel.border = element_blank(),
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     strip.background = element_blank(),
   )

transition_df <- stim.prob %>% filter(!is.na(martinserrano.class) & !is.na(content.bin) & !is.na(rna_class) & BTCID %in% iccas) %>% group_by(martinserrano.class, rna_class, content.bin) %>% tally()

colnames(transition_df) <- c("STIM", "CMS","Cellularity", "Freq")

transition_df$STIM <- gsub(' ','\n',transition_df$STIM)
transition_df$STIM.f <- factor(transition_df$STIM, levels=c( 'Hepatic\nStem-like', 'Immune\nClassical', 'Inflammatory\nStroma', 'Desert-like','Tumor\nClassical'))
transition_df$Cellularity.f <- factor(transition_df$Cellularity, levels=c( 'High (>=85%)', 'Mid (65-85%)', 'Low (<=65%)'))

stim_cms.iCCA.plot <- 
         ggplot(transition_df,
                aes(axis1 = CMS, axis2 = STIM.f, y = Freq)) +
         facet_grid(Cellularity.f ~ . , space='free_y',scale='free_y') +
         geom_alluvium(width=0.8,  aes(fill=STIM.f)) +
         geom_stratum(width=0.8, color = "black", aes(fill=STIM.f), alpha=0.2) +
         geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=1.5) +
         scale_x_discrete(limits = c("CCS", "STIM"), expand = c(.05, .05)) +
         scale_fill_manual(  
           values =  pnw_palette(name="Sailboat", n=5 )
           , na.value = 'white') +
         theme_bw(base_size=10) + 
         guides(fill='none')+
         labs(fill='STIM', y='Sample Count')+
         theme(
           plot.margin = unit(c(0, 0, 0, 0.5), "mm"),
           panel.spacing = unit(0, "mm"),
           panel.background = element_blank(),
           panel.border = element_blank(),
           plot.background = element_blank(),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
         )

stim_cms.iCCA.plot
       
ggsave(file.path(basedir, "/results/fig/FIG3g.stim_cms.iCCA.svg"),plot =
         stim_cms.iCCA.plot
       , device = "svg", width = 37, height = 63, units = 'mm', dpi= 72 )


         

```