# Explore Immune signatures

## NMF Immune signatures
First, let's look at the correlation between our immune signatures and their relationship with CCS. We'll find the strongest signal is from variance in tumor content in the sample.

```{r}
suppressWarnings(
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(reshape2)
  library(ggplot2)
  library(PNWColors)
  library(ggalluvial)
})
)

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'

H_matrix <- read.csv(file.path(data.dir,'/results/whole.H_matrix.txt'), row.names = 1)

H_matrix_ordered_norm <-   data.frame(t(apply(H_matrix, 1, scale)))
names(H_matrix_ordered_norm) <- names(H_matrix)

H_matrix_ordered_norm.df <- data.table(H_matrix_ordered_norm)
H_matrix_ordered_norm.df$sig.id <- paste0('sig', row.names(H_matrix_ordered_norm))

H_matrix_ordered_norm.melt <- melt(H_matrix_ordered_norm.df, id.vars='sig.id') 

y_order.convert <- fread(file.path(data.dir,'/results/whole.NMF.conversion.txt'))

h.new <- left_join(H_matrix_ordered_norm.melt, y_order.convert, by=c('sig.id'='og')) %>%
  reshape2::dcast( variable ~ new, value.var = 'value')

```




```{r}

cohort.info <- fread(file.path(data.dir,"/results/samples.info.txt")) %>%
  dplyr::select(Sample, cellularity, location)

h.new <- left_join(cohort.info, h.new,  by=c('Sample'='variable'))
h.new$content.bin <-NA
h.new$content.bin[h.new$cellularity <= 1 ] <- 'High (>=85%)'
h.new$content.bin[h.new$cellularity < 0.85 ] <- 'Mid (65-85%)'
h.new$content.bin[h.new$cellularity <= 0.65 ] <- 'Low (<=65%)'
h.new$Cellularity.f <- factor(h.new$content.bin, levels=c( 'High (>=85%)', 'Mid (65-85%)', 'Low (<=65%)'))

predicted.w.score <- fread(file.path(data.dir,"/results/all.eCCS.txt")) 
H_t_labs <- left_join(predicted.w.score, h.new,  by=c('Sample'='Sample'))

immune_nmf.tumour_content.plot <- 
         ggplot(H_t_labs %>% filter(!is.na(Cellularity.f) & !is.na(rna_class)) , 
                aes(y=sig.5, x=sig.8)) + 
  facet_grid(Cellularity.f~.)+
  geom_vline(xintercept = 0, color='grey80')+
  geom_hline(yintercept = 0, color='grey80')+
  
  geom_density_2d(color='cornflowerblue')+
  geom_point(aes(size=cellularity, color=rna_class))  + 
  coord_fixed() + 
  theme_bw(base_size=10)+
   
  scale_shape_manual(values=c(16,1)) + 
  #scale_color_manual(values = pnw_palette(name="Sailboat", n=5 ))+
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "mm"),
    panel.spacing = unit(2, "mm"),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    #strip.text = element_blank(),
    legend.key.height = unit(6, "pt"),
    legend.key.width  = unit(6, "pt"),
    legend.spacing.x  = unit(4, "pt"),
    legend.spacing.y  = unit(1, "pt"),
    legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0)
  ) +
  scale_color_manual(  values=c(
    "CCS-B"    = "#EDA71A" , 
    "CCS-A" = "#04597C" 
  ), na.value = 'grey90') +
    guides(color='none',size='none')+
    scale_size(range = c(0.25, 1.5)) + 
  labs(y='Signature 5', x='Signature 8',
        size='Cellularity',color='CCS')+
  scale_x_continuous(position = "top")

immune_nmf.tumour_content.plot

ggsave(paste0(data.dir, "/results/fig/fig3g.immune_nmf.tumour_content.svg"),plot =
  immune_nmf.tumour_content.plot
, device = "svg", height = 70, width = 45, units = 'mm' )

```

## Stroma-Tumor-Immune classification
Next, we compare our immune signatures with the Stroma-Tumor-Immune (STIM) classification of [Martin-Serrano et al. 2023](http://dx.doi.org/10.1136/gutjnl-2021-326514). This classifier was trained to predict iCCAs so we subset to this population only. When we noticed the strong segregation specific STIM classes with CCS, we also enquired into whether tumor content might impact inference.

```{r}

labels_df.raw <- fread(file.path(data.dir,"/results/classified.FF.txt"))
stim.class <- labels_df.raw  %>% dplyr::select( Sample | starts_with("martinserrano") ) 
stim.class[stim.class == 'Hepati Stem-like'] <- 'Hepatic Stem-like'
  
stim.prob <- left_join(H_t_labs, stim.class, by=c('Sample'='Sample')) %>% filter( !is.na(rna_class))

transition_df <- stim.prob %>% 
  filter(!is.na(martinserrano.class) & 
         !is.na(content.bin) & 
         !is.na(rna_class) & 
         location == 'Intrahepatic') %>% 
  group_by(martinserrano.class, rna_class, content.bin) %>% tally() %>% ungroup()

colnames(transition_df) <- c("STIM", "CCS","Cellularity", "Freq")

transition.max <- 
  transition_df %>% 
  group_by(STIM) %>% 
  filter(Freq == max(Freq) ) %>% 
  slice_sample(n = 1) %>% 
  ungroup()

transition_df$STIM <- gsub(' ','\n',transition_df$STIM)
transition_df$STIM.f <- factor(transition_df$STIM, levels=c( 'Hepatic\nStem-like', 'Immune\nClassical', 'Inflammatory\nStroma', 'Desert-like','Tumor\nClassical'))
transition_df$Cellularity.f <- factor(transition_df$Cellularity, levels=c( 'High (>=85%)', 'Mid (65-85%)', 'Low (<=65%)'))

stim_cms.iCCA.plot <- 
         ggplot(transition_df,
                aes(axis1 = CCS, axis2 = STIM.f, y = Freq)) +
         facet_grid(Cellularity.f ~ . , space='free_y',scale='free_y',switch = 'y') +
         geom_alluvium(width=0.9,  aes(fill=STIM.f)) +
         geom_stratum(width=0.9, color = "black", aes(fill=STIM.f), alpha=0.2, linewidth=0.2) +
         geom_text(stat = "stratum", aes(label = after_stat(stratum)), lineheight = 0.7, size=1.75) +
         scale_x_discrete(limits = c("CCS", "STIM"), expand = c(.05, .05), position = "top") +
         scale_y_continuous(position = "right") +

         scale_fill_manual(  
           values =  pnw_palette(name="Moth", n=5 )
           , na.value = 'white') +
         theme_bw(base_size=10) + 
         guides(fill='none')+
         labs(fill='STIM', y='Sample Count')+
         theme(
           plot.margin = unit(c(0, 0, 0, 0.5), "mm"),
           panel.spacing = unit(0, "mm"),
           panel.background = element_blank(),
           panel.border = element_blank(),
           plot.background = element_blank(),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
         ) 

stim_cms.iCCA.plot
       
ggsave(file.path(data.dir, "/results/fig/FIG3h.stim_cms.iCCA.svg"),plot =
         stim_cms.iCCA.plot
       , device = "svg", width = 37, height = 65, units = 'mm', dpi= 72 )


         

```

## Pathologist-estimated Cellularity


```{r}

BTC_clinical.tumours <- fread(file.path(data.dir,'/source_data/BTC_clinical.tumours.tsv'), header=T)

btc.cellularity <- fread(file.path(data.dir,'/results/btc.cellularity.curated.txt'), col.names = c('Sample','cellularity','ploidy'), header=F)
BTC_clinical.tumours <- full_join(BTC_clinical.tumours, btc.cellularity, by=c('Sample'='Sample'))

sample.list <- fread(file.path(data.dir,'/source_data/sample.list.txt'), col.names = c('donor','tumour','normal'), header=F)

BTC_clinical.tumours$curated <- F
BTC_clinical.tumours$curated[BTC_clinical.tumours$Sample %in% sample.list$tumour] <- T

replicates <- BTC_clinical.tumours %>% group_by(Donor) %>% tally() %>% filter(n>1)
BTC_clinical.tumours$replicates <- F
BTC_clinical.tumours$replicates[BTC_clinical.tumours$Donor %in% replicates$Donor] <- T

BTC_clinical.tumours$curated[BTC_clinical.tumours$curated == F & BTC_clinical.tumours$replicates == T] <- 'replicate'

pathologist_cellularity <- fread(file.path(data.dir,"/source_data/MOHCCN_LBR_Tumour_Percentage.txt"))

pathologist_cellularity <- left_join(BTC_clinical.tumours, pathologist_cellularity, by=c('Sample'='tumor_ID'))

btc.cellularity.all <- fread(file.path(data.dir,'/source_data/btc.cellularity.pipeline.txt'), col.names = c('Sample','cellularity','ploidy'), header=F)
pathologist_cellularity <- left_join(pathologist_cellularity, btc.cellularity.all, by=c('Sample'='Sample'))

pathologist_cellularity$QC <- NA
pathologist_cellularity$QC[pathologist_cellularity$cellularity.y < 0.3] <- 'FAIL'
pathologist_cellularity$QC[pathologist_cellularity$cellularity.y >= 0.3] <- 'PASS'

pathologist_cellularity <- pathologist_cellularity %>% 
  arrange(`Group ID(LCM/non-LCM)`, tumour_percentage_numeric, cellularity.y)

pathologist.plot <- 
ggplot(pathologist_cellularity %>% filter(!is.na(QC)), 
       aes(y=cellularity.y*100, x=tumour_percentage_numeric)) + 
  geom_hline(yintercept = 30, color='red', linetype='dotted')+
  geom_abline(slope=1)+
  geom_point(aes(color=`Group ID(LCM/non-LCM)`, shape=QC))+
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()) +
    scale_color_manual(values = c("LCM" = "lightblue",
                                "no LCM" = "darkblue")) +
           scale_shape_manual(values = c(4,1)) +
  labs(x='Pathologist Estimate (%)',y='Bioinformatic Estimate (%)', color='') + coord_fixed()

pathologist.plot

ggsave(file.path(data.dir, "/results/fig/FIGS3e.pathologist.plot.svg"),plot =
         pathologist.plot
       , device = "svg", width = 90, height = 75, units = 'mm', dpi= 72 )


pathologist_cellularity.filt <- pathologist_cellularity %>% 
  filter(`Group ID(LCM/non-LCM)` == "no LCM")

cor.test(pathologist_cellularity.filt$tumour_percentage_numeric,
         pathologist_cellularity.filt$cellularity.y * 100)

```

