# Assemble and review Non-negative Matrix Factorization (NMF) 
This is another section where we do something a little unique. Non-negative Matrix Factorization (NMF) is great for decomposing factors but you a little bit need to know what you are looking for (there's lot of optimization). First, we need to know how many components to break the data into (k). For some applications, keeping k small is great, but for in silico microdissection, too small of a k will confound different cell types. But if we remove the constraint of minimizing k, then how do we pick k? What we decided to do is, since we want to isolate signal from the tumor from other cell types, we took our only orthogonal estimate of tumor content: the estimate from the WGS. While not expected to be exactly the same, they should be pretty close. So here we optimize k to maximize for components that correlate with WGS-estimate of tumor content. We do the same to choose the numbers of genes to include in the decomposition. 

## Load Data


```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(stringr)
  library(NMF)
  library(clusterProfiler)
  library(msigdbr)
  library(gprofiler2)
  library(enrichplot)
  library(pheatmap)
  library(ggdendroplot)
  #library(ggrepel)
  library(cowplot)
  library(PNWColors)
  library(ggalluvial)
})

packageVersion('gprofiler2')

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'
cluster.dir = '/Volumes/pcsi/users/fbeaudry/nmf_analysis/'

gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))

key_tumour_features <- fread(file.path(basedir,'/source_data/BTC_clinical.tumours.tsv'), header=T)
ff_tumours <- na.omit(key_tumour_features$Sample[key_tumour_features$`Sample Source(FFPE, FF)` == 'FF' ])

sample_list_file_path = file.path(basedir,'/source_data/sample.list.txt')
sample_list <- fread(sample_list_file_path, header=F,  col.names = c('donor','tumour','normal'))
sample_list <- sample_list %>% filter(tumour %in% ff_tumours)

cohort.info <- fread(file.path(basedir,'/results/btc.cellularity.curated.txt'), col.names = c('tumour','cellularity','ploidy'))
cohort.info <- left_join(sample_list, cohort.info, by=c('tumour'='tumour'))

clinic.sample.info <- fread(file.path(basedir,"/results/clinic.sample.info.txt"))
clinic.sample.info$BTCID <- gsub('_','',clinic.sample.info$BTC_ID)
cohort.info <- full_join(cohort.info, clinic.sample.info,  by=c('donor'='BTCID'))

predicted.w.score <- fread(file.path(basedir,"/results/all.eCMS.txt")) 
cohort.info <- full_join(cohort.info, predicted.w.score,  by=c('tumour'='sample')) 


```

## Loop through gene and k
We pull the data from the HPC for each NMF iteration. We also use gost() to try to identify what component we've isolated - this is more closely related to the more classical "biologically informed" approach to NMF-optimization for in silico microdissection, which looks for geneset enrichment to establish the best k. We didn't know what we were looking for (and didn't want to be limited by this expectation) but used it to ensure our tumour-content approach wasn't radicaly divergent from the previous approach. We'll show the gost() results in the next section (step 2.1.1). 

```{r}
base_font_size = 16
COR_CUT = 0.2
rank_range <- c(2:16) #20
subtype = '' 

heat.cluster.all <- NULL
pull.nmf=F
if(pull.nmf){

  first.outer.loop=T
  first.loop=T
  
  for(gene_n in c(2000, 4000, 6000, 8000, 10000, 12000, 14000)){
    
    gene.file.path <- file.path(cluster.dir, paste0("nmf.",gene_n,"g.rds"))
    
    if(file.exists(gene.file.path)){
      
      filtered_genes_cts <- readRDS(gene.file.path)
      
      # Results storage
      mscd_list <- numeric(length(rank_range))
      stab_list <- numeric(length(rank_range))
      names(mscd_list) <- names(stab_list) <- rank_range
      
      for(component_n in rank_range){
        
        this.file.path =   paste0(cluster.dir,"/nmf.k",component_n,".",gene_n,"g.",subtype,"rds")
        
        if(file.exists(this.file.path)){
          
          cat('\n\n-------\ngene_n: ',gene_n,' k: ',component_n)
          
          nmf_results <- readRDS(this.file.path)
          
          ## Reconstruct and compute cosine distance (MSCD)
          V_rec <- fitted(nmf_results@fit)
          V_rec_norm <- t(apply(V_rec, 1, function(x) x / sum(x)))
          
          og_norm <- t(apply(filtered_genes_cts, 1, function(x) x / sum(x)))
          sample_cosine_distances <- proxy::dist(og_norm, V_rec_norm, method = "cosine")
          mscd_list[component_n - 1] <- mean(as.vector(sample_cosine_distances))
          
          ## load matrices
          H_matrix <- read.csv(paste0(cluster.dir,"./h_mat.k",component_n,".",gene_n,"g.",subtype,"csv"), row.names = 1)
          W_matrix <- read.csv(paste0(cluster.dir,"./w_mat.k",component_n,".",gene_n,"g.",subtype,"csv"), row.names= 1)
          
          ## scale H by sum or scale 
          # H_matrix_ordered_norm <- sweep(H_matrix_ordered, 2, colSums(H_matrix_ordered), FUN = "/") * 100
          
          H_matrix_ordered_norm <-   data.frame(t(apply(H_matrix, 1, scale)))
          names(H_matrix_ordered_norm) <- names(H_matrix)
          
          
          H_matrix_ordered_norm.df <- data.table(H_matrix_ordered_norm)
          H_matrix_ordered_norm.df$sig.id <- paste0('sig', row.names(H_matrix_ordered_norm))
          
          H_matrix_ordered_norm.melt <- melt(H_matrix_ordered_norm.df) %>% 
                                          left_join(cohort.info, by=c('variable'='tumour'))
          
          ## loop across components/signnatures
          for(this.sig in unique(H_matrix_ordered_norm.melt$sig.id)){
            
            ## filter for weight higher than median (zero, after scale)
            this.sig.w <- H_matrix_ordered_norm.melt %>% 
              filter(sig.id == this.sig & !is.na(cellularity) & value > 0 ) %>%
              dplyr::select(value, cellularity)
            
            ## calculate correaltion between sig weight and cellularity
            this.cor <- cor.test(this.sig.w$value, this.sig.w$cellularity)
      
            ## annotate sig cell type
            
            this.sig.col <- paste0('V',gsub('sig','',this.sig))
            
            this.W_matrix <- W_matrix %>% dplyr::select(all_of(this.sig.col)) 
            names(this.W_matrix) <- 'w'
            
            ## pull 100 top genes in component for annotation
            these.genes <- row.names(this.W_matrix %>% arrange(desc(w)) %>% slice_head(n=100))
            
            sig.profile <- gost(query = these.genes, organism = "hsapiens", sources = 'HPA')
            
            ## if succesfully matched to tissue, annotate in table
            if(length(sig.profile$result) > 0){
              
              sig.profile.res <- sig.profile$result
              sig.profile.res$term_name <- gsub('-', ';', sig.profile.res$term_name)
              top.tissue = paste0(unlist(sig.profile.res %>%  slice_head(n = 3) %>% dplyr::select(term_name)),  collapse = "-")
              top.tissue.p = paste0(unlist(sig.profile.res %>%  slice_head(n = 3) %>% dplyr::select(p_value)),  collapse = ";")
              top.tissue.precision = paste0(unlist(sig.profile.res %>%  slice_head(n = 3) %>% dplyr::select(precision)),  collapse = ";")
              top.tissue.recall = paste0(unlist(sig.profile.res %>%  slice_head(n = 3) %>% dplyr::select(recall)),  collapse = ";")
              
              cat('\n',this.sig,': ' ,top.tissue,'\n')
              
            } else {
              top.tissue = 'none'
            }
            
            ## make row
            this.cor <- cbind.data.frame(
              'genes' = gene_n,
              'k'= component_n,
              'n'= nrow(this.sig.w),
              'sig.id'=this.sig,
              'cor'=this.cor$estimate,
              'cor.p'=this.cor$p.value,
              'top.tissue'=top.tissue,
              'top.tissue.p'=top.tissue.p,
              'top.tissue.precision'=top.tissue.precision,
              'top.tissue.recall'=top.tissue.recall
            )
            
            if(first.loop){these.cor <- this.cor; first.loop=F}else{these.cor <- rbind.data.frame(these.cor,this.cor)}
          
          }
          
          ## filter for components where cellularity and weight are more correlated than COR_CUT value
          cellularity.sigs <- H_matrix_ordered_norm.melt %>% 
            filter(sig.id %in% these.cor$sig.id[these.cor$cor > COR_CUT] ) 
          
          ## if more than one component correlated to cellularity
          if(length(unique(cellularity.sigs$sig.id)) > 1){
            
            cellularity.sigs <- cellularity.sigs %>% 
              dplyr::select(variable, sig.id, value) %>%
              unique() %>%
              reshape2::dcast(variable ~ sig.id)
            
            cellularity.sigs.mat <- as.matrix(cellularity.sigs[,-1])
            row.names(cellularity.sigs.mat) <- cellularity.sigs[,1]
            
            annotation <- data.frame('tumour_id'=rownames(cellularity.sigs.mat))
            
            annotation <- 
              annotation %>%
              #rownames_to_column(var='tumour_id') %>%
              inner_join(.,
                         cohort.info[,c("tumour", "cellularity",'rna_class','Site_of_tissue','variable')],
                         by = c("tumour_id"="tumour")) %>%
              tibble::column_to_rownames(var = "tumour_id")
            
            ph <- pheatmap(
              t(cellularity.sigs.mat),
              annotation_col = annotation,
              cluster_rows = TRUE,
              cluster_cols = TRUE,
              clustering_distance_cols = "euclidean",
              clustering_method = "complete", # vs. average
              color = colorRampPalette(c("darkblue","white", "darkgreen"))(100),
              
              show_colnames = TRUE,
              fontsize = 8,
              border_color = "white"
            )
            
            
            # The order of the x-axis labels (left -> right) on the heatmap:
            x_order <- ph$tree_col$labels[ph$tree_col$order]
            
            # If you want a data.frame mapping sample to its plotting order index:
            heat.cluster <- data.frame(
              sample = x_order,
              order  = seq_along(x_order),
              stringsAsFactors = FALSE
            )
            
            heat.cluster$component_n = component_n
            heat.cluster$gene_n = gene_n
            
            if(is.null(heat.cluster.all)){heat.cluster.all <- heat.cluster}else{heat.cluster.all <- rbind.data.frame(heat.cluster.all,heat.cluster)}
            
            hc_rows <- hclust(dist(cellularity.sigs.mat))
            
            # Calculate the cophenetic correlation coefficient
            correlation <- cor(dist(cellularity.sigs.mat), cophenetic(hc_rows))
      
            ## add to table
            stab_list[component_n - 1] <- correlation
            
            cat('\nstability ', round(correlation, 3),'\n')
            
          } else {
            cat('\nno components correlate with cellularity\n')
          }
        } 
        
      }
      #assemble other table
      df <- data.frame(
        genes = gene_n,
        k = rank_range,
        MSCD = mscd_list,
        Stability = stab_list
      )
      if(first.outer.loop){k.stats.df <- df; first.outer.loop=F}else{k.stats.df <- rbind.data.frame(k.stats.df, df)}
      
    }
  }
  
  nmf.stats <- left_join(these.cor, k.stats.df)
  
  write.table( nmf.stats,  
               file = file.path(basedir,'/results/nmf.stats.txt'), 
               row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
  
  write.table( heat.cluster.all,  
               file = file.path(basedir,'/results/heat.cluster.txt'), 
               row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
}

```

## explore NMF components
Here, we visualize the optization trade-offs.

```{r}

nmf.stats <- fread(file.path(basedir,'/results/nmf.stats.txt'))

## filter for correlated components
top_labels <-  nmf.stats %>% filter(cor > COR_CUT)

top.cor <- nmf.stats %>% filter(cor > COR_CUT) %>% group_by(k, genes) %>% 
  summarise(mean_cor = mean(cor), Stability = mean(Stability), n=n(), MSCD = mean(MSCD)) %>% 
  dplyr::select(k, genes, mean_cor, Stability, MSCD, n) %>%
  unique() %>% arrange(desc(mean_cor)) 

## scale y-axis so both scales can be plotted
mc_range <- range(top.cor$mean_cor)
stab_scaled <- scales::rescale(top.cor$Stability, to = mc_range)
MSCD_scaled <- scales::rescale(top.cor$MSCD, to = mc_range)

nmf.optimization.plot <- 
ggplot(top.cor %>% filter(genes < 12000), aes(x = k, y = mean_cor, color=as.factor(genes))) +

  geom_vline(xintercept = 8, color='grey80', size=4)+
  
  geom_line() +
  geom_point(aes(size=n)) +
  geom_text(aes(size=n*0.5, label=n), color='white') +
  scale_color_manual(values =  pnw_palette(name="Sailboat", n=length(unique(top.cor$genes)) )) +
  

  labs(x="Number of Components (k)", y="Mean Correlation with Tumour Content", color='n genes', size='n components\ncor. w/ TC') +
  theme_bw(base_size=10) + 
  scale_size(range = c(1,3))  +
  
  theme(           plot.margin = unit(c(0, 0, 0, 0), "mm"),
                   
                   panel.grid.minor = element_blank(),
                   panel.grid.major.y = element_blank(),
                   plot.background = element_blank(),
                   panel.background = element_blank(),
                   strip.background = element_blank(),
                   
                   legend.key.size = unit(0.4, "cm"),   # Shrinks the legend keys (symbols)
                   legend.text = element_text(size = 8), # Shrinks the text for legend items
                   legend.title = element_text(size = 9) # Shrinks the legend title
  )

nmf.optimization.plot

ggsave(paste0(basedir, "/results/fig/FIGS3a.nmf.optimization.svg"),plot =
nmf.optimization.plot
, device = "svg",, width = 100, height = 75, units = 'mm', dpi=72 )

```

## Cluster samples across NMF parameter space
We show that CCS is very robust to this parameterization: it doesn't matter how many genes or which k you choose, CCS is a top factor of differentiation across all parameter space. (The same will not be said about the component signatures though.) This is likely why CCS has been found multiple times before, even in cohorts that did not use LCM and optimized k by minimizing k. We also show that anatomical location of the primary tumor does not have the same consistency in clustering as CCS.

```{r}

####
heat.cluster.all <- fread(file.path(basedir,'/results/heat.cluster.txt'))

heat.cluster.all <- heat.cluster.all %>% filter(gene_n != 2000)

heat.cluster.sub <- left_join(heat.cluster.all, cohort.info, by=c('sample'='tumour')) %>% filter(!is.na(order))

max_ord <- max(heat.cluster.sub$order, na.rm = TRUE)

heat.cluster.flipped <- heat.cluster.sub %>%
  group_by(component_n, gene_n) %>%
  mutate(
    meanA = mean(order[rna_class == "CMS-A"], na.rm = TRUE),
    meanB = mean(order[rna_class == "CMS-B"], na.rm = TRUE),
    flip  = meanA < meanB,
    order.mod = if_else(flip, max_ord - order + 1L, order)
  ) %>%
  ungroup() 

nmf.all_clusters.plot <- 
ggplot(heat.cluster.flipped, aes(x=order.mod, y=component_n, color=rna_class)) +
  facet_wrap('gene_n',nrow=3 )+
  geom_point(size=0.45, alpha=1, stroke=0) +
  theme_bw(base_size=10) + 
  guides(color='none')+
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    panel.spacing = unit(0, "mm"),
      
    panel.background = element_blank(),
    panel.border = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
  ) + scale_y_reverse() +
  scale_color_manual(  values=c(
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" ,
    'HCC-CCA' = "#b4b4b4ff",
    'Perihilar_CCA_pCCA' = "#e59d54ff",
    'Intrahepatic_CCA_iCCA' = "#fbcc7aff",
    'Distal_CCA_dCCA' = "#ba7a44ff",
    'Gallbladder' = "#855b40ff"
  ), na.value = 'grey90') + labs(x='cluster order',y='component number (k)')

nmf.all_clusters.plot
  
ggsave(paste0(basedir, "/results/fig/FIGS3b.nmf.all_clusters.svg"),plot =
         nmf.all_clusters.plot
, device = "svg",, width = 90, height = 60, units = 'mm', dpi=72 )

ggplot(heat.cluster.flipped, aes(x=order, y=component_n, color=variable)) +
  facet_wrap('gene_n',nrow=3 )+
  geom_line(aes(group=BTCID),alpha=0.2)+
  geom_point(size=0.5) +
  theme_bw(base_size=10) + 
  
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    panel.background = element_blank(),
    panel.border = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
  ) + scale_y_reverse() +
  scale_color_manual(  values=c(
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" ,
    'HCC-CCA' = "#b4b4b4ff",
    'Perihilar_CCA_pCCA' = "#e59d54ff",
    'Intrahepatic_CCA_iCCA' = "#fbcc7aff",
    'Distal_CCA_dCCA' = "#ba7a44ff",
    'Gallbladder' = "#855b40ff"
  ), na.value = 'grey95') 

```

## precise investigation
Now that we've picked our top parameters (4,00 genes across 8 components), we do a little more fine-grained evaluation of the components. Biological interpretation is left to the next step. This is also where we copy the best values of H and W from the hpc to the results directory.

```{r}
gene_n = 4000
component_n = 8

nmf.stats %>% filter(k == component_n & genes == gene_n & cor > 0.2) %>% dplyr::select(sig.id  ,     cor     ,   cor.p)

sigs = c(2,6)

over.write.matrices = F
if(over.write.matrices){
  H_matrix <- read.csv(paste0(cluster.dir,"./h_mat.k",component_n,".",gene_n,"g.csv"), row.names = 1)
  
  H_matrix.print <-   data.frame(apply(H_matrix, 1, scale))
  H_matrix.print <- cbind.data.frame('tumour.id'=names(H_matrix), H_matrix.print)
  names(H_matrix.print) <- gsub('X','sig', names(H_matrix.print))
  
  W_matrix <- read.csv(paste0(cluster.dir,"./w_mat.k",component_n,".",gene_n,"g.csv"), row.names= 1)
  
  W_matrix.print <- W_matrix
  W_matrix.print$gene_name <- row.names(W_matrix)
  W_matrix.print <- left_join(W_matrix.print, gene_set, by=c('gene_name'='gene_name'))


  write.table( H_matrix,  
               file = file.path(basedir,'/results/H_matrix.txt'), 
               row.names = T,col.names = T, quote = FALSE, sep = ",", append =F)
  write.table( H_matrix.print,  
               file = file.path(basedir,'/results/H_matrix.print.txt'), 
               row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
  
  write.table( W_matrix,  
               file = file.path(basedir,'/results/W_matrix.txt'), 
               row.names = T,col.names = T, quote = FALSE, sep = ",", append =F)
  write.table( W_matrix.print,  
               file = file.path(basedir,'/results/W_matrix.print.txt'), 
               row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
} else {
  H_matrix <- fread(file.path(basedir,'/results/H_matrix.txt'))
}


H_matrix_ordered_norm <-   data.frame(t(apply(H_matrix[sigs,], 1, scale)))
names(H_matrix_ordered_norm) <- names(H_matrix)
  
annotation <- data.frame('tumour_id'=names(H_matrix_ordered_norm))

annotation <- 
  annotation %>%
  inner_join(.,
             cohort.info[,c("tumour", "cellularity",'rna_class','Site_of_tissue','variable')],
             by = c("tumour_id"="tumour")) %>%
  tibble::column_to_rownames(var = "tumour_id")


ph <- pheatmap(
  H_matrix_ordered_norm,
  annotation_col = annotation,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_distance_cols = "euclidean",
  clustering_method = "complete", # vs. average
  color = colorRampPalette(c("darkblue","white", "darkgreen"))(100),
  show_colnames = TRUE,
  fontsize = 8,
  border_color = "white"
)

col_dend <- ph$tree_col

dendogram_cut = 6
clusters <- cutree(col_dend, k = dendogram_cut)

heat.cluster <- data.frame(clusters)
heat.cluster$V1 <- row.names(heat.cluster)
heat.cluster <- left_join(heat.cluster, predicted.w.score, by=c('V1'='sample'))
heat.cluster$rna_class[is.na(heat.cluster$rna_class)] <- 'CMS-O'

transition_df <- as.data.frame(table(heat.cluster$clusters, heat.cluster$rna_class))
colnames(transition_df) <- c("Cluster", "CMS", "Freq")

transition.max <- transition_df %>% group_by(Cluster) %>% filter(Freq == max(Freq) & CMS != 'CMS-O') 

transition_df$Cluster.f <- factor(transition_df$Cluster, 
                                  levels=transition.max$Cluster[order(transition.max$CMS)])

ggplot(transition_df,
       aes(axis1 = Cluster.f, axis2 = CMS, y = Freq)) +
  geom_alluvium(aes(fill = Cluster.f), width = 2/12) +
  geom_stratum(width = 2/12, fill = "white", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Cluster", "CCS"), expand = c(.05, .05)) +
  scale_fill_manual(values =  pnw_palette(name="Sailboat", n=dendogram_cut )) +
  theme_minimal() + theme(
    strip.background = element_blank(), panel.grid = element_blank()
    
  ) + guides(fill='none')

  

```
