# Assemble and review Non-negative Matrix Factorization (NMF) 
This is another section where we do something a little unique. Non-negative Matrix Factorization (NMF) is great for decomposing factors but you a little bit need to know what you are looking for (there's lot of optimization). First, we need to know how many components to break the data into (k). For some applications, keeping k small is great, but for in silico microdissection, too small of a k will confound different cell types. But if we remove the constraint of minimizing k, then how do we pick k? What we decided to do is, since we want to isolate signal from the tumor from other cell types, we took our only orthogonal estimate of tumor content: the estimate from the WGS. While not expected to be exactly the same, they should be pretty close. So here we optimize k to maximize for components that correlate with WGS-estimate of tumor content. We do the same to choose the numbers of genes to include in the decomposition. 

## Load Data


```{r}

suppressWarnings(
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(stringr)
  library(NMF)
  library(clusterProfiler)
  library(msigdbr)
  library(gprofiler2)
  library(enrichplot)
  library(pheatmap)
  library(ggdendroplot)
  library(cowplot)
  library(PNWColors)
  library(ggalluvial)
})
)

packageVersion('gprofiler2')

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'
source( file.path(code.dir, "bin/nmf.functions.R"))
source( file.path(data.dir, "paths.R"))
cluster.dir = file.path(cluster.dir, 'nmf_analysis')

base_font_size = 16
COR_CUT = 0.2

gene_set <- fread(file.path(data.dir,'/source_data/fgs.txt'))

cohort.info <- fread(file.path(data.dir,"/results/samples.info.txt")) %>%
  dplyr::select(Sample, cellularity)
names(cohort.info) <- c("tumour","cellularity")

```

## Loop through gene and k
We pull the data from the HPC for each NMF iteration. We also use gost() to try to identify what component we've isolated - this is more closely related to the more classical "biologically informed" approach to NMF-optimization for in silico microdissection, which looks for geneset enrichment to establish the best k. We didn't know what we were looking for (and didn't want to be limited by this expectation) but used it to ensure our tumour-content approach wasn't radicaly divergent from the previous approach. We'll show the gost() results in the next section (step 2.1.1). 

```{r}

pull.nmf = F
overwrite.nmf = T
run.clustering = T
rank_range <- c(5:16)

if(pull.nmf){

  for(analysis.cohort in c('whole','naive','icca','primary')){
    analysis.cluster.dir = file.path('/Volumes/pcsi/users/fbeaudry/nmf_analysis/',analysis.cohort)

    heat.cluster.all = NULL
    these.cor = NULL
    k.stats.df = NULL
    
    for(gene_n in c(2000, 4000, 6000, 8000, 10000)){
      
      gene.file.path <- file.path(analysis.cluster.dir, paste0("nmf.",gene_n,"g.rds"))
      
      if(file.exists(gene.file.path)){
        
        cat('\nanalyzing ',analysis.cohort," cohort using ",
             gene_n,' genes across k = ')

        filtered_genes_cts <- readRDS(gene.file.path)
        
        # Results storage
        mscd_list <- numeric(length(rank_range))
        stab_list <- numeric(length(rank_range))
        names(mscd_list) <- names(stab_list) <- rank_range
        
        for(component_n in rank_range){
          
          this.file.path =   paste0(analysis.cluster.dir,"/nmf.k",component_n,".",gene_n,"g.rds")
          
          if(file.exists(this.file.path)){
            cat(component_n, ', ')
            
            nmf_results <- readRDS(this.file.path)
            
            mscd_list[component_n - min(rank_range) + 1] <- get_mscd(filtered_genes_cts, nmf_results)
            
            ## load matrices
            W_matrix <- read.csv(file.path(analysis.cluster.dir, paste0("w_mat.k",component_n,".",gene_n,"g.csv")), row.names= 1)
            H_matrix <- read.csv(file.path(analysis.cluster.dir, paste0("h_mat.k",component_n,".",gene_n,"g.csv")), row.names = 1)
            
            H_matrix_ordered_norm.melt <- reformat_h_matrix(H_matrix, cohort.info)
            
            ## loop across components/signatures
            for(this.sig in unique(H_matrix_ordered_norm.melt$sig.id)){
              
                this.cor <- cbind.data.frame(
                  'genes' = gene_n, 'k'= component_n,
                  get_signature_stats(this.sig, 
                                      W_matrix, 
                                      H_matrix_ordered_norm.melt)
                )
              
              if(is.null(these.cor)){these.cor <- this.cor}else{these.cor <- rbind.data.frame(these.cor,this.cor)}
            
            }
            
            ## filter for components where cellularity and weight are more correlated than COR_CUT value
            cellularity.sigs <- H_matrix_ordered_norm.melt %>% 
              filter(sig.id %in% these.cor$sig.id[these.cor$cor > COR_CUT & 
                                                   these.cor$genes == gene_n & 
                                                    these.cor$k == component_n] ) 
            
            ## if more than one component correlated to cellularity
            if(length(unique(cellularity.sigs$sig.id)) > 1){
            #   message('2+ components correlate with cellularity, evaluating stability\n')
  
              cellularity.sigs <- cellularity.sigs %>% 
                dplyr::select(variable, sig.id, value) %>%
                unique() %>%
                reshape2::dcast(variable ~ sig.id)
              
              cellularity.sigs.mat <- as.matrix(cellularity.sigs[,-1])
              
              if(run.clustering){
                row.names(cellularity.sigs.mat) <- cellularity.sigs[,1]
                
                heat.cluster <- get_clustering_order(cellularity.sigs.mat)
                heat.cluster$component_n = component_n
                heat.cluster$gene_n = gene_n
                
                if(is.null(heat.cluster.all)){heat.cluster.all <- heat.cluster
                }else{ heat.cluster.all <- rbind.data.frame(heat.cluster.all,heat.cluster)
                }
              }
              
              stab_list[component_n - min(rank_range) + 1] <- get_cophenetic_coef(cellularity.sigs.mat)
              
            } 
          } 

        }
  
        component.table <- data.frame(
          genes = gene_n,
          k = rank_range,
          MSCD = mscd_list,
          Stability = stab_list
        )
        
        if(is.null(k.stats.df)){k.stats.df <- component.table}else{k.stats.df <- rbind.data.frame(k.stats.df, component.table)}
        
      } 
    }
    
    nmf.stats <- left_join(these.cor, k.stats.df, by=c('genes'='genes','k'='k'))
    
    ## filter for correlated components
    top_labels <-  nmf.stats %>% filter(cor > COR_CUT)
    
    top.cor <- nmf.stats %>% filter(cor > COR_CUT) %>% group_by(k, genes) %>% 
      summarise(mean_cor = mean(cor), Stability = mean(Stability), n=n(), MSCD = mean(MSCD), .groups = "drop_last") %>% 
      dplyr::select(k, genes, mean_cor, Stability, MSCD, n) %>%
      unique() %>% arrange(desc(mean_cor)) 
    
    ## scale y-axis so both scales can be plotted
    mc_range <- range(top.cor$mean_cor)
    stab_scaled <- scales::rescale(top.cor$Stability, to = mc_range)
    MSCD_scaled <- scales::rescale(top.cor$MSCD, to = mc_range)
    

    top.cor$k_scaled <-  (top.cor$k - min(top.cor$k)) /
                   (max(top.cor$k) - min(top.cor$k))

    top.cor$score = top.cor$mean_cor + 0.05 * top.cor$k_scaled
    top.cor <- top.cor %>% arrange(desc(score))
    component_n = top.cor$k[1]
    gene_n = top.cor$genes[1]
    component_n
    gene_n

    nmf.stats$top <- 0
    nmf.stats$top[nmf.stats$k == component_n & nmf.stats$genes == gene_n ] <- 1
    
    if(overwrite.nmf){
      write.table( nmf.stats,  
                   file = file.path(data.dir,'results', paste0(analysis.cohort,'.nmf.stats.txt')), 
                   row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
      
       if(run.clustering){
        write.table( heat.cluster.all,  
                     file = file.path(data.dir,'results', paste0(analysis.cohort,'.heat.cluster.txt')), 
                     row.names = F,col.names = T, quote = FALSE, sep = "\t", append =F)
       }
    }
      
    nmf.optimization.plot <- 
      ggplot(top.cor , aes(x = k, y = mean_cor, color=as.factor(genes))) +
    
      geom_vline(xintercept = component_n, color='grey80', linewidth=4)+
      
      geom_line() +
      geom_point(aes(size=n)) +
    #  geom_text(aes(size=n*0.5, label=n), color='white') +
      scale_color_manual(values =  pnw_palette(name="Starfish", n=length(unique(top.cor$genes)) )) +
      
    
      labs(x="Number of Components (k)", 
           y="Mean Correlation with Tumour Content", 
           color='n genes', 
           size='n components\ncor. w/ TC',
           title=analysis.cohort) +
      theme_bw(base_size=10) + 
      scale_size(range = c(1,3))  +
      
      theme(          
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
                       
       panel.grid.minor = element_blank(),
       panel.grid.major.y = element_blank(),
       plot.background = element_blank(),
       panel.background = element_blank(),
       strip.background = element_blank(),
       
       legend.key.size = unit(0.4, "cm"),   
       legend.text = element_text(size = 8), 
       legend.title = element_text(size = 9) 
      )
    
    print(nmf.optimization.plot)
  
    ggsave(file.path(data.dir, "results/fig",paste0("nmf.optimization.",analysis.cohort,".svg")),plot =
      nmf.optimization.plot
  , device = "svg", width = 100, height = 75, units = 'mm', dpi=72 )
    
  }
}

```


## Cluster samples across NMF parameter space

We show that CCS is very robust to this parameterization: it doesn't matter how many genes or which k you choose, CCS is a top factor of differentiation across all parameter space. (The same will not be said about the component signatures though.) This is likely why CCS has been found multiple times before, even in cohorts that did not use LCM and optimized k by minimizing k. We also show that anatomical location of the primary tumor does not have the same consistency in clustering as CCS.

```{r}

analysis.cohort = 'whole'

heat.cluster.all <- fread(file.path(data.dir,'results', paste0(analysis.cohort,'.heat.cluster.txt')))

predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt') )

heat.cluster.sub <- left_join(heat.cluster.all, predicted.w.score, by=c('sample'='Sample')) %>% 
  filter(!is.na(order))

max_ord <- max(heat.cluster.sub$order, na.rm = TRUE)

heat.cluster.flipped <- heat.cluster.sub %>%
  group_by(component_n, gene_n) %>%
  mutate(
    meanA = mean(order[rna_class == "CCS-A"], na.rm = TRUE),
    meanB = mean(order[rna_class == "CCS-B"], na.rm = TRUE),
    flip  = meanA < meanB,
    order.mod = if_else(flip, max_ord - order + 1L, order)
  ) %>%
  ungroup() 

nmf.all_clusters.plot <- 
  ggplot(heat.cluster.flipped, aes(x=order.mod, y=component_n, color=rna_class)) +
    facet_wrap('gene_n',nrow=3 )+
    geom_point(size=0.45, alpha=1, stroke=0) +
    theme_bw(base_size=10) + 
    guides(color='none')+
    theme(
      plot.margin = unit(c(0, 0, 0, 0), "mm"),
      panel.spacing = unit(0, "mm"),
        
      panel.background = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      strip.background = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
    ) + scale_y_reverse() +
    scale_color_manual(  values=c(
      "CCS-B"    = "#EDA71A" , 
      "CCS-A" = "#04597C" ,
      'HCC-CCA' = "#b4b4b4ff",
      'Perihilar_CCA_pCCA' = "#e59d54ff",
      'Intrahepatic_CCA_iCCA' = "#fbcc7aff",
      'Distal_CCA_dCCA' = "#ba7a44ff",
      'Gallbladder' = "#855b40ff"
    ), na.value = 'grey90') + labs(x='cluster order',y='component number (k)')

nmf.all_clusters.plot
  
ggsave(paste0(data.dir, "/results/fig/FIGS3b.nmf.all_clusters.svg"),plot =
         nmf.all_clusters.plot
, device = "svg", width = 50, height = 65, units = 'mm', dpi=72 )

```


