# Analyze scRNA from Zhang et al 2020
scRNA re-analysis of [Zhang et al. 2020](https://doi.org/10.1016/j.jhep.2020.05.039). Full full reasoning of steps, see walkthrough for reanalysis of data from SONG et al. The same logic is applied here.

## Data import
Data can be downloaded from [geo](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138709).

```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(data.table)
  library(tibble)
  library(openxlsx)
  library(Seurat) 
  library(SeuratObject)
  library(AUCell)
  library(ggplot2)
  library(ggrepel)
  library(PNWColors)
  library(cowplot)
  library(gghalves)
  library(stringr)
  library(tidyr)
})

'%ni%' <- function(x,y)!('%in%'(x,y))

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

source(paste0(basedir,"/bin/BTC.functions.R"))
source(paste0(basedir,"/bin/scrna_fxn.R"))

this.cohort = 'zhang_2020'
main_dir <- file.path(secdir, '/source_data/scrna/',this.cohort)
setwd(main_dir)

create.seurat=F
if(create.seurat){
    
  # sample info metadata 
  sample_info <- tribble(
    ~file , ~sample_id, ~ condition,
    "GSM4116579_ICC_18_Adjacent_UMI.csv", "ICC_18_Adjacent", "adjacent",
    "GSM4116582_ICC_23_Adjacent_UMI.csv", "ICC_23_Adjacent", "adjacent",
    "GSM4116586_ICC_25_Adjacent_UMI.csv", "ICC_25_Adjacent", "adjacent",
    "GSM4116580_ICC_18_Tumor_UMI.csv",    "ICC_18_Tumor",    "tumor",
    "GSM4116581_ICC_20_Tumor_UMI.csv",    "ICC_20_Tumor",    "tumor",
    "GSM4116583_ICC_23_Tumor_UMI.csv",    "ICC_23_Tumor",    "tumor",
    "GSM4116584_ICC_24_Tumor1_UMI.csv",   "ICC_24_Tumor1",   "tumor",
    "GSM4116585_ICC_24_Tumor2_UMI.csv",   "ICC_24_Tumor2",   "tumor"
  
  )
  
  
  seurat_list <- list()
  
  for (i in 1:nrow(sample_info)) {
    file <- sample_info$file[i]
    sample_id <- sample_info$sample_id[i]
    condition <- sample_info$condition[i]
  
    df <- read.csv(file.path(main_dir, file), row.names = 1)
  
    if (ncol(df) > nrow(df)) df <- t(df)
  
    obj <- CreateSeuratObject(counts = df, project = sample_id, min.cells = 3, min.features = 200)
    obj$sample_id <- sample_id
    obj$condition <- condition
    obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  
    # Auto thresholds based on IQR
    q <- obj@meta.data
    nFeature_bounds <- quantile(q$nFeature_RNA, probs = c(0.05, 0.95))
    nCount_bounds   <- quantile(q$nCount_RNA, probs = c(0.05, 0.95))
    mt_upper        <- quantile(q$percent.mt, probs = 0.95)
  
    VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), pt.size = 0.1, ncol = 3) 
    FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") 
    
    nFeature_min <- nFeature_bounds[1]
    nFeature_max <- nFeature_bounds[2]
    nCount_max   <- nCount_bounds[2]
    mt_max       <- mt_upper
    
  
    # Filter
    obj <- subset(obj, subset = 
                    nFeature_RNA > nFeature_min &
                    nFeature_RNA < nFeature_max &
                    nCount_RNA   < nCount_max &
                    percent.mt   < mt_max)
    
    # SCTransform
    obj <- SCTransform(obj, verbose = FALSE, method = "glmGamPoi")
  
    seurat_list[[sample_id]] <- obj
  
  
  }
  
  
  features <- SelectIntegrationFeatures(seurat_list, nfeatures = 2000) # adjust number of features as needed
  
  seurat_list <- purrr::map(seurat_list, ~ RunPCA(.x, verbose = FALSE))
  
  seurat_list <- PrepSCTIntegration(seurat_list, anchor.features = features)
  
  anchors <- FindIntegrationAnchors(
    object.list = seurat_list,
    normalization.method = "SCT",
    anchor.features = features,
    reduction = "rpca"
  )
  
  combined <- IntegrateData(anchorset = anchors, normalization.method = "SCT") 
  
  DefaultAssay(combined) <- "integrated"
  
  saveRDS(combined,file.path(main_dir,'/zhang.rds'))
} else {
  combined <- readRDS(file.path(main_dir,'/zhang.rds'))
  
}


# perform steps in order on the integrate data 
combined <- RunPCA(combined)

ElbowPlot(combined) # determining the number of dimensions to use to find neighbours 

combined <- FindNeighbors(combined, dims = 1:15)
combined <- RunUMAP(combined, dims = 1:15)
DimPlot(combined, reduction = "umap", group.by = "sample_id")
DimPlot(combined, reduction = "umap", group.by = "condition") 

DefaultAssay(combined) <- "integrated"

combined <- FindClusters(combined, resolution = 0.3)
DimPlot(combined, reduction = "umap", group.by = "seurat_clusters")

find.cell.ids.denovo=F
if(find.cell.ids.denovo){
  
  cell_marker_list <- get_cell_marker_list()
  TOP_N <- 50
  
  
  combined <- PrepSCTFindMarkers(combined)
  markers <- FindAllMarkers(combined, only.pos = TRUE,
                            min.pct = 0.5, logfc.threshold = 0.5)
  
  top_markers <- markers %>%
    group_by(cluster) %>%
    top_n(TOP_N, wt = avg_log2FC)
  
  marker_gene_list <- split(top_markers$gene, top_markers$cluster)
  
  cluster_annotations <- lapply(marker_gene_list, function(cluster_genes) {
    get_overlap_scores(cluster_genes, cell_marker_list)
  })
  
  lapply(cluster_annotations, function(x) x[c(1:3),c(1:2)])
}

cluster_celltypes <- c(
  
  '0'='0',
  '1'='1',
  '2'='CD4+ T cell',
  '3'='CD8+ T cell',
  '4'='4',
  '5'='Macrophage',
  '6'='Resting T cell',
  '7'='Monocyte',
  '8'='Endothelial cell',
  '9'='NK cell',
  '10'='Monocyte',
  '11'='T reg',
  '12'='Plasma cell',
  '13'='13',
  '14'='Hepatic cell',
  '15'='Fibroblast',
  '16'='Immature B cell',
  '17'='Immature B cell',
  '18'='pDC'

  
)

cluster_celltypes.df <- data.frame(cluster_celltypes)
cluster_celltypes.df$seurat_clusters <- rownames(cluster_celltypes.df)

emb <- Embeddings(combined, "umap") %>%   as.data.frame() 

seurat_clusters <- combined@meta.data[, "seurat_clusters", drop = FALSE]

df <- bind_cols(emb, seurat_clusters)
df <- left_join(df, cluster_celltypes.df)

centroids <- df %>%
  group_by(cluster_celltypes) %>%
  summarise(
    x_centroid = median(umap_1),
    y_centroid = median(umap_2)
  )


# Plot with centroid labels
ggplot(df, aes(x = umap_1, y = umap_2)) +
  
  geom_point(size=3, shape=16, color='black' , data = ~ subset(., cluster_celltypes %in% c('Cholangiocyte'))) +  
  geom_point(size=2.5, shape=16, color='white', data = ~ subset(., cluster_celltypes %in% c('Cholangiocyte'))) +  
  
  geom_point(size=1.5, shape=16,aes(color = cluster_celltypes), stroke = 0) +
  
  geom_label_repel(data = centroids  %>% filter( cluster_celltypes %ni% c('Cholangiocyte')),
                   color='grey50', aes(x = x_centroid, y = y_centroid, label = cluster_celltypes), 
                   size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  
  geom_label_repel(data = centroids  %>% filter( cluster_celltypes %in% c('Cholangiocyte')),
                   aes(x = x_centroid, y = y_centroid, label = cluster_celltypes), 
                   size=3,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  
  theme_void() +  # removes all axes and background
  scale_color_manual(values =  pnw_palette(name="Shuksan2", n=23 ) ) + 
  scale_fill_manual(values =  pnw_palette(name="Shuksan2", n=23 ) ) + 
  theme(
    panel.border = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
    plot.background = element_blank()
  ) + labs(fill='',color='') #+ guides(fill='none',color='none')


```

## Project NMF signatures onto cells with AUCcell

```{r}


y_order.convert <- fread( file.path(basedir,'/results/NMF.conversion.txt' ))
y_order.convert$og <- gsub('sig','V', y_order.convert$og)

w.matrix <- read.csv(file.path(basedir,'/results/W_matrix.txt'), row.names= 1)
genes <- row.names(w.matrix)
w.matrix <- w.matrix %>% dplyr::select(all_of(y_order.convert$og))
names(w.matrix) <- y_order.convert$new
w.matrix <- as.matrix(w.matrix)
row.names(w.matrix) <- genes


gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))

tpm_data <- convert_w_to_tpm_like(w.matrix, gene_set)
gene_sets <- make_gene_sets(tpm_data, ntile=0.9)

setClass("ExpData", contains = "matrix")
expr_matrix <- GetAssayData(combined, assay = 'SCT', slot = 'data')

expr_matrix.nmf <- expr_matrix 

cells_rankings <- AUCell_buildRankings(expr_matrix.nmf) #nCores = 3, 

nmf_AUC <- AUCell_calcAUC(gene_sets, 
                          cells_rankings, 
)

```

## AUCcell thresholding
We found AUCcell results to be very impacted by thresholding so we left the paramters at default. Below is the fit of the NMF signatures across the scRNA cell population.

```{r, fig.height=8}


par(mfrow=c(4,2)) 
cells_assignment <- AUCell_exploreThresholds(nmf_AUC, plotHist=TRUE, nCores=1, assign=TRUE)
par(mfrow=c(1,1))

auc_matrix <- as.data.frame(t(getAUC(nmf_AUC)))
auc_matrix <- auc_matrix[Cells(combined), ]

auc_matrix_zeroed <- auc_matrix

for(this.sig.index in c(1:length(unique(names(cells_assignment)))) ){
  this.sig <- paste0('sig',this.sig.index)
  this.ass <- cells_assignment[[this.sig]]
  
  auc_matrix_zeroed[this.sig.index][auc_matrix_zeroed[this.sig.index] < this.ass$aucThr$selected] <- 0
  
}

```

## Visualize NMF signatures on umap

```{r}

combined <- AddMetaData(combined, metadata = auc_matrix_zeroed)

NMF_cols <- names(gene_sets)
combined$Top_NMF <- NMF_cols[max.col(combined@meta.data[, NMF_cols], ties.method = "first")]
scores <- combined@meta.data[, "Top_NMF", drop = FALSE]

df <- bind_cols(emb, seurat_clusters)
df <- left_join(df, cluster_celltypes.df)
df <- bind_cols(df, scores)
df <- cbind.data.frame(df, combined@meta.data[, NMF_cols])


df <- df %>%
  mutate(Top_NMF = if_else(
    rowSums(select(., starts_with("sig")) != 0) == 0,
    NA,
    Top_NMF
  ))

df <- bind_cols(df, combined@meta.data[, c('sample_id','condition')])
df$barcode <- rownames(df)

df$donor <- NA
df$donor[df$sample_id %in% c("ICC_18_Adjacent" ,"ICC_18_Tumor")] <- '18'
df$donor[df$sample_id %in% c( "ICC_20_Tumor")] <- '20'
df$donor[df$sample_id %in% c( "ICC_23_Adjacent","ICC_23_Tumor")] <- '23'
df$donor[df$sample_id %in% c( "ICC_24_Tumor1","ICC_24_Tumor2")] <- '24'
df$donor[df$sample_id %in% c( "ICC_25_Adjacent")] <- '25'

color.set <- c(
  'sig.1' = 'grey30',
  'sig.7' = 'black',
  
  
  'sig.2' = 'lightblue',
  'sig.6' = 'cornflowerblue',
  
  'sig.5' = "forestgreen" ,
  'sig.8' = '#0A6F99' ,               #"#04597C" ,
  
  'sig.3' = 'yellow',
  'sig.4' = '#D98E04'            #"#EDA71A"
  
  
)


ggplot(df , aes(x = umap_1, y = umap_2)) +
  
  geom_point(size=0.5, aes(color = Top_NMF)) +
  geom_label_repel(data = centroids, aes(x = x_centroid, y = y_centroid, label = cluster_celltypes), 
                   size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  theme_void() +  # removes all axes and background
  scale_color_manual(values =  
                       color.set, na.value = 'grey'
                     
                     
  ) + 
  #  scale_fill_manual(values =  pnw_palette(name="Bay", n=12 ) ) + 
  theme(
    panel.border = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
    plot.background = element_blank()
  ) 

df_subset <- df %>% filter(!is.na(Top_NMF))
df_subset <- df_subset[sample(nrow(df_subset), floor(nrow(df_subset) / 4)), ]

centroids_subset <- centroids %>% filter(cluster_celltypes %in% c(  'CD4+ T cell',
                                                                    'CD8+ T cell',
                                                                    
                                                                    'Macrophage',
                                                                    'Resting T cell',
                                                                    'Monocyte',
                                                                    'Endothelial cell',
                                                                    'NK cell',
                                                                    'Monocyte',
                                                                    'T reg',
                                                                    'Plasma cell',
                                                                    
                                                                    'Hepatic cell',
                                                                    'Fibroblast',
                                                                    'Immature B cell',
                                                                    'Immature B cell',
                                                                    'pDC'
))

tumour_candidate_clusters <- c(0,1,4) 


    ggplot(df_subset, aes(x = umap_1, y = umap_2)) +
         
         geom_point(size=0.75, aes(color = Top_NMF)) +
         
         geom_point(size=4, shape=16, color='black' , data = ~ subset(., seurat_clusters %in% tumour_candidate_clusters)) +  
         geom_point(size=3, shape=16, color='grey90', data = ~ subset(., seurat_clusters %in% tumour_candidate_clusters)) +  
         
         
         geom_label_repel(data = centroids_subset, aes(x = x_centroid, y = y_centroid, label = cluster_celltypes), 
                          size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
         theme_void() +  # removes all axes and background
      scale_color_manual(values =  
                           color.set, na.value = 'grey'
                         
                         
      ) + 
         guides(color='none') + 
         theme(
           panel.border = element_blank(),
           plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
           plot.background = element_blank()
         ) 
       



#### subset ####

subset_obj <- subset(combined, subset = seurat_clusters %in% tumour_candidate_clusters & 
                       sample_id %in% c('ICC_18_Tumor','ICC_20_Tumor','ICC_23_Tumor','ICC_24_Tumor1','ICC_24_Tumor2') 
)
    
subset_obj <- SCTransform(subset_obj, verbose = FALSE)
subset_obj <- RunPCA(subset_obj)
ElbowPlot(subset_obj) # determining the number of dimensions to use to find neighbours 

#for(i in c(3:20)){
  i = 5
  subset_obj <- RunUMAP(subset_obj, dims = 1:i)  # Adjust dims as needed
  subset_obj <- FindNeighbors(subset_obj, dims = 1:i )
  
  print(DimPlot(subset_obj, reduction = "umap", group.by = "sample_id"))
#}

subset_obj <- FindClusters(subset_obj, resolution = 0.1)  # Tune resolution
DimPlot(subset_obj, reduction = "umap", group.by = "seurat_clusters")

emb.sub <- Embeddings(subset_obj, "umap") %>%   as.data.frame() 
emb.sub$barcode <- rownames(emb.sub)
df.sub <- left_join(emb.sub, df[,-c(1:2)], by=c('barcode'='barcode'))

chol.centroids <- df.sub %>%
  group_by(sample_id, condition) %>%
  summarise(
    x_centroid = median(umap_1),
    y_centroid = median(umap_2)
  )



     ggplot(df.sub , aes(x = umap_1, y = umap_2)) +
         
         geom_point(size=3.5, color = 'black') + #, shape=disease
         geom_point(size=2.5, color = 'white') + #, shape=disease
         
         
         geom_point(size=1.5, aes(color = Top_NMF)) + #, shape=disease
         geom_label_repel(data = chol.centroids, aes(x = x_centroid, y = y_centroid, label = sample_id, fill=condition), color='white', 
                          size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
         
         
         
         theme_void() +  # removes all axes and background
         scale_fill_manual(values =  
                             c(
                               
                               'tumor' = 'black',
                               'adjacent' = 'grey80'
                               
                             ), na.value = 'grey'
         ) + 
       scale_color_manual(values =  
                            color.set, na.value = 'grey'
                          
                          
       ) + 
           #guides(color='none')+
         #  scale_fill_manual(values =  pnw_palette(name="Bay", n=12 ) ) + 
         theme(
           panel.border = element_blank(),
           legend.position = 'bottom',
           plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
           plot.background = element_blank()
         ) 

```

## infercnv
Now, let's take our cells and infer the copy number profile from gene expression. Aneuploid cells are more likely to be malignant.

```{r}

     
run.infercnv=F
     
if(run.infercnv){
  library(infercnv)
  library(Matrix)
  library(hdf5r)
  '%ni%' <- function(x,y)!('%in%'(x,y))
  options(scipen = 100)
  
  all_genes <- fread(file.path(basedir,'/source_data/fgs.txt'))
  
  #cluster_celltypes
  immune_celltypes <- c( "CD4+ memory T cell"    ,    "CD8+ T cell" ,
     "T reg"     ,       "NK cell"    ,     "Macrophage"       ,    
     "Monocyte" ,"Endothelial cell"    ,    "Plasma cell"    , "Dendritic cell" ,
     "Fibroblast"     ,        "B cell"           ,  "B cell" ,    "Dendritic cell" )
  
  df$meta_celltype <- 'epithelial'
  df$meta_celltype[df$cluster_celltypes %in%  immune_celltypes] <- 'immune'
  
  combined <- AddMetaData(combined, metadata = df$meta_celltype, col.name = 'meta_celltype')
  
  ref_group_names <- c("immune")
  
  for(this.sample in unique(combined$sample_id)){
    
    if (!dir.exists(paste0("./infercnv/",this.sample))) {
      dir.create(paste0("./infercnv/",this.sample), recursive = T)
    }
    
    tumour.subset_obj <- subset(combined, subset = 
                        sample_id %in% this.sample) 
    
    annotations <- data.frame(
      Cell = colnames(tumour.subset_obj),
      Group = tumour.subset_obj$meta_celltype
    )
    
    write.table(annotations, paste0("./infercnv/",this.sample,"/cellAnnotations.txt"), 
                sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
    
    gene_names <-  data.frame('gene_name'=rownames(tumour.subset_obj@assays$RNA))
    
    gene_names <- left_join(gene_names, all_genes) %>% dplyr::select(gene_name, Chr,  Start,   End)
    duplicate.genes <- gene_names %>% group_by(gene_name) %>% tally() %>% arrange(desc(n)) %>% filter(n > 1)
    
    gene_names <- gene_names %>% filter(gene_name %ni% duplicate.genes$gene_name & !is.na(Chr) & Chr < 23) %>% arrange(Chr, Start,   End)
    
    write.table(gene_names, paste0("./infercnv/",this.sample,"/geneAnnotations.txt"), sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
    
    tumour.subset_obj <- subset(tumour.subset_obj, features = gene_names$gene_name)
    
    count_mat <- tumour.subset_obj@assays$RNA$counts
    
    infercnv.obj <- CreateInfercnvObject(raw_counts_matrix = count_mat,
                                         annotations_file = paste0("./infercnv/",this.sample,"/cellAnnotations.txt"),
                                         delim = "\t",
                                         ref_group_names = ref_group_names ,
                                         gene_order_file = paste0("./infercnv/",this.sample,"/geneAnnotations.txt")
                                         )
  
    infercnv.res <- infercnv::run(infercnv.obj,
                                  # noise_filter = 0.2,  
                                  # cutoff=0.5,
                                  out_dir=paste0("./infercnv/",this.sample,"/output/"),
                                  cluster_by_groups=TRUE,
                                  denoise=TRUE,  
                                  write_expr_matrix = T,
                                  no_plot=T,
                                 # num_threads=1,
                                  HMM=T
                                  )
  
    cnv_matrix <- data.frame(infercnv.res@expr.data)
    
    write.table( cnv_matrix,  
                 file = paste0("./infercnv/",this.sample,"/",this.sample,'.cnv_matrix.tsv'), 
                 row.names = T,col.names = T, quote = FALSE, sep = "\t", append =F)
    
    
    
    
  }
}


```

## Predict gCCS from inferCNV profile

```{r}


cytoBand <- fread(file.path(basedir,'/source_data/cytoBand.txt'), na.strings = '') %>% 
  filter(!is.na(Chromosome))

cytoBand<- separate(cytoBand, col = Chromosome, into = c('band','period'), sep = '\\.', remove=F)

cytoBand <- cytoBand %>%
  mutate(
    parts = str_match(band, "^(.*?[pq])(.*)$"),
    arm = parts[,2],
    armband = parts[,3]
  ) %>%
  dplyr::select(-parts) %>% filter(!is.na(arm) & arm %ni% c('Xq','Xp','Yp','Yq'))

full_cnv_model <- readRDS(file.path(basedir,'/results/cnv.model.rds'))

first.sample = T
for(this.sample in unique(subset_obj$sample_id) ){
  
  infercnv.path <- paste0(secdir,"/source_data/scrna/zhang_2020/infercnv/",this.sample,"/",this.sample,'.cnv_matrix.tsv')
  
  if(file.exists(infercnv.path)){
    
    cat('analyzing ', this.sample,'\n')
    
    infer.cnv.data <- fread(infercnv.path )
    
    
    #get gCMS per cell
    
    segment.gene.cyto <- inner_join(cytoBand, infer.cnv.data,  by=c('Hugo_Symbol'='V1')) %>%
      dplyr::select(-c("Hugo_Symbol" , "Chromosome"  , "band" , "period" ,   "armband" ))
    
    ggplot(segment.gene.cyto, aes(x=band, y=))
    

    arm.med <- segment.gene.cyto %>%
      group_by(arm ) %>% 
      summarise(across(where(is.numeric), median, na.rm = F)) 

    arm.med.t <- data.frame(t(arm.med[,-1]))
    arm.med.t <- log(arm.med.t, 2)
    names(arm.med.t) <- paste0('arm.',arm.med$arm)
    arm.med.t$mean.vaf <- 0.5
    
    #not a great solution for missing arms...
    #arm.med.t$arm.18q <- 0
    
    arm.med.t$prob <- predict(full_cnv_model, arm.med.t, type = "response")
    
    #get cnv-score per cell
    arm.med.t$cnv_scores <- sqrt(colMeans(infer.cnv.data[,-1]^2, na.rm = TRUE))
    
    arm.med.t$barcode <-  rownames(arm.med.t)
    
    if(first.sample){cnv.loads <- arm.med.t; first.sample=F} else {cnv.loads <- rbind.data.frame(cnv.loads,  arm.med.t) }
    
  } else {
    
    cat('\nno infercnv profile for ',this.sample,'\n')
    
  }
}


### mismatched lengths are due to not running all samples 
df.cnv <- left_join(df, cnv.loads, by=c('barcode'='barcode')) %>% filter(!is.na(mean.vaf))

#cluster_celltypes
immune_celltypes <- c( "CD4+ memory T cell"    ,    "CD8+ T cell" ,
                       "T reg"     ,       "NK cell"    ,     "Macrophage"       ,    
                       "Monocyte" ,"Endothelial cell"    ,    "Plasma cell"    , "Dendritic cell" ,
                       "Fibroblast"     ,        "B cell"           ,  "B cell" ,    "Dendritic cell" )

df.cnv$meta_celltype <- 'epithelial'
df.cnv$meta_celltype[df.cnv$cluster_celltypes %in%  immune_celltypes] <- 'immune'

THRESHOLD = 0.34

ggplot(df.cnv , aes(x=cnv_scores, fill=Top_NMF)) +
  geom_vline(xintercept = 1, color="black",  linetype='longdash')+
  
  geom_histogram(position = position_dodge())+
  facet_grid(donor~meta_celltype, scales='free') +
  theme_bw()+
  theme(panel.grid=element_blank(),
        strip.background = element_blank()) +
  
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  ) 

zhang.gCMS_prob.NMF.plot <- 
ggplot(df.cnv %>% filter(!is.na(donor)), aes(x=prob, fill=Top_NMF)) +
  geom_vline(xintercept = THRESHOLD, color="black",  linetype='longdash')+
  
  geom_vline(xintercept = THRESHOLD + 0.1, color="#04597C",  linetype='longdash')+
  geom_vline(xintercept = THRESHOLD - 0.1, color="#EDA71A",  linetype='longdash')+
  
  geom_vline(xintercept = 0.8, color="#04597C",  linetype='dashed')+
  geom_vline(xintercept = 0.2, color="#EDA71A",  linetype='dashed')+
  
  
  geom_histogram(position = position_dodge())+
  facet_grid(donor+sample_id~meta_celltype, scales='free') +
  theme_bw()+
  theme(panel.grid=element_blank(),
        strip.background = element_blank()) +
  
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  ) 
zhang.gCMS_prob.NMF.plot
ggsave(file.path(basedir, "/results/zhang.gCMS_prob.NMF.svg"), plot = 

zhang.gCMS_prob.NMF.plot
     , device = "svg", height = 100,width = 150, units = 'mm' )

df.cnv$gCMS <- NA
df.cnv$gCMS[df.cnv$prob > 0.8 ] <- 'CMS-A'
df.cnv$gCMS[df.cnv$prob < 0.2 ] <- 'CMS-B'


df.cnv.filt <- df.cnv[!is.na(df.cnv$gCMS) & df.cnv$meta_celltype ==  'epithelial' & df.cnv$sample_id != 'ICC_25_Adjacent',]


```

## eCCS

```{r}


tumour.subset_obj <- subset(subset_obj, cells = df.cnv.filt$barcode) 

whole.estimate <- get_cms_from_pseudobulk(tumour.subset_obj)
whole.estimate$sample_id <- rownames(whole.estimate)

meta <- tumour.subset_obj@meta.data
meta$cell <- rownames(meta)

for(this.boot in c(1:100)){
  
  seed=format(Sys.time(), "%OS3")
  set.seed(seed)
  
  
  
  sampled_cells <- meta %>%
    group_by(sample_id) %>%
    slice_sample(n = 20, replace = F) %>%
    pull(cell)
  
  
  boot.subset <- subset(tumour.subset_obj, cells = sampled_cells)
  
  suppressMessages(
    boot.estimate <- get_cms_from_pseudobulk(boot.subset)
  )
  
  boot.estimate$sample_id <- rownames(boot.estimate)
  boot.estimate$boot <- this.boot
  
  if(this.boot == 1){all.boot.estimate <- boot.estimate}else{all.boot.estimate <- rbind.data.frame(all.boot.estimate, boot.estimate)}
  
}

tsp_summary_stats <- all.boot.estimate %>%
  group_by(sample_id) %>%
  summarise(
    median = median(confidence.polarized ),
    mean = mean(confidence.polarized ),
    se = sd(confidence.polarized ) / sqrt(n()),
    ci_low = mean - qt(0.975, df = n() - 1) * se,
    ci_high = mean + qt(0.975, df = n() - 1) * se
  )

tsp.confidence <- left_join(whole.estimate, tsp_summary_stats)

tsp.confidence$Sample <- factor(tsp.confidence$sample_id, levels = tsp.confidence$sample_id[order(tsp.confidence$confidence.polarized )])


cohort <- df %>% dplyr::select(sample_id, donor, condition) %>% unique()


tsp.confidence <- left_join(tsp.confidence, cohort, by=c('sample_id'='sample_id'))



donor.confidence <- tsp.confidence %>% group_by(donor) %>% summarise(mean.conf = mean(confidence.polarized, na.rm=T))
tsp.confidence$Donor <- factor(tsp.confidence$donor, levels = donor.confidence$donor[order(donor.confidence$mean.conf)])


tsp.confidence$disease <- 'iCCA'



   ggplot(tsp.confidence , aes(x=sample_id,  color=rna_class)) + 
   
   geom_errorbar(aes(ymin = ci_low*100, ymax = ci_high*100), width = 0.2)  +
   
   #  geom_line(aes(group=Donor, y=confidence.polarized*100))+
   geom_point(aes(y=confidence.polarized*100), size=1.5)+
   geom_point(aes(y=mean*100), shape=1)+
   
   facet_grid(. ~  Donor+disease , space='free_x',scales='free_x') +
   
   theme_bw(base_size=10) +scale_color_manual(values =  
                                                c(
                                                  "CMS-A"="#04597C", "CMS-B"="#EDA71A"
                                                  
                                                  
                                                ), na.value = 'grey'
   ) +labs(y='% of TSP for CMS-A')+ theme(panel.grid = element_blank(), 
                                          axis.ticks.x = element_blank(),
                                          axis.title.x = element_blank(),
                                          axis.text.x = element_blank(),
                                          
                                          #strip.text.x = element_text(size = 6),
                                          
                                          plot.background=element_blank(),
                                          panel.background = element_blank(),
                                          panel.spacing = unit(0, "lines"),
                                          strip.background = element_blank()) + guides(color='none')+ ylim(0,100)
 

df.cms <- left_join(df.cnv.filt, whole.estimate, by=c('sample_id'='sample_id'))

df.cms.n <- df.cms %>% group_by(sample_id, donor, condition, rna_class, confidence.polarized, Top_NMF) %>% tally() 
df.cms.n$esig <- factor(df.cms.n$Top_NMF, levels = paste0('sig.',c(1:10)))

df.cms.n$disease <- 'iCCA'

ggplot(df.cms.n , aes(x=sample_id, y=n, fill=esig)) + 
  geom_bar(stat='identity')+
  facet_grid(.~donor+disease , space='free_x',scales='free_x' ) +
  theme_bw(base_size=10)+ 
  theme(panel.grid = element_blank(), 
        legend.position = 'bottom',
        legend.background = element_blank(),
        legend.text = element_text(size = 6), 
        legend.key.size = unit(0.5, "lines"),     
        plot.background=element_blank(),
        panel.background = element_blank(),
        legend.title = element_blank(),
        
        axis.title.x =  element_blank(),
        
        
        panel.spacing = unit(0, "lines"),
        strip.text = element_blank(),
        strip.background = element_blank())+ labs(y='n of aneuploid cells')+
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  )


```

## cell gCCS to tumour eCCS
Combine the cell-level genetic CCS calls to the pseudo-bulk eCCS calls and to the single cell NMF-signature assignments to show all three lines of evidence line up. Based on this analysis, we suspect ICC_25_Adjacent is an HCC tumor and we remove it here.

```{r}
 
 
 tumour.cnv.arms <- df.cnv.filt %>%  dplyr::select(c("donor","sample_id","barcode","gCMS","prob") |  starts_with("arm.") )
 
 tumour.cnv.arms <- tumour.cnv.arms %>% 
   reshape2::melt(id.vars=c("donor","sample_id",'barcode','prob','gCMS'))
 
 tumour.cnv.arms$band <- gsub("arm.","", tumour.cnv.arms$variable)
 
 tumour.cnv.arms <- tumour.cnv.arms %>%
   dplyr::mutate(
     arm = substr(band, nchar(band), nchar(band)),
     chr = substr(band, 1, nchar(band) - 1)
   ) %>%
   dplyr::select(donor,sample_id,barcode, prob, gCMS, chr, arm,band, value)
 
 model.coeff <- data.frame('band'=gsub('arm.','',names(full_cnv_model$coefficients))) %>%  mutate(
   arm = substr(band, nchar(band), nchar(band)),
   chr = substr(band, 1, nchar(band) - 1)
 ) %>%   select(chr, arm, band)
 
 tumour.cnv.model.arms <- tumour.cnv.arms %>% filter(band %in% model.coeff$band & !is.na(gCMS))
 tumour.cnv.model.arms$Band <- factor(tumour.cnv.model.arms$band, levels = paste0(rep(1:21, each = 2), c("p", "q")) )
 
 
 tumour.cnv.model.arms %>%
   group_by(sample_id) %>%
   summarise(
     most_common = names(sort(table(gCMS), decreasing = TRUE))[1],
     percentage = max(table(gCMS)) / n() * 100
   ) %>% summarise(mean_percentage=mean(percentage))
 
 
 
 tumour.cnv.model.arms$eCMS <- NA
 tumour.cnv.model.arms$eCMS[tumour.cnv.model.arms$donor %in% tsp.confidence$donor[tsp.confidence$rna_class == 'CMS-A']] <- 'CMS-A'
 tumour.cnv.model.arms$eCMS[tumour.cnv.model.arms$donor %in% tsp.confidence$donor[tsp.confidence$rna_class == 'CMS-B']] <- 'CMS-B'
 
 tumour.cnv.model.arms.match <- tumour.cnv.model.arms[!is.na(tumour.cnv.model.arms$eCMS) &  !is.na(tumour.cnv.model.arms$gCMS),] #[tumour.cnv.model.arms$eCMS == tumour.cnv.model.arms$gCMS,]
 tumour.cnv.model.arms.match <- tumour.cnv.model.arms %>% filter(sample_id != 'ICC_25_Adjacent')
 
 tumour.cnv.model.arms.match$Donor <- factor(tumour.cnv.model.arms.match$donor, levels = donor.confidence$donor[order(donor.confidence$mean.conf)])
 
 
 df.cms.n.filt <- df.cms %>% filter(barcode %in% tumour.cnv.model.arms.match$barcode) %>% 
   group_by(sample_id, donor, condition, rna_class, confidence.polarized, Top_NMF,gCMS ) %>% tally() 
 df.cms.n.filt$n[df.cms.n.filt$gCMS == 'CMS-A'] <- -df.cms.n.filt$n[df.cms.n.filt$gCMS == 'CMS-A']
 
 df.cms.n.filt$esig <- factor(df.cms.n.filt$Top_NMF, levels = paste0('sig.',c(1:10)))
 
 df.cms.n.filt$disease <- 'iCCA'
 df.cms.n.filt$Donor <- factor(df.cms.n.filt$donor, levels = donor.confidence$donor[order(donor.confidence$mean.conf)])
 
 df.cms.n.filt$gCMS.f <- factor(paste0('g',df.cms.n.filt$gCMS), levels = rev(c('gCMS-A','gCMS-B')))
 
 zhang.nmf.tally.plot <- 
 plot_grid(
   
 ggplot(tsp.confidence %>% filter(sample_id != 'ICC_25_Adjacent'), aes(x=sample_id,  color=rna_class)) + 
   
   geom_hline(yintercept = 75, color='grey', linetype='dotted')+
   geom_hline(yintercept = 25, color='grey', linetype='dotted')+
   
   geom_errorbar(aes(ymin = ci_low*100, ymax = ci_high*100), width = 0)  +
   
   geom_point(aes(y=confidence.polarized*100), size=1.5)+
 #  geom_point(aes(y=mean*100), shape=1)+
   
   facet_grid(. ~  Donor + disease, space='free_x',scales='free_x') +
   
   theme_bw(base_size=10) +scale_color_manual(values =  
                                                c(
                                                  "CMS-A"="#04597C", "CMS-B"="#EDA71A"
                                                  
                                                  
                                                ), na.value = 'grey'
   ) +labs(y='% of TSP')+ theme(panel.grid = element_blank(), 
                                          axis.ticks.x = element_blank(),
                                          axis.title.x = element_blank(),
                                          axis.text.x = element_blank(),
                                          
                                strip.text.x = element_text(size = 4,margin = margin(t = 2, b = 2)),
                                          
                                          plot.background=element_blank(),
                                          panel.background = element_blank(),
                                          
                                          panel.spacing = unit(0, "lines"),
                                          strip.background = element_blank()) + guides(color='none')+
   scale_y_continuous(limits = c(0,100), breaks = c(20,80))
   ,NULL,
 
 ggplot(df.cms.n.filt, aes(x=sample_id, y=n, fill=esig)) + 
   geom_bar(stat='identity')+
   facet_grid(gCMS.f~Donor , space='free',scales='free' ) +
   theme_bw(base_size=10)+ 
   theme(panel.grid = element_blank(), 
         legend.position = 'bottom',
         legend.background = element_blank(),
         legend.text = element_text(size = 6), 
         legend.key.size = unit(0.5, "lines"),     
         plot.background=element_blank(),
         panel.background = element_blank(),
         legend.title = element_blank(),
         
         axis.title.x =  element_blank(),
         axis.text.x =  element_blank(),
         axis.ticks.x =  element_blank(),
         
         
         panel.spacing = unit(0, "lines"),
         strip.text = element_blank(),
         strip.background = element_blank()
   )+ labs(y='n cells')+
   guides(color='none',fill='none')+
   scale_fill_manual(values =  
                       color.set, na.value = 'grey'
   )
 
 ,ncol = 1, align = 'hv', axis = "rlbt", 
 rel_heights =c(0.3,-0.1,0.7)) 
 
zhang.nmf.tally.plot
 
ggsave(file.path(basedir, "/results/fig/FIGS3e.zhang.nmf.tally.svg"), plot = 
 zhang.nmf.tally.plot
, device = "svg", height = 60,width = 40, units = 'mm', dpi=72 )

 chol.centroids.donor <- df.sub %>%
   group_by(donor) %>%
   summarise(
     x_centroid = median(umap_1),
     y_centroid = median(umap_2)
   )
 

ggplot(df.sub %>% filter(sample_id != 'ICC_25_Adjacent' & barcode %in% tumour.cnv.model.arms.match$barcode), aes(x = umap_1, y = umap_2)) +
          
          geom_point(size=0.5, aes(color = Top_NMF)) +
          geom_label_repel(data = chol.centroids.donor, aes(x = x_centroid, y = y_centroid, label = donor), 
                           size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
          theme_void() +  # removes all axes and background
          scale_color_manual(values =  
                               color.set, na.value = 'white'
                             
                             
          ) + 
          guides(fill='none',color='none')+
          theme(
            panel.border = element_blank(),
            plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
            plot.background = element_blank()
          ) 
        

 
```
