# Analyze scRNA from Song G et al 2022
scRNA re-analysis of [Song G. et al. 2022](https://doi.org/10.1038/s41467-022-29164-0)

## Data import
These data can be accessed by connecting with the authors of that manuscript. We thank Dr. Guohe Song for giving us access.

```{r}
suppressPackageStartupMessages({

library(dplyr)
library(data.table)
library(tibble)
library(openxlsx)
library(tidyr)
library(stringr)
library(Seurat) 
library(SeuratObject)
library(AUCell)
library(ggplot2)
library(ggrepel)
library(PNWColors)
library(cowplot)
library(gghalves)
  
})

packageVersion('Seurat')
packageVersion('AUCell')

'%ni%' <- function(x,y)!('%in%'(x,y))

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

source(paste0(basedir,"/bin/BTC.functions.R"))
source(paste0(basedir,"/bin/scrna_fxn.R"))

this.cohort = 'song_2022'
main_dir <- file.path(secdir, '/source_data/scrna/',this.cohort)
setwd(main_dir)

save.ind.nonepi = F
if(save.ind.nonepi){
  noepi <- readRDS(file.path(secdir,'source_data/scrna/song_2022/nonepi.rds'))
  no.epi.tumours <- unique(noepi@meta.data[,'orig.ident'])[grepl("T$", unique(noepi@meta.data[,'orig.ident']))]
  
  for(this.sample in no.epi.tumours){
  
    noepi.sub <- subset(noepi, subset = orig.ident == this.sample)
  
    noepi.sub <- SCTransform(noepi.sub, verbose = FALSE, method = "glmGamPoi")
    noepi.sub$cellmeta <- 'noepi'
    
    if (!dir.exists(paste0("./infercnv/",this.sample))) {
      dir.create(paste0("./infercnv/",this.sample), recursive = T)
    }
    
    saveRDS(noepi.sub, paste0("./infercnv/",this.sample,'/nonepi.',this.sample,'.rds'))
  
  }
}

song.epi <- readRDS(file.path(secdir,'/source_data/scrna/song_2022/epi.rds'))

song.epi <- SCTransform(song.epi, verbose = FALSE)
song.epi <- subset(song.epi, subset = orig.ident %in% c('I02T', 'I03T', 'I04T', 'I06T', 'I09T', 'I10T', 
                                                                 'I12T', 'I13T', 'I14T', 'I15T',  'I16T', 'I17T', 'I18T')) 


```

## Explore dataset at a high level

```{r}

song.epi <- RunPCA(song.epi)

ElbowPlot(song.epi) # determining the number of dimensions to use to find neighbours 

song.epi <- FindNeighbors(song.epi, dims = 1:15)
song.epi <- RunUMAP(song.epi, dims = 1:15)

DimPlot(song.epi, reduction = "umap", group.by = "orig.ident")

```

## Project NMF signatures onto cells with AUCcell
We are mostly interested in the scRNA data because NMF decomposition does not necessarily resolve single cells, so we combine the two approaches. There's quite a bit of manual parametrization here again. The first is that we are transiting from gene weights from the NMF to gene lists for AUCcell. So we take the distribution of weights from the NMF and fit a bimodal distribution with mcclust using our make_gene_sets() function, most genes have small weights so are discarded in this step. However, the genesets were still highly correlated with just the removal of low weight genes (see corplot) so we go one step further and subset to the nth-tile of the distribution (in this case 90th percentile). 

```{r}

y_order.convert <- fread( file.path(basedir,'/results/NMF.conversion.txt' ))
y_order.convert$og <- gsub('sig','V', y_order.convert$og)

w.matrix <- read.csv(file.path(basedir,'/results/W_matrix.txt'), row.names= 1)
genes <- row.names(w.matrix)
w.matrix <- w.matrix %>% dplyr::select(all_of(y_order.convert$og))
names(w.matrix) <- y_order.convert$new
w.matrix <- as.matrix(w.matrix)
row.names(w.matrix) <- genes

gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))
tpm_data <- convert_w_to_tpm_like(w.matrix, gene_set)
gene_sets <- make_gene_sets(tpm_data, ntile=0.9)

setClass("ExpData", contains = "matrix")
expr_matrix <- GetAssayData(song.epi, assay = 'SCT', slot = 'data')

expr_matrix.nmf <- expr_matrix[rownames(expr_matrix) %in% unique(unlist(gene_sets)), ]

cells_rankings <- AUCell_buildRankings(expr_matrix.nmf) 

nmf_AUC <- AUCell_calcAUC(gene_sets, 
                          cells_rankings, 
)

```

## AUCcell thresholding
We found AUCcell results to be very impacted by thresholding so we left the paramters at default. Below is the fit of the NMF signatures across the scRNA cell population.

```{r, fig.height=8}

## find thresholds
par(mfrow=c(4,2)) 
cells_assignment <- AUCell_exploreThresholds(nmf_AUC, plotHist=TRUE, nCores=1, assign=TRUE)
par(mfrow=c(1,1))

auc_matrix <- as.data.frame(t(getAUC(nmf_AUC)))
auc_matrix <- auc_matrix[Cells(song.epi), ]

auc_matrix_zeroed <- auc_matrix

for(this.sig.index in c(1:length(unique(names(cells_assignment)))) ){
  this.sig <- paste0('sig.',this.sig.index)
  this.ass <- cells_assignment[[this.sig]]
  
  
  auc_matrix_zeroed[this.sig.index][auc_matrix_zeroed[this.sig.index] < this.ass$aucThr$selected] <- 0
  
}

```

## Visualize NMF signatures on umap
With our signatures assigned, we project them in UMAP space.

```{r}

song.epi <- AddMetaData(song.epi, metadata = auc_matrix_zeroed)

NMF_cols <- names(gene_sets)
song.epi$Top_NMF <- NMF_cols[max.col(song.epi@meta.data[, NMF_cols], ties.method = "first")]
scores <- song.epi@meta.data[, "Top_NMF", drop = FALSE]

emb <- Embeddings(song.epi, "umap") %>%   as.data.frame() 

df <- bind_cols(emb, scores)
df <- cbind.data.frame(df, song.epi@meta.data[, NMF_cols])
df$donor <- song.epi@meta.data[, 'orig.ident'] 

centroids <- df %>%
  group_by(donor) %>%
  summarise(
    x_centroid = median(umap_1),
    y_centroid = median(umap_2)
  )

color.set <- c(
  'sig.1' = 'grey30',
  'sig.7' = 'black',
  
  
  'sig.2' = 'lightblue',
  'sig.6' = 'cornflowerblue',
  
  'sig.5' = "forestgreen" ,
  'sig.8' = '#0A6F99' ,               #"#04597C" ,
  
  'sig.3' = 'yellow',
  'sig.4' = '#D98E04'            #"#EDA71A"
  
  
)

ggplot(df, aes(x = umap_1, y = umap_2)) +
  
  geom_point(size=0.5, aes(color = Top_NMF)) +
  geom_label_repel(data = centroids, aes(x = x_centroid, y = y_centroid, label = donor), 
                   size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  theme_void() +  # removes all axes and background
  scale_color_manual(values =  
                       color.set, na.value = 'white'
                     
                     
  ) + 
  #  scale_fill_manual(values =  pnw_palette(name="Bay", n=12 ) ) + 
  theme(
    panel.border = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
    plot.background = element_blank()
  ) 

```

## infercnv
Now, let's take our cells and infer the copy number profile from gene expression. Aneuploid cells are more likely to be malignant.

```{r}

run.infercnv=F

if(run.infercnv){
  
  library(infercnv)
  packageVersion('infercnv')
  library(Matrix)
  library(hdf5r)
  '%ni%' <- function(x,y)!('%in%'(x,y))
  options(scipen = 100)
  
  all_genes <- fread(file.path(basedir,'/source_data/fgs.txt'))
  

  song.epi$cellmeta <- 'epi'
  
  these.samples <- unique(song.epi$orig.ident)
  these.samples <- c('I19T')
  
  for(this.sample in these.samples){
    
    seurat_list <- list()
    
    seurat_list[['epi']] <- subset(song.epi, subset = orig.ident == this.sample) 
    seurat_list[['noepi']] <- readRDS(paste0(basedir,'/source_data/scrna/song_2022/infercnv/',this.sample,'/nonepi.',this.sample,'.rds'))
    
    features <- SelectIntegrationFeatures(seurat_list, nfeatures = 2000) # adjust number of features as needed
    
    seurat_list <- purrr::map(seurat_list, ~ RunPCA(.x, verbose = FALSE))
    
    seurat_list <- PrepSCTIntegration(seurat_list, anchor.features = features)
    
    anchors <- FindIntegrationAnchors(
      object.list = seurat_list,
      normalization.method = "SCT",
      anchor.features = features,
      reduction = "rpca",
      k.anchor = 10
      
    )
    
    
    combined <- IntegrateData(anchorset = anchors, normalization.method = "SCT") 
    
    DefaultAssay(combined) <- "integrated"
  
    

    ref_group_names <- c("noepi")
    
    
    annotations <- data.frame(
      Cell = colnames(combined),
      Group = combined$cellmeta
    )
    
    write.table(annotations, paste0("./infercnv/",this.sample,"/cellAnnotations.txt"), 
                sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
    
    gene_names <-  data.frame('gene_name'=rownames(combined@assays$RNA))
    
    gene_names <- left_join(gene_names, all_genes) %>% dplyr::select(gene_name, Chr,  Start,   End)
    duplicate.genes <- gene_names %>% group_by(gene_name) %>% tally() %>% arrange(desc(n)) %>% filter(n > 1)
    
    gene_names <- gene_names %>% filter(gene_name %ni% duplicate.genes$gene_name & !is.na(Chr) & Chr < 23) %>% arrange(Chr, Start,   End)
    
    write.table(gene_names, paste0("./infercnv/",this.sample,"/geneAnnotations.txt"), sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
    
    combined <- subset(combined, features = gene_names$gene_name)
    
    count_mat <- combined@assays$RNA$counts
    
    infercnv.obj <- CreateInfercnvObject(raw_counts_matrix = count_mat,
                                         annotations_file = paste0("./infercnv/",this.sample,"/cellAnnotations.txt"),
                                         delim = "\t",
                                         ref_group_names = ref_group_names ,
                                         gene_order_file = paste0("./infercnv/",this.sample,"/geneAnnotations.txt")
    )
    
    infercnv.res <- infercnv::run(infercnv.obj,
                                  # noise_filter = 0.2,  
                                  # cutoff=0.5,
                                  out_dir=paste0("./infercnv/",this.sample,"/output/"),
                                  cluster_by_groups=TRUE,
                                  denoise=TRUE,  
                                  write_expr_matrix = T,
                                  no_plot=T,
                                  # num_threads=1,
                                  HMM=T
    )
    
    cnv_matrix <- data.frame(infercnv.res@expr.data)
    
    write.table( cnv_matrix,  
                 file = paste0("./infercnv/",this.sample,"/",this.sample,'.cnv_matrix.tsv'), 
                 row.names = T,col.names = T, quote = FALSE, sep = "\t", append =F)
    
  }
}

```

## Predict gCCS from inferCNV profile
We use our gCCS model to predict the CCS of individual cells form the CNV profiles inferred from inferCNV

```{r}

cytoBand <- fread(file.path(basedir,'/source_data/cytoBand.txt'), na.strings = '') %>% 
  filter(!is.na(Chromosome))


cytoBand<- separate(cytoBand, col = Chromosome, into = c('band','period'), sep = '\\.', remove=F, extra = 'drop', fill='right')

cytoBand <- cytoBand %>%
  mutate(
    parts = str_match(band, "^(.*?[pq])(.*)$"),
    arm = parts[,2],
    armband = parts[,3]
  ) %>%
  dplyr::select(-parts) %>% filter(!is.na(arm) & arm %ni% c('Xq','Xp','Yp','Yq'))

full_cnv_model <- readRDS(file.path(basedir,'/results/cnv.model.rds'))

first.sample = T
for(this.sample in unique(song.epi$orig.ident) ){
  
  infercnv.path <- paste0(secdir,"/source_data/scrna/song_2022/infercnv/",this.sample,"/",this.sample,'.cnv_matrix.tsv')
  
  if(file.exists(infercnv.path)){
    
    cat('analyzing ', this.sample,'\n')
    
    suppressWarnings(
    
      infer.cnv.data <- fread(infercnv.path )
    
    )
    
    #get gCMS per cell
    
    segment.gene.cyto <- inner_join(cytoBand, infer.cnv.data,  by=c('Hugo_Symbol'='V1')) %>%
      dplyr::select(-c("Hugo_Symbol" , "Chromosome"  , "band" , "period" ,   "armband" ))
    
    ggplot(segment.gene.cyto, aes(x=band, y=))
    
    
    arm.med <- segment.gene.cyto %>%
      group_by(arm ) %>% 
      summarise(across(where(is.numeric), \(x) median(x, na.rm = TRUE))) 
    
    arm.med.t <- data.frame(t(arm.med[,-1]))
    arm.med.t <- log(arm.med.t, 2)
    names(arm.med.t) <- paste0('arm.',arm.med$arm)
    arm.med.t$mean.vaf <- 0.5
    
    arm.med.t$prob <- predict(full_cnv_model, arm.med.t, type = "response")
    
    #get cnv-score per cell
    arm.med.t$cnv_scores <- sqrt(colMeans(infer.cnv.data[,-1]^2, na.rm = TRUE))
    
    arm.med.t$barcode <-  rownames(arm.med.t)
    
    if(first.sample){cnv.loads <- arm.med.t; first.sample=F} else {cnv.loads <- rbind.data.frame(cnv.loads,  arm.med.t) }
    
  } else {
    
    cat('!!no infercnv profile for ',this.sample,'\n')
    
  }
}

df$barcode <- rownames(df)

df.cnv <- left_join(df, cnv.loads, by=c('barcode'='barcode')) %>% filter(!is.na(mean.vaf))

```

## Visualize the effect of using the gCCS model
The classic "cnv-score" (global sum of copy variant segments) is not very sensitive. When we switch to the probabilities from the gCCS model, we see stronger separation of cells by their NMF signature. The probability threshold set using the bulk samples gave a ton of false positives, so we tried a more strict approach, subsetting to only the top 20% of cells for each CCS. 

```{r}



ggplot(df.cnv , aes(x=cnv_scores, fill=Top_NMF)) +
  geom_vline(xintercept = 1, color="black",  linetype='longdash')+
  
  geom_histogram(position = position_dodge())+
  facet_grid(donor~., scales='free') +
  theme_bw()+
  theme(panel.grid=element_blank(),
        strip.background = element_blank()) +
  
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  ) 


THRESHOLD = 0.34

df.cnv$gCMS <- NA
df.cnv$gCMS[df.cnv$prob > 0.8 ] <- 'CMS-A'
df.cnv$gCMS[df.cnv$prob < 0.2 ] <- 'CMS-B'

ggplot(df.cnv %>% filter(!is.na(donor)), aes(x=prob, fill=Top_NMF)) +
  facet_grid(donor~., scales='free') +
  
  geom_vline(xintercept = THRESHOLD, color="black")+
  

  geom_vline(xintercept = 0.8, color="#04597C",  linetype='dashed')+
  geom_vline(xintercept = 0.2, color="#EDA71A",  linetype='dashed')+
  
  geom_histogram(position = position_dodge())+
  
  theme_bw()+
  theme(panel.grid=element_blank(),
        strip.background = element_blank()) +
  
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  ) 


df.cnv.filt <- df.cnv

```

## eCCS
The TSP algorithm did not function at single cell resolution from high drop-out rates. So we calculate eCCS at the tumor-level from pseudo bulk. We bootstrap our pseudobulk estimate 100 to get a sense in our confidence of our assignment.

```{r}

tumour.subset_obj <- subset(song.epi, cells = df.cnv.filt$barcode) 

whole.estimate <- get_cms_from_pseudobulk(tumour.subset_obj, grouping_var = "orig.ident")
whole.estimate$sample_id <- rownames(whole.estimate)

meta <- tumour.subset_obj@meta.data
meta$cell <- rownames(meta)

for(this.boot in c(1:100)){
  
  seed=format(Sys.time(), "%OS3")
  set.seed(seed)
  
  sampled_cells <- meta %>%
    group_by(orig.ident) %>%
    slice_sample(n = 20, replace = F) %>%
    pull(cell)
  
  
  boot.subset <- subset(tumour.subset_obj, cells = sampled_cells)
  
  suppressMessages(
    boot.estimate <- get_cms_from_pseudobulk(boot.subset, grouping_var = "orig.ident")
  )
  
  boot.estimate$sample_id <- rownames(boot.estimate)
  boot.estimate$boot <- this.boot
  
  if(this.boot == 1){all.boot.estimate <- boot.estimate}else{all.boot.estimate <- rbind.data.frame(all.boot.estimate, boot.estimate)}
  
}

tsp_summary_stats <- all.boot.estimate %>%
  group_by(sample_id) %>%
  summarise(
    median = median(confidence.polarized ),
    mean = mean(confidence.polarized ),
    se = sd(confidence.polarized ) / sqrt(n()),
    ci_low = mean - qt(0.975, df = n() - 1) * se,
    ci_high = mean + qt(0.975, df = n() - 1) * se
  )

tsp.confidence <- left_join(whole.estimate, tsp_summary_stats, by=c('sample_id'='sample_id'))

tsp.confidence$Sample <- factor(tsp.confidence$sample_id, levels = tsp.confidence$sample_id[order(tsp.confidence$confidence.polarized )])

df$condition <- 'iCCA'
cohort <- df %>% dplyr::select(donor, condition) %>% unique()


tsp.confidence <- left_join(tsp.confidence, cohort, by=c('Sample'='donor'))

donor.confidence <- tsp.confidence %>% group_by(Sample) %>% summarise(mean.conf = mean(confidence.polarized, na.rm=T))
tsp.confidence$Donor <- factor(tsp.confidence$Sample, levels = donor.confidence$Sample[order(donor.confidence$mean.conf)])



ggplot(tsp.confidence , aes(x=Donor,  color=rna_class)) + 
  
  geom_errorbar(aes(ymin = ci_low*100, ymax = ci_high*100), width = 0.2)  +
  
  #  geom_line(aes(group=Donor, y=confidence.polarized*100))+
  geom_point(aes(y=confidence.polarized*100), size=1.5)+
  geom_point(aes(y=mean*100), shape=1)+
  
  facet_grid(. ~  Donor+condition , space='free_x',scales='free_x') +
  
  theme_bw(base_size=10) +scale_color_manual(values =  
                                               c(
                                                 "CMS-A"="#04597C", "CMS-B"="#EDA71A"
                                                 
                                                 
                                               ), na.value = 'grey'
  ) +labs(y='% of TSP for CMS-A')+ theme(panel.grid = element_blank(), 
                                         axis.ticks.x = element_blank(),
                                         axis.title.x = element_blank(),
                                         axis.text.x = element_blank(),
                                         
                                         #strip.text.x = element_text(size = 6),
                                         
                                         plot.background=element_blank(),
                                         panel.background = element_blank(),
                                         panel.spacing = unit(0, "lines"),
                                         strip.background = element_blank()) + guides(color='none')+ ylim(0,100)


df.cms <- left_join(df.cnv.filt, tsp.confidence, by=c('donor'='sample_id'))

df.cms.n <- df.cms %>% group_by(Donor, condition, rna_class, confidence.polarized, Top_NMF) %>% tally() 
df.cms.n$esig <- factor(df.cms.n$Top_NMF, levels = paste0('sig.',c(1:10)))


ggplot(df.cms.n , aes(x=Donor, y=n, fill=esig)) + 
  geom_bar(stat='identity')+
  facet_grid(.~Donor+condition , space='free_x',scales='free_x' ) +
  theme_bw(base_size=10)+ 
  theme(panel.grid = element_blank(), 
        legend.position = 'bottom',
        legend.background = element_blank(),
        legend.text = element_text(size = 6), 
        legend.key.size = unit(0.5, "lines"),     
        plot.background=element_blank(),
        panel.background = element_blank(),
        legend.title = element_blank(),
        
        axis.title.x =  element_blank(),
        
        
        panel.spacing = unit(0, "lines"),
        strip.text = element_blank(),
        strip.background = element_blank())+ labs(y='n of aneuploid cells')+
  scale_fill_manual(values =  
                      color.set, na.value = 'grey'
  )

```

## cell gCCS to tumour eCCS
Combine the cell-level genetic CCS calls to the pseudo-bulk eCCS calls and to the single cell NMF-signature assignments to show all three lines of evidence line up. Two samples had inclusive pseudo-bulk assignment and were removed in this step.

```{r}

tumour.cnv.arms <- df.cnv.filt %>%  dplyr::select(c("donor","barcode","gCMS","prob") |  starts_with("arm.") )

tumour.cnv.arms <- tumour.cnv.arms %>% 
  reshape2::melt(id.vars=c("donor",'barcode','prob','gCMS'))

tumour.cnv.arms$band <- gsub("arm.","", tumour.cnv.arms$variable)

tumour.cnv.arms <- tumour.cnv.arms %>%
  dplyr::mutate(
    arm = substr(band, nchar(band), nchar(band)),
    chr = substr(band, 1, nchar(band) - 1)
  ) %>%
  dplyr::select(donor,barcode, prob, gCMS, chr, arm,band, value)

model.coeff <- data.frame('band'=gsub('arm.','',names(full_cnv_model$coefficients))) %>%  mutate(
  arm = substr(band, nchar(band), nchar(band)),
  chr = substr(band, 1, nchar(band) - 1)
) %>%   select(chr, arm, band)

tumour.cnv.model.arms <- tumour.cnv.arms %>% filter(band %in% model.coeff$band & !is.na(gCMS))
tumour.cnv.model.arms$Band <- factor(tumour.cnv.model.arms$band, levels = paste0(rep(1:21, each = 2), c("p", "q")) )

tumour.cnv.model.arms %>%
  group_by(donor) %>%
  summarise(
    most_common = names(sort(table(gCMS), decreasing = TRUE))[1],
    percentage = max(table(gCMS)) / n() * 100
  ) %>% summarise(mean_percentage=mean(percentage), min_perc=min(percentage), max_perc=max(percentage))


tumour.cnv.model.arms$eCMS <- NA
tumour.cnv.model.arms$eCMS[tumour.cnv.model.arms$donor %in% tsp.confidence$Donor[tsp.confidence$rna_class == 'CMS-A']] <- 'CMS-A'
tumour.cnv.model.arms$eCMS[tumour.cnv.model.arms$donor %in% tsp.confidence$Donor[tsp.confidence$rna_class == 'CMS-B']] <- 'CMS-B'


tumour.cnv.model.arms.match <- tumour.cnv.model.arms %>% filter(!is.na(eCMS) & !is.na(gCMS)) #filter(eCMS == gCMS)
tumour.cnv.model.arms.match$Donor <- factor(tumour.cnv.model.arms.match$donor, levels = donor.confidence$Sample[order(donor.confidence$mean.conf)])




df.cms.n.filt <- df.cms %>% filter(barcode %in% tumour.cnv.model.arms.match$barcode) %>% 
  group_by(Donor, condition, rna_class, confidence.polarized, Top_NMF, gCMS) %>% tally() 
df.cms.n.filt$n[df.cms.n.filt$gCMS == 'CMS-A'] <- -df.cms.n.filt$n[df.cms.n.filt$gCMS == 'CMS-A']

df.cms.n.filt$esig <- factor(df.cms.n.filt$Top_NMF, levels = paste0('sig.',c(1:10)))


df.cms.n.filt$gCMS.f <- factor(paste0('g',df.cms.n.filt$gCMS), levels = rev(c('gCMS-A','gCMS-B')))




nmf.tally.gCMS.plot <- 
         plot_grid(
          
           ggplot(tsp.confidence %>% filter(Donor %in% df.cms.n.filt$Donor), aes(x=Donor,  color=rna_class)) + 
             
             geom_hline(yintercept = 80, color='grey', linetype='dotted')+
             geom_hline(yintercept = 20, color='grey', linetype='dotted')+
             
             geom_errorbar(aes(ymin = ci_low*100, ymax = ci_high*100), width = 0.2)  +
             
             geom_point(aes(y=confidence.polarized*100), size=1.5)+
             geom_point(aes(y=mean*100), shape=1)+
             
             facet_grid(. ~  Donor , space='free_x',scales='free_x') +
             
             theme_bw(base_size=10) +scale_color_manual(values =  
                                                          c(
                                                            "CMS-A"="#04597C", "CMS-B"="#EDA71A"
                                                            
                                                            
                                                          ), na.value = 'grey'
             ) +labs(y='% of TSP')+ theme(panel.grid = element_blank(), 
                                                     axis.ticks.x = element_blank(),
                                                     axis.title.x = element_blank(),
                                                     axis.text.x = element_blank(),
                                                     
                                                    
                                                      strip.text.x = element_text(size = 4,margin = margin(t = 2, b = 2)),
                                                     plot.background=element_blank(),
                                                     panel.background = element_blank(),
                                                     panel.spacing = unit(0, "lines"),
                                                     strip.background = element_blank()) + guides(color='none')+
             scale_y_continuous(limits = c(0,100), breaks = c(20,80))
           
           
           ,NULL,
           ggplot(df.cms.n.filt, aes(x=Donor, y=n, fill=esig)) + 
             geom_bar(stat='identity')+
             facet_grid(gCMS.f~Donor , space='free',scales='free' ) +
             theme_bw(base_size=10)+ 
             theme(panel.grid = element_blank(), 
                   legend.position = 'bottom',
                   legend.background = element_blank(),
                   legend.text = element_text(size = 6), 
                   legend.key.size = unit(0.5, "lines"),     
                   plot.background=element_blank(),
                   panel.background = element_blank(),
                   legend.title = element_blank(),
                   
                   axis.title.x =  element_blank(),
                   axis.text.x =  element_blank(),
                   axis.ticks.x =  element_blank(),
                   
                   
                   panel.spacing = unit(0, "lines"),
                   strip.text.x = element_blank(),
                   strip.background = element_blank()
             )+ labs(y='n cells')+
             guides(color='none',fill='none')+
             scale_fill_manual(values =  
                                 color.set, na.value = 'grey'
             )
           ,ncol = 1, align = 'hv', axis = "rlbt", 
           rel_heights =c(0.3,-0.05,0.7)) 
       
nmf.tally.gCMS.plot

ggsave(file.path(basedir, "/results/fig/FIG3h.nmf.tally.gCMS.svg"), plot = 
         nmf.tally.gCMS.plot
       , device = "svg", height = 65,width = 65, units = 'mm', dpi=72 )


ggplot(df.cms %>% filter(Donor %in% df.cms.n.filt$Donor), aes(x = umap_1, y = umap_2)) +
  
  geom_point(size=0.01, aes(color = Top_NMF)) +
  geom_label_repel(data = centroids %>% filter(donor %in% df.cms.n.filt$Donor), aes(x = x_centroid, y = y_centroid, label = donor), 
                   size=2,   box.padding = 0.5, force = 0.5, max.time = 4,  min.segment.length = 0, max.overlaps =15) +
  theme_void() +  # removes all axes and background
  scale_color_manual(values =  
                       color.set, na.value = 'white'
                     
                     
  ) + 
  guides(fill='none',color='none')+
  theme(
    panel.border = element_blank(),
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
    plot.background = element_blank()
  ) 


```

