# Small substitution drivers
Working with the gene-centric dndscv approach and the gene-agnostic ActiveDriverWGS approach, we explore small substitution (snv to in/del size) drivers.

## Prepare data

```{r}

#BiocManager::install(c('BSgenome.Hsapiens.UCSC.hg19', 'BSgenome.Hsapiens.UCSC.hg38', 'BSgenome.Mmusculus.UCSC.mm9', 'BSgenome.Mmusculus.UCSC.mm10'))
#devtools::install_github('https://github.com/reimandlab/ActiveDriverWGSR')
#devtools::install_github("im3sanger/dndscv")

suppressWarnings(
suppressPackageStartupMessages({
  library(data.table)
  library(tidyr)
  library(dplyr)
  library(GenomicRanges)
  #library(CNTools)
  library(ActiveDriverWGS)
  library(dndscv)
  library(ggplot2)
  library(ggrepel)
})
)

packageVersion('ActiveDriverWGS')
packageVersion('dndscv')

'%ni%' <- function(x,y)!('%in%'(x,y))

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'

source(paste0(code.dir,"/bin/driver_functions.R"))
source( file.path(data.dir, "paths.R"))
cluster.dir = file.path(cluster.dir, 'more.data')


##define genes (using dnds Granges)
dndscv_reference_data_path <- file.path(data.dir,'/source_data/RefCDS_human_GRCh38_GencodeV18_recommended.rda')
load(file=dndscv_reference_data_path)

candidate_gene_exons <- data.frame(chr=paste0('chr',seqnames(gr_genes)),
                           start=start(gr_genes),
                           end=end(gr_genes),
                           id=gr_genes$names)

##writes to file
make_vcfs=F
if(make_vcfs){
  ## define sample set ##
  
  sample.list <- fread(file.path(data.dir,'/source_data/sample.list.txt'), header=F)
  
  collate_barebones_vcf_write(sample.list, 
                              inpath=cluster.dir, 
                              outpath=file.path(data.dir,'/results/'), 
                              variant_types = c("indel","snv"))
}


all_mutations <- read.delim(file.path(data.dir,"/results/all_mutations.txt"))
length(unique(all_mutations$patient))

all_mutations <- all_mutations %>% arrange(patient,chr,pos1,pos2)

#removes scaffolds without candidate genes
all_mutations <- all_mutations[all_mutations$chr %in% unique(candidate_gene_exons$chr),]
#also a check that the candidate annotation and mutation annotation match

dndscv_muts <- all_mutations[,c("patient", "chr", "pos1", "ref", "alt")]
names(dndscv_muts) <- c("sampleID", "chr", "pos", "ref", "mut")
dndscv_muts$chr <- gsub(pattern = 'chr', replacement = '', dndscv_muts$chr)

```

## run driver analyses
activedriver takes up to 12 hours so we don't run that every time.

```{r}

get_drivers = F
if(get_drivers){
  
  result <- tryCatch({
    dndsout = dndscv(dndscv_muts, refdb =  dndscv_reference_data_path, outp = 3, cv = NULL)
  }, error = function(e) {
    message("Error: ", e)
    NULL
  })
  

  dndsout.annotmuts <- dndsout$annotmuts
  
  write.table(
    dndsout$sel_cv,
    file = paste0(data.dir, "/results/sel_cv.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
  write.table(
    dndsout$annotmuts,
    file = paste0(data.dir, "/results/dnds_annotation.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
  active_driver_results = ActiveDriverWGS(mutations = all_mutations, 
                                          elements = candidate_gene_exons, 
                                          ref_genome = 'hg38')
  
  write.table(
    active_driver_results,
    file = paste0(data.dir, "/results/active_driver_results.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
  #### promoters ####
  regulatory_features <- fread(file.path(data.dir,'source_data/homo_sapiens.GRCh38.Regulatory_Build.regulatory_features.20231016.gff'))
  regulatory_features$V1 <- paste0('chr',regulatory_features$V1)
  promoters <- regulatory_features[regulatory_features$V3 == 'promoter',]
  
  #https://genome-euro.ucsc.edu/cgi-bin/hgTables?db=hg38&hgta_group=hub_669493&hgta_track=hub_669493_epdnew&hgta_table=hub_669493_epdnew6&hgta_doSchema=describe+table+schema
  #EPD based on experimental and bioinformatics TSS, so likely more strict
  EPD_promoters <- fread(file.path(data.dir,'source_data/Hs_EPDnew_006_hg38.bed'))
  
  promoters$epd <- NA
  for(this.epd.promoter in EPD_promoters$V4){
    
    this.chr <- EPD_promoters$V1[EPD_promoters$V4 == this.epd.promoter]
    this.start <- EPD_promoters$V2[EPD_promoters$V4 == this.epd.promoter]
    this.end <- EPD_promoters$V3[EPD_promoters$V4 == this.epd.promoter]
    
    this.ens.match <- promoters$epd[(promoters$V4 - 20) <= this.start & this.end  <= (promoters$V5 + 20) & promoters$V1 == this.chr ] 
    if(length(this.ens.match) == 0 ){
      cat(this.epd.promoter,' not found\n')
    } else if(is.na(this.ens.match)){
      promoters$epd[promoters$V4 <= this.start & this.end  <= promoters$V5 & promoters$V1 == this.chr ] <- this.epd.promoter
    } else {
      promoters$epd[promoters$V4 <= this.start & this.end  <= promoters$V5 & promoters$V1 == this.chr ] <- paste(this.ens.match, this.epd.promoter, sep = '; ')
      
    }
    
  }
  
  promoters <- promoters[!is.na(promoters$epd),]
  
  candidate_promoters <- data.frame(chr=promoters$V1,
                                    start=promoters$V4,
                                    end=promoters$V5,
                                    id=promoters$epd)
  
  
  all_mutations <- fread(file.path(data.dir,"/results/all_mutations.txt"))
  all_mutations <- all_mutations[all_mutations$chr %in% unique(candidate_promoters$chr),]
  
  active_driver_results.promoters = ActiveDriverWGS(mutations = all_mutations, 
                                                    elements = candidate_promoters, 
                                                    ref_genome = 'hg38')
  
  write.table(
    active_driver_results.promoters,
    file = file.path(data.dir,"/results/active_driver_results.promoters.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
}

```

## review results
Let's compare drivers discovered by the two methods. We manually review cases with mutations in new drivers (eg PTH2) using IGV; spoiler alert: they are artefacts.

```{r}

active_driver_results <- fread(paste0(data.dir, "/results/active_driver_results.txt"))
active_drivers <- active_driver_results$id[active_driver_results$fdr_element < 1 & active_driver_results$element_muts_obs > 4]

sel_cv <- fread(paste0(data.dir, "/results/sel_cv.txt"))
dnds_annotation <- fread(paste0(data.dir, "/results/dnds_annotation.txt"))

all_genes <- fread(file.path(data.dir,'/source_data/fgs.txt')) 
gene_set <- all_genes[all_genes$biotype=="protein_coding" & all_genes$Length >= 500,]
#cancerGeneList <- fread(file.path(data.dir,'/source_data/cancerGeneList.tsv'))


positioned_drivers <- left_join( active_driver_results, all_genes, by=c("id"="gene_name"))
positioned_drivers_dnds <- full_join( positioned_drivers, sel_cv, by=c("id"="gene_name"))

positioned_drivers_dnds$chr_fac <- factor(gsub("chr","",positioned_drivers_dnds$Chr), levels=c(1:22,'X'))
positioned_drivers_dnds <- positioned_drivers_dnds %>% filter(!is.na(chr_fac))

positioned_drivers_dnds$fdr_element_sigfig <- signif(positioned_drivers_dnds$fdr_element, 2)
positioned_drivers_dnds$qglobal_cv_sigfig <- signif(positioned_drivers_dnds$qglobal_cv, 2)

## filter False Positives
#PTH2: low VAF, in repeat region, same spot
#USP8: unsure, complex recurrent mutation: always same two in-phase mutations ~8 bp apart
positioned_drivers_dnds <- positioned_drivers_dnds %>% filter(id %ni% c('PTH2','USP8')) 

sig_id <- positioned_drivers_dnds$id[positioned_drivers_dnds$fdr_element_sigfig < 0.05 | positioned_drivers_dnds$qglobal_cv_sigfig < 0.05] 

positioned_drivers_dnds.plot <- positioned_drivers_dnds
positioned_drivers_dnds.plot$fdr_element_sigfig[-log(positioned_drivers_dnds.plot$fdr_element_sigfig) > 30] <- 0

#subsampling rows in bottom left corner of dots that would just stack, for file size
keep_rows <- positioned_drivers_dnds.plot[positioned_drivers_dnds.plot$fdr_element_sigfig < 1 |positioned_drivers_dnds.plot$qglobal_cv_sigfig < 1 ]
sample_rows <- positioned_drivers_dnds.plot[positioned_drivers_dnds.plot$fdr_element_sigfig == 1, ]
sampled_subset <- sample_rows[sample(nrow(sample_rows), 20), ]

positioned_drivers_dnds.plot <- rbind.data.frame(keep_rows, sampled_subset)

drivers.pvalues.plot <- 
ggplot(positioned_drivers_dnds.plot, aes(y=-log(fdr_element_sigfig), x=-log(qglobal_cv_sigfig))) +
  
  geom_vline(xintercept = -log(0.05), color='grey', linetype='dotted')+
  geom_hline(yintercept = -log(0.05), color='grey', linetype='dotted')+
  
  geom_point(size=2.5, color='grey') +  
  geom_point(size=2.5, data = ~ subset(., id %in% sig_id)) +  
  geom_label_repel(aes(label=id),  
                   data = ~ subset(., id %in% sig_id), size=3, max.overlaps =20,
                   box.padding = 0.75, force = 0.1, label.padding = 0.1,
                   max.time = 3) +
  
  theme_bw(base_size=10) + 
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) + 
  
  labs(y="region enrichment FDR (log)",x="dnds q-value (log)")

drivers.pvalues.plot

ggsave(file.path(data.dir, "/results/fig/FIG4A.drivers.pvalues.svg"), plot = 
       drivers.pvalues.plot  
, device = "svg",width = 75, height = 75 , units='mm', dpi=72 )

```

## analyze drivers by subtype
We also compared global signals of selection between the subtypes - although these results were tantalizing, we found the conclusion were extremely inconsistent and the signal could completely flip just by the addition/removal of a couple samples - so this was ultimately removed from the final manuscript.

```{r}

predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt') )

cms.drivers.pvalues.plot <- compare_ccs_dnds(dndscv_muts, predicted.w.score, 'all')
  
cms.drivers.pvalues.plot

ggsave(paste0(data.dir, "/results/fig/FIGS4b.ccs.drivers.pvalues.svg"), plot = 
  cms.drivers.pvalues.plot         
, device = "svg",width = 65, height = 65 , units='mm', dpi=72 )

clinic.sample.info <- fread(file.path(data.dir,"/results/samples.info.txt"), header=T)
primaries <- clinic.sample.info$Sample[clinic.sample.info$Sample_Progression == 'Primary']
naive <- clinic.sample.info$Sample[clinic.sample.info$tissue_before_chemo == T]

ccs.dnds.primaries.plot <- 
  compare_ccs_dnds(dndscv_muts %>% filter(sampleID %in% primaries), 
                 predicted.w.score, 
                 'primaries')

ccs.dnds.primaries.plot

ggsave(paste0(data.dir, "/results/fig/FIGS4b.ccs.drivers.primaries.svg"), plot = 
  ccs.dnds.primaries.plot         
, device = "svg",width = 65, height = 65 , units='mm', dpi=72 )

ccs.dnds.naive.plot <- 
  compare_ccs_dnds(dndscv_muts %>% filter(sampleID %in% naive), 
                 predicted.w.score, 
                 'treatment naive')

ccs.dnds.naive.plot

ggsave(paste0(data.dir, "/results/fig/FIGS4b.ccs.drivers.naive.svg"), plot = 
  ccs.dnds.naive.plot         
, device = "svg",width = 65, height = 65 , units='mm', dpi=72 )

```

## test loss-of-heterozygosity
I think this is a new type of analysis. What we wanted to do originally was rescue tumor suppressor genes that were just under the FDR threshold by showing that there was more LOH at those sites than expected by chance. We actually didn't end up recovering anything this way, but we found interesting results nonetheless. 

```{r}

gene_set <- fread(file.path(data.dir,'/source_data/fgs.txt'))
gene_set$Chr <- paste0('chr', gene_set$Chr)

genebed <- gene_set
genebed <- genebed %>% filter(biotype=="protein_coding" & Length >= 500)
genebed <- genebed %>% dplyr::select(Chr, Start,   End,ensid, gene_name)
names(genebed) <- c('chrom',	'start',	'end',	'geneid',	'genename')

dnds_annotation <- fread(paste0(data.dir, "/results/dnds_annotation.txt"))

get_loh= F
if(get_loh){
  
  
  all.cnv <- fread(file.path(cluster.dir, 'all.cnv.txt'))
  length(unique(all.cnv$sampleID))
  
  all.cnv$toy.marker <- 1
  all.cn <- all.cnv %>% dplyr::select(sampleID, chrom,start.pos,end.pos,toy.marker, imean) 
  names(all.cn) <- c('ID','chrom','loc.start','loc.end','num.mark','seg.mean')
  
  all.cnv <- all.cnv %>% dplyr::select(sampleID, chrom,start.pos,end.pos,toy.marker, labels) %>% filter(!is.na(labels))
  names(all.cnv) <- c('ID','chrom','loc.start','loc.end','num.mark','seg.mean')
  
  allele.cn.seg <- CNSeg(all.cnv)
  allele.cn.gene <- getRS(allele.cn.seg, by="gene", imput=FALSE, XY=FALSE, geneMap=genebed, what="min")
  gene.allele <- rs(allele.cn.gene)
  
  these.genes = intersect(unique(dnds_annotation$gene), gene.allele$genename)
  
  rescue_id <- positioned_drivers_dnds$id[positioned_drivers_dnds$element_enriched == T & positioned_drivers_dnds$pglobal_cv < 0.05] 
  
  these.genes = intersect(these.genes, rescue_id)
  
  
  first.row = T 
  for(this.gene in these.genes){
    
    this.cn <- gene.allele[gene.allele$genename == this.gene,]
    if(nrow(this.cn) == 1){
      
      this.cn <- as.data.frame(t(this.cn[,-c(1:5)]))
      names(this.cn) <- 'cn'
      
      these.alleles <- separate(this.cn, col = 'cn', into = c('a','b'), sep = '\\.')
      these.alleles$loh <- F
      these.alleles$loh[these.alleles$a == 0 | these.alleles$b == 0  ] <- T
      
      this.annot <- dnds_annotation[dnds_annotation$gene == this.gene,]
      these.alleles$hit <- F
      these.alleles$hit[row.names(these.alleles) %in% this.annot$sampleID] <- T
      
      this.table <- these.alleles %>% group_by(loh,hit) %>% tally() %>% reshape2::dcast(formula = loh ~ hit, value.var = 'n', fill = 0  )
      
      if(nrow(this.table) * ncol(this.table) == 6){
        
        fish.res <- fisher.test(this.table[,-1], conf.int = T)
        
        tmp.tbl <- cbind.data.frame('gene'=this.gene, 
                                    'loh.pvalue'=fish.res$p.value, 
                                    'loh.TTn'= this.table[2,3],
                                    'loh.LOR'=log(fish.res$estimate)
        )
        
        if(first.row){
          loh.hit.table<- tmp.tbl
          first.row =F 
        } else {
          loh.hit.table<- rbind.data.frame(loh.hit.table, tmp.tbl)
          
        }
      }
    } else {
      message('two rows with the gene name: ',this.gene)
    }
    
  }
  loh.hit.table <- loh.hit.table %>% arrange( loh.pvalue)
  write.table(
    loh.hit.table,
    file = file.path(data.dir, "/results/loh.p.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
}

loh.hit.table <- fread(file.path(data.dir, "/results/loh.p.txt"))

loh.hit.table$fdr <- p.adjust(loh.hit.table$loh.pvalue, method='fdr')
loh_sig_id <- loh.hit.table$gene[loh.hit.table$fdr < 0.05 ]

drivers.loh.plot <- 
ggplot(loh.hit.table, aes(x=loh.LOR, y= -log(fdr))) +
  
  geom_hline(yintercept = -log(0.05), color='grey', linetype='dotted')+
  
  geom_vline(xintercept = 0, color='grey')+
  geom_point(size=2.5, color='grey') +  
  geom_point(size=2.5, data = ~ subset(., gene %in% loh_sig_id)) +  
  geom_label_repel(aes(label=gene),  data = ~ subset(., gene %in% loh_sig_id), size=3, 
                   box.padding = 0.5, force = 1, max.time = 3,  label.padding = 0.1, min.segment.length = 0) +

  
  theme_bw(base_size=10) + 
  
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) + labs(x='LOH + mutant odds (log)', y='FDR (-log)')

drivers.loh.plot

ggsave(paste0(data.dir, "/results/fig/FIGS4d.drivers.loh.svg"), plot = 
         drivers.loh.plot
, device = "svg",width = 60, height = 75 , units='mm' , dpi=72 )

```

## combine LOH and dnds for table

```{r}

smm.drivrs.analysis <- full_join(positioned_drivers_dnds, loh.hit.table, by=c('id'='gene'))

positioned_drivers_only <- smm.drivrs.analysis[smm.drivrs.analysis$fdr_element_sigfig < 0.05  | 
                                                 
                                                 smm.drivrs.analysis$qglobal_cv_sigfig < 0.05 ,]


positioned_drivers_only$n_mut <- positioned_drivers_only$n_syn + positioned_drivers_only$n_mis + positioned_drivers_only$n_non +
  positioned_drivers_only$n_spl + positioned_drivers_only$n_ind 

positioned_drivers_only$loh.fdr <- p.adjust(positioned_drivers_only$loh.pvalue, method = "BH")


positioned_drivers_table <- positioned_drivers_only %>% 
  dplyr::select(id, chr_fac, n_mut, fdr_element_sigfig,  qglobal_cv_sigfig, loh.fdr, loh.TTn) %>% arrange(desc(n_mut))


format_sigfig <- function(x, sigfig = 2) {
  if (is.numeric(x)) {
    sapply(x, function(val) {
      if (is.na(val)) return(NA)
      if (abs(val) < 0.0001 && val != 0) {
        format(signif(val, sigfig), scientific = TRUE)
      } else {
        format(signif(val, sigfig), scientific = FALSE, trim = TRUE)
      }
    })
  } else {
    x
  }
}


positioned_drivers_table[] <- lapply(positioned_drivers_table, format_sigfig)

write.table(
  positioned_drivers_table,
  file = paste0(data.dir, "/results/TableS3.positioned_drivers_table.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

positioned_drivers_table

```

