# Analyze recurrence and impact of structural variants
Here, we want to find structural variants that reccur more than expected using fishhook. The positive control here is FGFR2 (fusions). We found a ton of hits but many of them seemed suspicious as functional drivers. So we overlapped the recurrence analysis with strctural variants that have a significant effect on expression (this is the same method undelrying the identification of expression quatitiative trait loci eQTL in germline analyses). The eQTL analysis also returned a ton of hits, but many were very rare with huge effect sizes. We also look at extrachromosomal DNA in this analysis. 



## Data prep
Pull data from hpc

```{r}
#library(devtools)
#devtools::install_github('mskilab-org/gUtils')
#devtools::install_github('mskilab-org/fishHook')
#devtools::install_github('mskilab-org/gTrack')
#devtools::install_github('mskilab-org/skitools') ##corrupt
#BiocManager::install("ggbio")

suppressPackageStartupMessages({

  library(rtracklayer)
  library(gTrack)
  library(BSgenome)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(ggbio)
  library(GenomicRanges)
  library(data.table)
  library(dplyr)
  library(reshape2)
  library(circlize)
  library(tidyr)
  library(PNWColors)
  library(ggrepel)
  
})

'%ni%' <- function(x,y)!('%in%'(x,y))
basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'
hpcdir = '/Volumes/pcsi/users/fbeaudry/more.data/'

source( file.path(basedir, "bin/integrated_driver.functions.R"))

sample_list <- read.table(file.path(basedir,'/source_data/sample.list.txt'), header=F)

rerun.pulls = F
if(rerun.pulls){
  first.sv = T

  for(tumour_id in sample_list$V2){
    cat('processing ',tumour_id,'\n')
    donor_id <- sample_list$V1[sample_list$V2 == tumour_id]
    
    sv.cnv.file <- paste0(hpcdir,tumour_id,'/sv.strict/',tumour_id,'.sv.cnv.txt')
    
    if(file.exists(sv.cnv.file)){
      sv.cnv <- fread(sv.cnv.file)
      
      if(first.sv ){
        full.sv.cnv<-sv.cnv
        first.sv=F
      }else{
        full.sv.cnv<-rbind.data.frame(full.sv.cnv,sv.cnv)
      }
    } else {
      cat('skipping ',sv.cnv.file,'\n')
    }

  }
  
  full.sv.cnv$chr <- factor(full.sv.cnv$CHROM, levels = paste0('chr', c(c(1:22), 'X', 'Y')))
  
  write.table(
    full.sv.cnv,
    file = file.path(secdir,'/results/sv.cnv.txt'),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )

}

```


## breakpoint eQTL
Find structural variants that have a significant (and large) impact on nearby gene expression (+- 5000bp)

```{r}

predicted.w.score <- fread(file.path(basedir,'/results/all.eCMS.txt') )
predicted.w.score <- predicted.w.score %>% filter(`Sample Source(FFPE, FF)` == 'FF')

full.sv.cnv <- fread(file.path(secdir,'/results/sv.cnv.txt'))

all.LBR.rna.df <- fread(file.path(basedir,"/results/rna.tpm.ff.txt"), header = T)
all.LBR.rna <- as.matrix(all.LBR.rna.df[,-1])
rownames(all.LBR.rna) <- all.LBR.rna.df$gene_name

gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))
gene_set$Chr <- paste0('chr', gene_set$Chr)
gene_list <- gene_set$gene_name[gene_set$biotype=="protein_coding" & 
                                  gene_set$Length >= 500 &
                                  gene_set$Chr %in% paste0('chr',c(1:22)) &
                                  gene_set$gene_name %in% rownames(all.LBR.rna)
]

# ~ mean(gene_set$Length)
WINDOW_SIZE = 5000

run.eqtl = F
if(run.eqtl){
  
  first.loop = T
  for(this.gene in gene_list){
    
    rearranged.tumours <- get.sv.window(this.gene, gene_set, WINDOW_SIZE, full.sv.cnv)
    
    this_rna <- all.LBR.rna[this.gene,]
    this_rna_full <- as.data.frame(this_rna)
    this_rna_full$tumour <- rownames(this_rna_full)
    
    this_rna_full$is.sv <- 0
    this_rna_full$is.sv[this_rna_full$tumour %in% rearranged.tumours$tumour] <- 1
    
    for(this.cms in c('all','CMS-A','CMS-B')){
      
      if(this.cms %in% c('CMS-A','CMS-B')){
        
        this_rna_t <- this_rna_full %>% filter(tumour %in% predicted.w.score$sample[predicted.w.score$rna_class == this.cms])
        
      } else {
        this_rna_t <- this_rna_full 
      }
      
      if(all(sum(this_rna_t$is.sv) > 2 & sum(this_rna_t$this_rna) > 1) & sum(this_rna_t$is.sv) < nrow(this_rna_t)){
          
        this_rna_t <- this_rna_t %>%
          group_by(is.sv) %>%
          mutate(weight = 1 / n()) %>%
          ungroup()
        
        this_rna_t$weight_prop <- this_rna_t$weight / sum(this_rna_t$weight )
        
        tpm.sv.lm <- lm(log(as.numeric(this_rna) + 1) ~ is.sv, data=this_rna_t, weights = this_rna_t$weight_prop)
        tpm.sv.lm.summary <- summary(tpm.sv.lm)
        
        this.length = (gene_set$End[gene_set$gene_name == this.gene] + WINDOW_SIZE) - (gene_set$Start[gene_set$gene_name == this.gene] - WINDOW_SIZE)
        
        this.p <- cbind.data.frame('gene'=this.gene, 
                                   
                                   'chr'=gene_set$Chr[gene_set$gene_name == this.gene], 
                                   'win.start'=gene_set$Start[gene_set$gene_name == this.gene] - WINDOW_SIZE, 
                                   'win.end'=gene_set$End[gene_set$gene_name == this.gene] + WINDOW_SIZE, 
                                   'true_window_size' = this.length,
                                   'n'=length(unique(rearranged.tumours$tumour)),
                                   
                                   'lm.p'= tpm.sv.lm.summary$coefficients[2,4],
                                   'lm.fstat' = tpm.sv.lm.summary$fstatistic[['value']],
                                   'lm.estimate' = tpm.sv.lm.summary$coefficients[2,1],
                                   'lm.tstat' = tpm.sv.lm.summary$coefficients[2,3]
        )
        
  
        if(first.loop){sv.eqtl <- this.p; first.loop=F} else {sv.eqtl <- rbind.data.frame(sv.eqtl, this.p)}
      
      }
      
    }
    
  }

  write.table(
    sv.eqtl,
    file = paste0(basedir, "/results/sv.eqtl.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
}

sv.eqtl <- fread(paste0(basedir, "/results/sv.eqtl.txt"))

sv.eqtl$eqtl.fdr <- p.adjust(sv.eqtl$lm.p, method = "BH")

sv.eqtl$chr_fac <- factor(gsub("chr","",sv.eqtl$chr), levels=c(1:22,'X'))

large.sv.eqtl <- sv.eqtl %>% filter(eqtl.fdr < 0.05 & abs(lm.estimate) > 1)

ggplot(sv.eqtl %>% filter(cms != 'all'), aes(y=-log(eqtl.fdr), x=win.start)) + 
  
  geom_line()+
  facet_grid(.~chr_fac, space='free',scales='free')+
  
  geom_label_repel(aes(label=gene,color=cms),    data = ~ subset(., eqtl.fdr < 0.05 & abs(lm.estimate) > 1), size=2, 
                  box.padding = 1, force = 4, max.time = 2,  min.segment.length = 0) +
  
  geom_point(   data = ~ subset(., eqtl.fdr < 0.05 & abs(lm.estimate) > 1), aes(color=cms, size=abs(lm.estimate) )) +
  
  scale_color_manual(values = c(
    "all" = 'grey',
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" 
  )) +
  theme_bw(base_size=12) +
  theme(strip.background = element_blank(),
        
        panel.spacing.y = unit(0,'lines'),
        legend.position="top",
        
        axis.title.y =element_blank(),
        axis.text=element_blank(),
        axis.ticks  =element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='transparent'), #transparent panel bg
        plot.background = element_rect(fill='transparent', color='transparent'), #transparent plot bg
        panel.border = element_blank(),
        
        #strip.text.x = element_blank(),
        strip.text.y = element_text(size = 18, angle = 0),
        strip.placement = "outside") + coord_cartesian(clip = "off")

sv.eqtl.cms <- sv.eqtl %>% filter(cms != 'all' & eqtl.fdr < 0.05 & n < 50 )
sv.eqtl.cms$effect.size.mod <- abs(sv.eqtl.cms$lm.estimate)
sv.eqtl.cms$effect.size.mod[sv.eqtl.cms$cms == 'CMS-A'] <- -sv.eqtl.cms$effect.size.mod[sv.eqtl.cms$cms == 'CMS-A']

eqtl.n <- sv.eqtl.cms %>% group_by(gene) %>% tally()

sv.eqtl.cms <- sv.eqtl.cms %>% filter(gene %ni% eqtl.n$gene[eqtl.n$n > 1])

sv.eqtl.cms$Gene <- factor(sv.eqtl.cms$gene, levels=sv.eqtl.cms$gene[order(sv.eqtl.cms$effect.size.mod)])

sv.eqtl.plot <- 
ggplot(sv.eqtl.cms, aes(x = effect.size.mod, y = -log(eqtl.fdr))) +
  geom_hline(yintercept = 0, color='grey')+
  
  #geom_point(aes(color=cms, size=-log(eqtl.fdr))) +
  geom_point(aes(color=cms, size=n)) +
  geom_point(aes( size=n), shape=1, color='white', stroke = 0.5, alpha=0.2) +
  
  geom_label_repel(aes(label=gene,color=cms),    data = ~ subset(., gene %in% c('TERT','CDK6','FGFR2','MDM2')), size=2, 
                   box.padding = 1, force = 4, max.time = 2,  min.segment.length = 0) +
  

  scale_color_manual(values = c(
    "all" = 'grey',
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" 
  )) +
  theme_bw(base_size=10) +
  guides(color='none')+
  scale_size(range = c(1,8))  +
  
  theme(
        
    plot.margin = unit(c(2, 0, 0, 2), "mm"),
   # axis.text.x=element_blank(),
   # axis.ticks.x  =element_blank(), 
legend.position = 'bottom',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='transparent'), #transparent panel bg
        plot.background = element_rect(fill='transparent', color='transparent'), #transparent plot bg

) + labs(x='eQTL effect size, CMS-specific', y='-log FDR', size='sv n')

sv.eqtl.plot

ggsave(paste0(basedir, "/results/fig/FIGS2b.sv.eqtl.svg"), plot = 
         sv.eqtl.plot
, device = "svg",width = 100, height = 95 , units='mm' , dpi=72 )

sv.eqtl.cms %>% filter( gene %in% c('TERT','EGFR','CDK6','FGFR2','MDM2','TP53','DCDC2','RB1'))

```

## ecDNA
How common is extra-chromosomal DNA and how does it align with our eQTL?

```{r}

split_string <- function(input_string) {
  # Remove the square brackets and single quotes, then split by comma
  items <- strsplit(gsub("\\['|\\']", "", input_string), ",")[[1]]
  # Trim any leading or trailing whitespace from each item
  items <- trimws(items)
  items <- gsub("'", "", items)
  return(items)
}

rerun.pulls = F
local.run = T

if(rerun.pulls){
  
  sample_list <- read.table(file.path(basedir,'/source_data/sample.list.txt'), header=F)

  first.loop = T
  first.ecDNA = T
  for(tumour_id in sample_list$V2){
    
    cat('\n\nprocessing ',tumour_id,'\n')
    donor_id <- sample_list$V1[sample_list$V2 == tumour_id]
    
    #cluster here will be a mush of ("population/tumour-specific") hotspots
    #and complex rearrangements
    #should I stratify clusters for VAF?
    this.sv.cnv.path <- paste0(hpcdir,tumour_id,'/sv.strict/',tumour_id,'.sv.cnv.txt')
    
    #amplicon-architect, by definition, filters for CNV clusters (first) 
    #so SV-only clusters won't show up
    this.sample.amplicons.path <- paste0(hpcdir
                                         ,tumour_id,'/amplicon/results/',tumour_id,'_classification/',
                                         tumour_id, '_result_table.tsv')
    
    if(file.exists(this.sv.cnv.path) & file.exists(this.sample.amplicons.path)){
      
      this.sv.cnv <- fread(this.sv.cnv.path)
      
      this.sample.amplicons <- fread(this.sample.amplicons.path)
      this.sample.amplicons <- this.sample.amplicons[!is.na(this.sample.amplicons$`AA amplicon number`),]
      
      if(nrow(this.sample.amplicons) > 0){
        
        for(this.amplicon.id in this.sample.amplicons$`Feature ID`){
          
          this.bed.file <- this.sample.amplicons$`Feature BED file`[this.sample.amplicons$`Feature ID` == this.amplicon.id]
          
          #if R is running on cluster, set to False
          if(local.run){this.bed.file <- gsub('/.mounts/labs/PCSI/','/Volumes/pcsi/',this.bed.file)}
          
          #TODO: this was a patch for a change in directory structure (25/06/20), 
          #can eventually be deleted
          if(!file.exists(this.bed.file)){this.bed.file <- gsub('/amplicon/','/amplicon/results/',this.bed.file)}
          
          ##read bed file
          this.bed <- fread(this.bed.file)
          
          ##check how often AA calls overlap with SV calls
          for(this.chrom in unique(this.bed$V1)){ 
            
            sub.bed <- this.bed[this.bed$V1 == this.chrom,]
            this.start <- sub.bed$V2[1] - 200000
            this.end <- sub.bed$V3[nrow(sub.bed)] + 200000
            
            this.caught.in.cluster <- this.sv.cnv[this.sv.cnv$CHROM == this.chrom & 
                                                    this.sv.cnv$POS >= this.start & 
                                                    this.sv.cnv$POS <= this.end]
            
            this.caught.in.cluster$amplicon.id <- this.sample.amplicons$`AA amplicon number`[this.sample.amplicons$`Feature ID` == this.amplicon.id]
            this.caught.in.cluster$amplicon.type <- this.sample.amplicons$Classification[this.sample.amplicons$`Feature ID` == this.amplicon.id]
            
            if(nrow(this.caught.in.cluster) == 0){
              cat('no SVs at all in ',this.amplicon.id,'\n')
            }
            
            if(first.loop){caught.in.cluster <- this.caught.in.cluster; first.loop = F} else {
              caught.in.cluster <- rbind.data.frame(caught.in.cluster, this.caught.in.cluster)
            }
          }
          
          ## pull ecDNA out seperatly 
          if(this.sample.amplicons$Classification[this.sample.amplicons$`Feature ID` == this.amplicon.id] == 'ecDNA'){
            
            cat('ecDNA found!\n')
            thi.gene_list <- split_string(this.sample.amplicons$`All genes`[this.sample.amplicons$`Feature ID` == this.amplicon.id])
            this.ecdna.genes <- cbind.data.frame(tumour_id, this.amplicon.id, thi.gene_list )
            
            if(first.ecDNA){ecdna.genes <- this.ecdna.genes; first.ecDNA = F} else {
              ecdna.genes <- rbind.data.frame(ecdna.genes, this.ecdna.genes)
            }
          }
        }
      } else { cat('no amplicons')}
    } else { cat('file missing')}
  }
  

  write.table(
    ecdna.genes,
    file = file.path(basedir,'/results/ecdna.genes.txt'),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
  
}

ecdna.genes <- fread(file.path(basedir,'/results/ecdna.genes.txt'))

ecdna.genes.tally <- ecdna.genes %>% group_by(tumour_id, thi.gene_list) %>% tally()
ecdna.genes.tally <- ecdna.genes.tally %>% group_by( thi.gene_list) %>% tally()
ecdna.genes.tally <- ecdna.genes.tally %>% filter(thi.gene_list != '[]' )

gene_set <- fread(file.path(basedir,'/source_data/fgs.txt'))
ecdna.genes.tally <- left_join(ecdna.genes.tally, gene_set, by=c('thi.gene_list'='gene_name'))

ggplot(ecdna.genes.tally %>% filter(!is.na(Chr)), aes(x=Start, y=n)) +
  geom_point()+
  geom_line()+
  facet_grid(.~Chr)+
  theme_bw() + theme(panel.grid = element_blank(),
                     axis.text.x = element_blank(),
                     
                     strip.background = element_blank())

ecdna.genes.tally %>% arrange(desc(n)) %>% filter(Chr == 12)
ecdna.genes.tally %>% arrange(desc(n)) %>% filter(Chr == 7)

```

## run fishhook
Run recurrence analysis and combine with eQTL results. 

You may notice RHD comes up as the second hit after FGFR2. This is statistically accurate but upon manual review of some of the cases with this variant, we found it was just a common germline variant, aka the one defining blood groups. This also felt like a good, although unexpected, proof-of-concept (although suggests some leakage of germline variants into our somatic sv calls).


```{r}

run_fishhook = F

if(run_fishhook){
  
  ## careful, the fishHook package over-writes a bunch of base functions 
  ## and you basically have to restart your environment to run anything else
  library(fishHook)
  packageVersion('fishHook')
  
  dndscv_reference_data_path <- file.path(secdir,'/code/src/dndscv_data/data/RefCDS_human_GRCh38_GencodeV18_recommended.rda')
  load(file=dndscv_reference_data_path)
  
  genes <- gr_genes
  current_seqlevels <- seqlevels(genes)
  new_seqlevels <- paste0('chr', current_seqlevels)
  seqlevels(genes) <- new_seqlevels
  
  ## Reading in mutation calls from WGS
  
  raw.full.sv.cnv <- fread(file.path(secdir,'~/Documents/scripts/github/evolution/results/sv.cnv.txt'))
  
  for(this.cms in c('all')){
    

    full.sv.cnv <- raw.full.sv.cnv 

    mutations.wgs <- GRanges(
      seqnames = full.sv.cnv$CHROM,
      ranges = IRanges(start = full.sv.cnv$POS, end = full.sv.cnv$POS),
      Variant_Type = full.sv.cnv$SV,
      uuid = full.sv.cnv$tumour
    )
    
    genome(mutations.wgs) <- "hg38"
    seqlengths(mutations.wgs) <- seqlengths(BSgenome.Hsapiens.UCSC.hg38)[seqlevels(mutations.wgs)]
    
    tiles = gr.tile(seqinfo(mutations.wgs), WINDOW_SIZE)
    
    # %+% is a gUtils operator that shifts ech GRange by given distance
    # Here we create additional GRanges by shifting all tiles by 5000 Kb
    supertiles = c(tiles, tiles %+% 5000)
    
    ## Annotating by nearest gene
    supertiles$nearest.gene = genes$names[nearest(supertiles, gr.stripstrand(genes))]
  
    
    ## Supertiles
    fish.supertiles.snv = Fish(hypotheses = supertiles, 
                               events = mutations.wgs,  
                               idcol = 'uuid', 
                               #use_local_mut_density = TRUE, 
                               mc.cores = 40)
    
    
    
    fish.supertiles.snv = fish.supertiles.snv[fish.supertiles.snv$data$frac.eligible > 0.75, ]
    fish.supertiles.snv$score()
    
    p1 = fish.supertiles.snv$qqp(plotly = FALSE)
    fish.res <- gr2dt(fish.supertiles.snv$res)[order(p), ]
    
    fish.res.gene <- fish.res %>%
      group_by(nearest.gene) %>%
      filter(fdr == min(fdr) & p == min(p)) %>% 
      dplyr::select(nearest.gene, p   , fdr ,count, effectsize, count.pred)
    
    fish.res.gene <- unique(fish.res.gene)
    fish.res.gene$cms <- this.cms
    
    if(this.cms == 'all'){all.fish.res.gene <- fish.res.gene} else {
      all.fish.res.gene <- rbind.data.frame(all.fish.res.gene, fish.res.gene)
    }

  }
  
  write.table(
    all.fish.res.gene,
    file = file.path(basedir, "/results/fish.res.gene.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
}

all.fish.res.gene <- fread(file.path(basedir, "/results/fish.res.gene.txt"))

sv.fishhook <- left_join(sv.eqtl, all.fish.res.gene, by=c('gene'='nearest.gene','cms'='cms'))

ggplot(sv.fishhook %>% filter(cms == 'all'), aes(x=-log(eqtl.fdr),-log(fdr))) + 
  geom_point()+
  geom_point(aes(color=cms),  data = ~ subset(., (eqtl.fdr < 0.05 & fdr < 0.05)  )) +
  geom_label_repel(aes(label=gene),    data = ~ subset(., (eqtl.fdr < 0.05 & fdr < 0.05)  ), size=5) +
  theme_bw(base_size=22) + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill='transparent'), #transparent panel bg
    plot.background = element_rect(fill='transparent', color='transparent') #transparent plot bg
  ) + 
  
  
  scale_color_manual(values = c(
    "all" = 'grey',
    "CMS-B"    = "#EDA71A" , 
    "CMS-A" = "#04597C" 
  )) 



```