# Test co-occurence of drivers
Now that we have our drivers, we ask how often do they occur together and are some mutually exclusive (and of course does this connect to CCS)? 

## Data prep
You may notice in data prep that there are only 168 (not 169) samples in this section: one case has no statistically-supported driver so there is not co-occurence to test.

```{r}
suppressWarnings(
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(pvclust)
  library(cluster)
  library(Rediscover)
  library(qvalue)
  library(PNWColors)
  library(ggplot2)
  library(cowplot)
  library(ggrepel)
  library(ggdendroplot)
  library(survival)
  library(survminer)
}))

packageVersion('Rediscover')

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'

source( file.path(code.dir, "bin/BTC.functions.R"))
source( file.path(code.dir, "bin/integrated_driver.functions.R"))

cols_names = c('donor', 'gene', 'type')

ssm_drivers <- fread(file.path(data.dir, "/results/TableS3.positioned_drivers_table.txt")) 
dnds_annotation <- fread(file.path(data.dir, "/results/dnds_annotation.txt"))
length(unique(dnds_annotation$sampleID))
gene.ssm <- dnds_annotation %>% dplyr::select(sampleID, gene, impact)
ssm.driver.hits <- gene.ssm %>% filter(gene %in% c(ssm_drivers$id, 'FGFR2','CDKN2A.p16INK4a', 'CDKN2A.p14arf'))
names(ssm.driver.hits) <- cols_names

all_mutations <- fread(paste0(data.dir, "/results/all_mutations.txt"))
tert_mutations <- all_mutations %>% filter(chr == 'chr5' & pos1 %in% c(1295135, 1295113))
tert_mutations$gene <- 'TERT'
tert_mutations$type <- 'promoter'
tert_mutations <- tert_mutations %>% dplyr::select(patient, gene, type)
names(tert_mutations) <- cols_names

ssm.driver.hits <- rbind.data.frame(ssm.driver.hits, tert_mutations)

full.sv.cnv <- fread(file.path(data.dir, '/results/sv.cnv.txt'))
gene_set <- fread(file.path(data.dir, '/source_data/fgs.txt'))
gene_set$Chr <- paste0('chr', gene_set$Chr)

driver.sv <- 
  rbind.data.frame(
    get.sv.window('TERT', gene_set, 5000, full.sv.cnv),
    get.sv.window('FGFR2', gene_set, 5000, full.sv.cnv),
    get.sv.window('MDM2', gene_set, 5000, full.sv.cnv),
    get.sv.window('CDK6', gene_set, 5000, full.sv.cnv),
    get.sv.window('SMAD4', gene_set, 5000, full.sv.cnv),
    get.sv.window('CDKN2A', gene_set, 5000, full.sv.cnv)
  )

gene.sv <- unique(driver.sv %>% dplyr::select(tumour, GENE, SV))
gene.sv %>% group_by(GENE) %>% tally()
sv.driver.hits <- gene.sv %>% filter(GENE %in% c('TERT','FGFR2','SMAD4','CDKN2A'))
names(sv.driver.hits) <- cols_names

full.mutations <- rbind.data.frame(sv.driver.hits, ssm.driver.hits) #cnv.driver.hits, btc.fusions, 

## a little filtering
full.mutations <-full.mutations %>% filter( type %ni% c('Synonymous'))

full.mutations$gene[full.mutations$gene %in% c('CDKN2A.p16INK4a', 'CDKN2A.p14arf','CDKN2B-AS1')] <- 'CDKN2A'

length(unique(dnds_annotation$sampleID))

# incl. donors without actions in the onco plot
no_variants <- dnds_annotation[dnds_annotation$sampleID %ni% unique(full.mutations$donor) ,]
names(no_variants)[names(no_variants) == "sampleID"] <- "donor"
no_variants$gene <-NA
no_variants$type <- NA
no_variants <- unique(no_variants[,c("donor","gene","type")])

full.mutations <- rbind.data.frame(full.mutations, no_variants)


full.mutations$type <- full.mutations$type
full.mutations$type[full.mutations$type %in% c('DEL','Deletion')] <- 'Deletion'
full.mutations$type[full.mutations$type %in% c('DUP','Amplification')] <- 'Amplification'
full.mutations$type[full.mutations$type %in% c('INV','TRA', 'Fusion')] <- 'Translocation/Fusion'
full.mutations$type[full.mutations$type %in% c('no-SNV','Missense','Nonsense','Essential_Splice','promoter')] <- 'Small Substitution'
full.mutations$type[full.mutations$type %in% c('CLU','BND')] <- 'Complex SV'
full.mutations <- unique(full.mutations)

```

## Visualize occurences using oncoplot

```{r}

driver.mutations <- make_oncoplot_factor_order(full.mutations)
driver.mutations <- unique(driver.mutations)

key_frequency <- get_gene_frequencies(driver.mutations)

fz_plot <- 
  ggplot(key_frequency , aes(x=frequency*100, y=gene_factor)) + 
  geom_bar(stat = "identity")+
  scale_y_discrete(limits=rev) +
  theme_bw(base_size=10)+ 
  labs(x="Cohort Frequency (%)",fill="OncoKb Level") + 
  guides(fill = guide_legend(title.position="top")) +
  #scale_fill_grey()+
  this_theme + 
  theme(
    legend.position = "top" ,
    plot.margin = unit(c(0,1,0,-1), "lines"),
    axis.text.y =element_blank(),
    axis.title.y =element_blank(),
    axis.ticks.y =element_blank(),
  ) 

onco_plot <- 
  ggplot(driver.mutations, aes(x=donor_factor, y=gene_factor)) + 
  geom_tile(aes(fill=type)) + 
    theme_bw(base_size=10) +
  scale_y_discrete(limits=rev, breaks=na.omit(unique(driver.mutations$gene_factor))) +
  labs(x='',y='' )+ 
  
  scale_fill_manual(values=pnw_palette(name="Lake",n=8)[c(2:7)], na.value = 'white') +
    this_theme + 
    theme(legend.position = "bottom" , 
       #   axis.text.y = element_text(size = 10),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()
          
          )


width_gap = -0.09

this.plot <- plot_grid(onco_plot,NULL, fz_plot,  rel_widths = c(0.8,width_gap,0.2 ), ncol = 3, align = 'hv', axis = "rlbt")
this.plot
ggsave( paste0(data.dir, "/results/fig/figs4c.current_oncoplot.svg"), 
        plot =  this.plot , 
        device = "svg", width = 200, height = 80 , units='mm' , dpi=72)

write.table(
  driver.mutations %>% dplyr::select(donor,gene,type),
  file = paste0(data.dir, "/results/driver.landscape.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

```

## associate driver mutations with eCCS
How likely are drivers to be mutated in cases from each subtype?

```{r}

predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt') )

full.mutations.classed <- full_join(full.mutations, predicted.w.score, by=c('donor'='Sample')) %>% 
  filter( !is.na(rna_class) & !is.na(gene))

full.mutations.classed <- dcast(full.mutations.classed, formula = 'donor + rna_class ~ gene', fun.aggregate=length)

odds.ratio.df <- get.odds.ratio(
  this.col = c('ARID1A'), 
  odds.ratio = F, 
  muts.clinic=full.mutations.classed
)

for(this.gene in names(full.mutations.classed)[-c(1:3)]){
  
  odds.ratio.df <- get.odds.ratio(
    this.col = this.gene, 
    odds.ratio = odds.ratio.df, 
    muts.clinic=full.mutations.classed
  )
  
}

odds.ratio.df$sig <- NA
odds.ratio.df$sig[odds.ratio.df$p.value < 0.05 & odds.ratio.df$lor < 0] <- 'CCS-B'
odds.ratio.df$sig[odds.ratio.df$p.value < 0.05 & odds.ratio.df$lor > 0] <- 'CCS-A'

write.table(
  odds.ratio.df,
  file = paste0(data.dir, "/results/driver.odds.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

order_vec <- c(na.omit(odds.ratio.df$var[order(odds.ratio.df$lor)]))
order_vec <- order_vec[order_vec != 'NA']

```



## associate co-occurence with eCCS
We use the package Rediscovery here to ensure that the significance of co-occurence is corrected for high occurence

```{r}

mutation.dcast <- reshape2::dcast(full.mutations, donor ~ gene, fun.aggregate=length)

full.mutations.bin <- full.mutations %>% filter(!is.na(gene))
full.mutations.bin$type <- 1
full.mutations.bin.mat <- as.matrix(dcast(data = full.mutations.bin, formula = donor ~  gene, fun.aggregate=length))
row.names(full.mutations.bin.mat) <- full.mutations.bin.mat[,1]
full.mutations.bin.mat <- full.mutations.bin.mat[,-1]
full.mutations.bin.mat[full.mutations.bin.mat > 1] <- 1
full.mutations.bin.mat <- t(full.mutations.bin.mat)
full.mutations.bin.mat <- apply(full.mutations.bin.mat, c(1, 2), as.numeric)

mutation.prob <- getPM(full.mutations.bin.mat)
mutex.res <- getMutex(full.mutations.bin.mat, mutation.prob)
mutex.res <- as.matrix(mutex.res)

mutex.res_qvalue <- getMutex(full.mutations.bin.mat, mutation.prob)
mutex.res_qvalue@x <- qvalue::qvalue(mutex.res_qvalue@x)$qvalue

colnames(mutex.res) = rownames(mutex.res) = row.names(full.mutations.bin.mat)

## rearrange matrix for plot
## you have to reorder the mutex matrix before zeroing the upper.tri
mutex.res <- mutex.res[order_vec, order_vec]

mutex.res[upper.tri(mutex.res, diag = TRUE)] <- NA

mutex.res.df <- data.frame(mutex.res )
mutex.res.df$gene <- order_vec
mutex.res.melt <- reshape2::melt(mutex.res.df, id.vars='gene')

mutex.res.melt$value.adj <- mutex.res.melt$value - 0.5
mutex.res.melt$value.adj.abs <- abs(mutex.res.melt$value.adj)
mutex.res.melt$value.adj.abs.sig <- mutex.res.melt$value.adj.abs
mutex.res.melt$value.adj.abs.sig[mutex.res.melt$value.adj.abs < 0.45] <- NA

mutex.res.melt$gene <- factor(mutex.res.melt$gene, levels = order_vec)
mutex.res.melt$gene2 <- factor(mutex.res.melt$variable, levels = order_vec)

mutex.df <- as.data.frame(as.matrix(mutex.res))
names(mutex.df) <- row.names(full.mutations.bin.mat)

mutex.df$gene <- names(mutex.df)
mutex.melt <- reshape2::melt(mutex.df, id.vars='gene')

mutex.melt.na <- mutex.melt
mutex.melt.na$value[mutex.melt.na$value < 0.9 & mutex.melt.na$value > 0.1] <- NA
mutex.melt.na$value[mutex.melt.na$value < 0.95 & mutex.melt.na$value > 0.05] <- NA
mutex.melt.na$value[mutex.melt.na$value == 0] <- NA

mutex.res.melt$combined <- paste(pmin(as.character(mutex.res.melt$gene), as.character(mutex.res.melt$variable)), 
                                 pmax(as.character(mutex.res.melt$gene), as.character(mutex.res.melt$variable)), sep = ", ")

sig.pairs <- mutex.res.melt %>% filter(value < 0.05 | value > 0.95 )
strict.sig.pairs <- mutex.res.melt %>% filter(value < 0.01 | value > 0.99  )

odds.ratio.df$gene <- factor(odds.ratio.df$var, levels = order_vec)
mutex.res.melt$gene2 <- factor(mutex.res.melt$gene2, levels = order_vec)

subtype.driver.LOR <- 
  plot_grid(
    
    ggplot(mutex.res.melt , aes(x=gene, y=gene2) )+ 
  
      geom_point(aes(size=value.adj.abs,alpha=value.adj.abs, color=value.adj), shape=16) +        
      geom_point(aes(size=value.adj.abs.sig), shape =1, stroke = 1) + 
      scale_color_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
      theme_bw(base_size=10)+  guides(color='none',size='none',alpha='none')+
      theme( 
  
        legend.position = 'bottom',
        panel.border = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = -45,  hjust=0.01),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank() ,
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
        axis.text.y =element_blank(),
        panel.background = element_blank(), 
        plot.background = element_blank() 
      ) +
     coord_fixed()+
      scale_size_continuous(range = c(0.5, 3))  
      
     
    ,
    NULL,
    
    ggplot(odds.ratio.df, aes(x=lor, y=gene)) +
      geom_vline(xintercept = 0, color='grey80')+
      geom_errorbar( aes(xmin =  ci.lo, xmax = ci.hi), width=0) +
      geom_point(aes(color=sig), size=2)+ 
      theme_bw(base_size=10) +
      scale_size_continuous(range = c(0.5, 3))  +
      
      scale_color_manual(values = pnw_palette(name="Bay", n=30)[c(3,20)], na.value = 'grey90') +
      
      theme(panel.grid = element_blank(), 
            panel.border = element_blank(),
            panel.spacing = unit(1, "lines"),
            strip.background = element_blank(),
            axis.ticks.y = element_blank(),
            axis.text.y = element_text(hjust = 0.5),
            plot.margin = unit(c(0, 0, 0, 0), "mm"),
            axis.line.x = element_line(color = "grey80"),
            legend.position = 'bottom',
            panel.background = element_blank(), 
            plot.background = element_blank() 

      ) +
      guides(color='none',size='none')+
      labs(x='CCS-A log odds ratio',y='')+
      scale_y_discrete(position = "left")
  
  , rel_widths = c(0.25,-0.01,0.2 ), ncol = 3,align = 'h', axis = "rlbt")

subtype.driver.LOR

ggsave(file.path(data.dir,"/results/fig/FIG4BC.subtype.driver.LOR.svg"), plot = 
  subtype.driver.LOR
, device = "svg", height = 75, width = 115, units='mm', dpi=72 )
 
```
## TP53 and survival

```{r}
predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt')) %>%
   dplyr::select(Sample, rna_class, confidence.polarized)
predicted.w.score$rna_class[is.na(predicted.w.score$rna_class)] <- 'Other'
predicted.w.score$final.cms <- factor(predicted.w.score$rna_class, levels = rev(c("CCS-A" ,'Other',"CCS-B")))

all.samples <- fread(file.path(data.dir,"/results/samples.info.txt")) %>% 
  dplyr::select(Study_ID, Sample, location, FF, Site_of_tissue, Sample_Progression, tissue_before_chemo)

all.samples <- left_join(all.samples, predicted.w.score)

timeline.outcome <- fread(file.path(data.dir,"/results/timeline.txt"))

## adjust timeline
timeline.outcome <- timeline.outcome %>% 
  filter(relative_latest_date > 0)

timeline.outcome$last.from.surgery = timeline.outcome$relative_latest_date - timeline.outcome$relative_Date_of_Surgery
timeline.outcome$last.from.L1 = timeline.outcome$relative_latest_date - timeline.outcome$relative_Date_of_first_infusion_of_first_line_palliative_chemotherapy

all.clinic <- fread(file.path(data.dir,"/results/clinic.info.txt")) %>% 
  dplyr::select(Study_ID, Status, Overall_Stage_at_Diagnosis, Specify_other_cause_of_death,
                Sex_assigned_at_birth, Age_at_first_histological_diagnosis
                )

na_values <- c("No information", "Not available", "Not Done", 
               "Unknown", "Not stated", "Not applicable", "","N/A")

all.clinic <- all.clinic %>%
  mutate(across(Overall_Stage_at_Diagnosis, ~replace(., . %in% na_values, NA)))

all.clinic$OS_STATUS <- NA
all.clinic$OS_STATUS[all.clinic$Status == 'Alive'] <- 0
all.clinic$OS_STATUS[all.clinic$Status == 'Expired'] <- 1
all.clinic %>% group_by(Status, OS_STATUS) %>% tally()

timeline.outcome <- full_join(timeline.outcome, all.clinic, by=c('Study_ID'='Study_ID'))  %>% 
  filter(Specify_other_cause_of_death %ni% c('coagulopathy and MSOF post hepatectomy') )

alluvial <- fread(file.path(data.dir,"/results/alluvial.txt")) %>% 
  dplyr::select(Study_ID, Type_of_Surgery, first_palliative, first_palliative_specific, second_palliative)

timeline.outcome <- full_join(timeline.outcome, alluvial, by=c('Study_ID'='Study_ID'))

timeline.outcome <- left_join(all.samples, timeline.outcome, by=c('Study_ID'='Study_ID'))  %>%
  filter( rna_class %in% c('CCS-A','CCS-B'))

table(timeline.outcome$rna_class)

tp53.hit <- c(na.omit(full.mutations$donor[full.mutations$gene == 'TP53']))

palliative.cohort <- timeline.outcome %>% 
  filter(  last.from.L1 > 0 & rna_class != 'Other' & Sample %ni% tp53.hit &
             first_palliative %in% c( "Other Drug", "Chemotherapy + Immunotherapy","Chemotherapy" ) 
             ) 

palliative.cohort$OS_STATUS[palliative.cohort$last.from.L1 > 12 *5] <- 0
palliative.cohort$last.from.L1[palliative.cohort$last.from.L1 > 12 *5] <- 12 *5

os.palliative.cms <- survfit(Surv(last.from.L1, OS_STATUS) ~ rna_class , 
                             data = palliative.cohort )

os.palliative.cms
surv_pvalue(os.palliative.cms, data = palliative.cohort)

```
