# Test co-occurence of drivers
Now that we have our drivers, we ask how often do they occur together and are some mutually exclusive (and of course does this connect to CCS)? 

## Data prep
You may notice in data prep that there are only 168 (not 169) samples in this section: one case has no statistically-supported driver so there is not co-occurence to test.

```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(pvclust)
  library(cluster)
  library(Rediscover)
  library(qvalue)
  library(PNWColors)
  library(ggplot2)
  library(cowplot)
  library(ggrepel)
  library(ggdendroplot)
})

packageVersion('Rediscover')

basedir = '~/Documents/scripts/github/btc-ccs/'
secdir = '~/Documents/scripts/github/evolution/'

source( file.path(basedir, "bin/BTC.functions.R"))
source( file.path(basedir, "bin/integrated_driver.functions.R"))

cols_names = c('donor', 'gene', 'type')

ssm_drivers <- fread(file.path(basedir, "/results/TableS3.positioned_drivers_table.txt")) 
dnds_annotation <- fread(file.path(basedir, "/results/dnds_annotation.txt"))
length(unique(dnds_annotation$sampleID))
gene.ssm <- dnds_annotation %>% dplyr::select(sampleID, gene, impact)
ssm.driver.hits <- gene.ssm %>% filter(gene %in% c(ssm_drivers$id, 'FGFR2','CDKN2A.p16INK4a', 'CDKN2A.p14arf'))
names(ssm.driver.hits) <- cols_names

all_mutations <- fread(paste0(secdir, "/results/all_mutations.txt"))
tert_mutations <- all_mutations %>% filter(chr == 'chr5' & pos1 %in% c(1295135, 1295113))
tert_mutations$gene <- 'TERT'
tert_mutations$type <- 'promoter'
tert_mutations <- tert_mutations %>% dplyr::select(patient, gene, type)
names(tert_mutations) <- cols_names

ssm.driver.hits <- rbind.data.frame(ssm.driver.hits, tert_mutations)

full.sv.cnv <- fread(file.path(secdir, '/results/sv.cnv.txt'))
gene_set <- fread(file.path(basedir, '/source_data/fgs.txt'))
gene_set$Chr <- paste0('chr', gene_set$Chr)

driver.sv <- 
  rbind.data.frame(
    get.sv.window('TERT', gene_set, 5000, full.sv.cnv),
    get.sv.window('FGFR2', gene_set, 5000, full.sv.cnv),
    get.sv.window('SMAD4', gene_set, 5000, full.sv.cnv),
    get.sv.window('CDKN2A', gene_set, 5000, full.sv.cnv)
  )

gene.sv <- unique(driver.sv %>% dplyr::select(tumour, GENE, SV))
sv.driver.hits <- gene.sv 
names(sv.driver.hits) <- cols_names

full.mutations <- rbind.data.frame(sv.driver.hits, ssm.driver.hits) #cnv.driver.hits, btc.fusions, 

## a little filtering
full.mutations <-full.mutations %>% filter( type %ni% c('Synonymous'))

full.mutations$gene[full.mutations$gene %in% c('CDKN2A.p16INK4a', 'CDKN2A.p14arf','CDKN2B-AS1')] <- 'CDKN2A'

length(unique(dnds_annotation$sampleID))

# incl. donors without actions in the onco plot
no_variants <- dnds_annotation[dnds_annotation$sampleID %ni% unique(full.mutations$donor) ,]
names(no_variants)[names(no_variants) == "sampleID"] <- "donor"
no_variants$gene <-NA
no_variants$type <- NA
no_variants <- unique(no_variants[,c("donor","gene","type")])

full.mutations <- rbind.data.frame(full.mutations, no_variants)


full.mutations$type <- full.mutations$type
full.mutations$type[full.mutations$type %in% c('DEL','Deletion')] <- 'Deletion'
full.mutations$type[full.mutations$type %in% c('DUP','Amplification')] <- 'Amplification'
full.mutations$type[full.mutations$type %in% c('INV','TRA', 'Fusion')] <- 'Translocation/Fusion'
full.mutations$type[full.mutations$type %in% c('no-SNV','Missense','Nonsense','Essential_Splice')] <- 'Small Substitution'
full.mutations$type[full.mutations$type %in% c('CLU','BND')] <- 'Complex SV'
full.mutations <- unique(full.mutations)

```

## Visualize occurences using oncoplot

```{r}

driver.mutations <- make_oncoplot_factor_order(full.mutations)
driver.mutations <- unique(driver.mutations)

key_frequency <- get_gene_frequencies(driver.mutations)

this_theme <- theme(        
  panel.background = element_rect(fill='transparent'), #transparent panel bg
  plot.background = element_rect(fill='transparent', color='transparent'), #transparent plot bg
  panel.grid.major.x = element_blank(), #remove major gridlines
  panel.grid.minor = element_blank(), #remove minor gridlines
  legend.background = element_rect(color = NA,fill=NA), #transparent legend bg
  legend.box.background = element_rect(color = NA,fill=NA), #transparent legend panel
  
  plot.title = element_blank(),
  legend.spacing = unit(0, 'cm'),)+   
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    
    panel.grid = element_blank(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    
    legend.key.size = unit(0.4, "cm"),   # Shrinks the legend keys (symbols)
    legend.text = element_text(size = 8), # Shrinks the text for legend items
    legend.title = element_text(size = 9) # Shrinks the legend title
    
    
  )  

fz_plot <- 
  ggplot(key_frequency , aes(x=frequency*100, y=gene_factor)) + 
  geom_bar(stat = "identity")+
  scale_y_discrete(limits=rev) +
  theme_bw(base_size=10)+ 
  labs(x="Cohort Frequency (%)",fill="OncoKb Level") + 
  guides(fill = guide_legend(title.position="top")) +
  #scale_fill_grey()+
  this_theme + 
  theme(
    legend.position = "top" ,
    plot.margin = unit(c(0,1,0,-1), "lines"),
    axis.text.y =element_blank(),
    axis.title.y =element_blank(),
    axis.ticks.y =element_blank(),
  ) 

onco_plot <- 
  ggplot(driver.mutations, aes(x=donor_factor, y=gene_factor)) + 
  geom_tile(aes(fill=type)) + 
    theme_bw(base_size=10) +
  scale_y_discrete(limits=rev, breaks=na.omit(unique(driver.mutations$gene_factor))) +
  labs(x='',y='' )+ 
  
  scale_fill_manual(values=pnw_palette(name="Lake",n=8)[c(2:7)], na.value = 'white') +
    this_theme + 
    theme(legend.position = "bottom" , 
       #   axis.text.y = element_text(size = 10),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()
          
          )


width_gap = -0.09

this.plot <- plot_grid(onco_plot,NULL, fz_plot,  rel_widths = c(0.8,width_gap,0.2 ), ncol = 3, align = 'hv', axis = "rlbt")
this.plot
ggsave( paste0(basedir, "/results/fig/figs4c.current_oncoplot.svg"), 
        plot =  this.plot , 
        device = "svg", width = 200, height = 80 , units='mm' , dpi=72)

write.table(
  driver.mutations %>% dplyr::select(donor,gene,type),
  file = paste0(basedir, "/results/driver.landscape.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

```

## associate driver mutations with eCCS
How likely are drivers to be mutated in cases from each subtype?

```{r}

get.odds.ratio <- function(odds.ratio,  this.col, muts.clinic, var.string=NULL, pos.var=1, neg.var=0, subclass.factors = c("CMS-A", "CMS-B") ){
  
  this.df <- tibble(muts.clinic) %>% dplyr::select(rna_class, all_of(this.col)) 
  names(this.df)[2] <- 'var'
  this.df$var[this.df$var > 1] <- 1
  this.df.filt <- this.df %>% filter(var %in% c(pos.var, neg.var)) 
  this.df.filt$var <- factor(this.df.filt$var, levels=c(pos.var, neg.var))
  this.df.filt$rna_class <- factor(this.df.filt$rna_class, levels=subclass.factors)
  
  this.table <- table(this.df.filt)
  
  this.odds_ratio <- fisher.test(t(this.table), conf.int = T)
  
  if(is.null(var.string)){
    var.string = this.col
  }
  
  res.tmp <- cbind.data.frame('var'=var.string, 
                              'lor'=log(this.odds_ratio$estimate), 
                              'ci.lo' = log(this.odds_ratio$conf.int[1]),
                              'ci.hi' = log(this.odds_ratio$conf.int[2]), 
                              'p.value' = this.odds_ratio$p.value, 
                              'k'=sum(this.table[,1]),
                              'n'=sum(this.table)
                              )
  
  if(class(odds.ratio) == "logical"){
    odds.ratio <- res.tmp
  } else {
    odds.ratio <- rbind.data.frame(odds.ratio, res.tmp)
    
  }
  
  return(odds.ratio)
}

predicted.w.score <- fread(file.path(basedir,'/results/all.eCMS.txt') )
predicted.w.score <- predicted.w.score %>% filter(`Sample Source(FFPE, FF)` == 'FF')

full.mutations.classed <- full_join(full.mutations, predicted.w.score, by=c('donor'='sample')) %>% filter( !is.na(rna_class))

full.mutations.classed <- dcast(full.mutations.classed, formula = 'donor + rna_class ~ gene', fun.aggregate=length)

odds.ratio.df <- get.odds.ratio(
  this.col = c('ARID1A'), 
  odds.ratio = F, 
  muts.clinic=full.mutations.classed
)

for(this.gene in names(full.mutations.classed)[-c(1:4)]){
  
  odds.ratio.df <- get.odds.ratio(
    this.col = this.gene, 
    odds.ratio = odds.ratio.df, 
    muts.clinic=full.mutations.classed
  )
  
}

odds.ratio.df$sig <- NA
odds.ratio.df$sig[odds.ratio.df$p.value < 0.05 & odds.ratio.df$lor < 0] <- 'CMS-B'
odds.ratio.df$sig[odds.ratio.df$p.value < 0.05 & odds.ratio.df$lor > 0] <- 'CMS-A'


write.table(
  odds.ratio.df,
  file = paste0(basedir, "/results/driver.odds.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

order_vec <- odds.ratio.df$var[order(odds.ratio.df$lor)]

```

## significance of co-variance

```{r}

sample_count <- length(unique(full.mutations$donor))
key_genes_var <- full.mutations %>% filter(!is.na(gene))
first=T
first_pair=T

for(this_gene in unique(key_genes_var$gene)){
  this_annotation <- key_genes_var[key_genes_var$gene == this_gene,]
  this_sample_fz <- length(unique(this_annotation$sampleID)) / sample_count
  
  if(first){this_driver_table = cbind.data.frame('gene'=this_gene, 'fz'=this_sample_fz); first=F}else{
    this_driver_table = rbind.data.frame(this_driver_table, cbind.data.frame('gene'=this_gene, 'fz'=this_sample_fz))
  }
  
  for(this_paired_gene in unique(key_genes_var$gene)){
    these_samples_have_it <- unique(this_annotation$donor)
    these_samples_have_the_other <- unique(key_genes_var$donor[key_genes_var$gene == this_paired_gene])
    
    co_occurs <- length(unique(key_genes_var$donor[key_genes_var$gene == this_paired_gene & key_genes_var$donor %in% these_samples_have_it]))
    
    
    
    contingency_table <- data.frame('this_gene_occurs' = c(co_occurs,
                                                           length(unique(key_genes_var$donor[key_genes_var$gene == this_gene & key_genes_var$donor %ni% these_samples_have_the_other]))), 
                                    'this_gene_wt' = c(length(unique(key_genes_var$donor[key_genes_var$gene == this_paired_gene & key_genes_var$donor %ni% these_samples_have_it])),
                                                       (sample_count - length(unique(key_genes_var$donor[key_genes_var$gene %in% c(this_paired_gene, this_gene)]))) ),
                                    row.names = c('paired_gene_occurs','paired_gene_wt'))
    fisher.res <- fisher.test(contingency_table)
    
    
    
    this_pair_table_tmp = cbind.data.frame('gene'=this_gene, 'pair'=this_paired_gene, 'count_match'=co_occurs,  'fz_match'=co_occurs/length(these_samples_have_it),'fz_pair'=co_occurs/sample_count,
                                           'odd.ratio' = fisher.res$estimate, 'p.value' = fisher.res$p.value, 'sorted.pair' = paste(sort(c(this_gene, this_paired_gene)), collapse = ', ') ) 
    
    if(first_pair){this_pair_table <- this_pair_table_tmp; first_pair=F}else{
      this_pair_table = rbind.data.frame(this_pair_table, this_pair_table_tmp)
    }
  }
}

paired_frequency <- dcast(unique(as.data.table(this_pair_table[,c(1,2,4)])), gene ~ pair, fill = NA, fun.aggregate=length)
paired_frequency_matrix <- as.matrix(paired_frequency[,-1])
rownames(paired_frequency_matrix) <- paired_frequency$gene

paired_frequency_matrix.melt <- reshape2::melt(paired_frequency_matrix)

or.table <- this_pair_table %>% filter(gene  != pair  ) %>% dplyr::select(sorted.pair, odd.ratio, p.value, fz_pair)
or.table <- or.table[order(or.table$sorted.pair),]
or.table <- or.table[seq(1,nrow(or.table),2),]

ggplot(or.table , aes(x=log(odd.ratio), y=-log(p.value))) + 
  geom_vline(xintercept = 0, color='red', linetype='dashed')+
  
  geom_point(aes( size=fz_pair),  color='grey80') +
  geom_point(aes( size=fz_pair), 
             data = ~ subset(., p.value < 0.01 )) +
  geom_label_repel(aes(label=sorted.pair), 
                   data = ~ subset(., p.value < 0.01 ), size=2, max.overlaps =30,
                   box.padding = 0.8, force = 0.8, max.time = 3, min.segment.length = 0) +
  guides(size='none')+
  theme_minimal(base_size=10)+ theme( 
    legend.position = 'bottom',
    panel.background = element_blank(),
    plot.background = element_blank(),
    
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank() #remove minor gridlines
  ) + labs(x='log Odds Ratio', y='p-value (-log)', size='frequency') +
  scale_x_continuous(expand = expansion(mult = 0.25), breaks = c(-2,0,2)) +
  scale_y_continuous(expand = expansion(mult = 0.05)) 

```

## associate co-occurence with eCCS
We use the package Rediscovery here to ensure that the significance of co-occurence is corrected for high occurence

```{r}

mutation.dcast <- reshape2::dcast(full.mutations, donor ~ gene, fun.aggregate=length)

full.mutations.bin <- full.mutations %>% filter(!is.na(gene))
full.mutations.bin$type <- 1
full.mutations.bin.mat <- as.matrix(dcast(data = full.mutations.bin, formula = donor ~  gene, fun.aggregate=length))
row.names(full.mutations.bin.mat) <- full.mutations.bin.mat[,1]
full.mutations.bin.mat <- full.mutations.bin.mat[,-1]
full.mutations.bin.mat[full.mutations.bin.mat > 1] <- 1
full.mutations.bin.mat <- t(full.mutations.bin.mat)
full.mutations.bin.mat <- apply(full.mutations.bin.mat, c(1, 2), as.numeric)

mutation.prob <- getPM(full.mutations.bin.mat)
mutex.res <- getMutex(full.mutations.bin.mat, mutation.prob)
mutex.res <- as.matrix(mutex.res)

mutex.res_qvalue <- getMutex(full.mutations.bin.mat, mutation.prob)
mutex.res_qvalue@x <- qvalue::qvalue(mutex.res_qvalue@x)$qvalue



colnames(mutex.res) = rownames(mutex.res) = row.names(full.mutations.bin.mat)

## rearrange matrix for plot
## you have to reorder the mutex matrix before zeroing the upper.tri
mutex.res <- mutex.res[order_vec, order_vec]

mutex.res[upper.tri(mutex.res, diag = TRUE)] <- NA

mutex.res.df <- data.frame(mutex.res )
mutex.res.df$gene <- order_vec
mutex.res.melt <- reshape2::melt(mutex.res.df, id.vars='gene')

mutex.res.melt$value.adj <- mutex.res.melt$value - 0.5
mutex.res.melt$value.adj.abs <- abs(mutex.res.melt$value.adj)
mutex.res.melt$value.adj.abs.sig <- mutex.res.melt$value.adj.abs
mutex.res.melt$value.adj.abs.sig[mutex.res.melt$value.adj.abs < 0.45] <- NA

mutex.res.melt$gene <- factor(mutex.res.melt$gene, levels = order_vec)
mutex.res.melt$gene2 <- factor(mutex.res.melt$variable, levels = order_vec)

mutex.df <- as.data.frame(as.matrix(mutex.res))
names(mutex.df) <- row.names(full.mutations.bin.mat)

mutex.df$gene <- names(mutex.df)
mutex.melt <- reshape2::melt(mutex.df, id.vars='gene')

mutex.melt.na <- mutex.melt
mutex.melt.na$value[mutex.melt.na$value < 0.9 & mutex.melt.na$value > 0.1] <- NA
mutex.melt.na$value[mutex.melt.na$value < 0.95 & mutex.melt.na$value > 0.05] <- NA
mutex.melt.na$value[mutex.melt.na$value == 0] <- NA

paired_frequency_matrix.join <- full_join(paired_frequency_matrix.melt, mutex.melt.na, by=c('Var1'='gene','Var2'='variable') )

paired_frequency_matrix.join$gene <- factor(paired_frequency_matrix.join$Var1, levels = order_vec)
paired_frequency_matrix.join$gene2 <- factor(paired_frequency_matrix.join$Var2, levels = rev(order_vec))

paired_frequency_matrix.join$value.y[paired_frequency_matrix.join$value.x == 0] <- NA

mutex.res.melt$combined <- paste(pmin(as.character(mutex.res.melt$gene), as.character(mutex.res.melt$variable)), 
                                 pmax(as.character(mutex.res.melt$gene), as.character(mutex.res.melt$variable)), sep = ", ")

sig.pairs <- mutex.res.melt %>% filter(value < 0.05 | value > 0.95 )
strict.sig.pairs <- mutex.res.melt %>% filter(value < 0.01 | value > 0.99  )

odds.ratio.df$gene <- factor(odds.ratio.df$var, levels = order_vec)
mutex.res.melt$gene2 <- factor(mutex.res.melt$gene2, levels = order_vec)

subtype.driver.LOR <- 
  plot_grid(
    
    ggplot(mutex.res.melt , aes(x=gene, y=gene2) )+ 
  
      geom_point(aes(size=value.adj.abs,alpha=value.adj.abs, color=value.adj), shape=16) +        
      geom_point(aes(size=value.adj.abs.sig), shape =1, stroke = 1) + 
      scale_color_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
      theme_bw(base_size=10)+  guides(color='none',size='none',alpha='none')+
      theme( 
  
        legend.position = 'bottom',
        panel.border = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = -45,  hjust=0.01),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank() ,
        plot.margin = unit(c(0, 0, 0, 0), "mm"),
        axis.text.y =element_blank(),
        panel.background = element_blank(), 
        plot.background = element_blank() 
      ) +
     coord_fixed()+
      scale_size_continuous(range = c(0.5, 3))  
      
     
    ,
    NULL,
    
    ggplot(odds.ratio.df, aes(x=lor, y=gene)) +
      geom_vline(xintercept = 0, color='grey80')+
      geom_errorbar( aes(xmin =  ci.lo, xmax = ci.hi), width=0) +
      geom_point(aes(color=sig))+ 
      theme_bw(base_size=10) +
      scale_size_continuous(range = c(0.5, 3))  +
      
      scale_color_manual(values = pnw_palette(name="Bay", n=30)[c(3,20)], na.value = 'grey90') +
      
      theme(panel.grid = element_blank(), 
            panel.border = element_blank(),
            panel.spacing = unit(1, "lines"),
            strip.background = element_blank(),
            axis.ticks.y = element_blank(),
            axis.text.y = element_text(hjust = 0.5),
            plot.margin = unit(c(0, 0, 0, 0), "mm"),
            axis.line.x = element_line(color = "grey80"),
            legend.position = 'bottom',
            panel.background = element_blank(), 
            plot.background = element_blank() 

      ) +
      guides(color='none',size='none')+
      labs(x='CMS-A log odds ratio',y='')+
      scale_y_discrete(position = "left")
  
  , rel_widths = c(0.25,-0.01,0.2 ), ncol = 3,align = 'h', axis = "rlbt")

subtype.driver.LOR

ggsave(file.path(basedir,"/results/fig/FIG4BC.subtype.driver.LOR.svg"), plot = 
  subtype.driver.LOR
, device = "svg", height = 75, width = 115, units='mm', dpi=72 )
 


```