# Compare mutational profiles between CCS
Let's compare the mutational profiles between the substypes

## Data Intake

```{r}
suppressWarnings(
suppressPackageStartupMessages({
  library(FactoMineR)
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(ggrepel)
  library(broom) #tidy
  library(logistf) #Firth's bias-reduced logistic regressions
  library(vegan) # shannon index
  library(stringr )
  library(ggupset)
  library(PNWColors)
  library(cowplot)
  library(car)
}))

'%ni%' <- function(x,y)!('%in%'(x,y))

this_theme <- theme(        
  panel.background = element_rect(fill='transparent'), #transparent panel bg
  plot.background = element_rect(fill='transparent', color='transparent'), #transparent plot bg
  panel.grid.major.x = element_blank(), #remove major gridlines
  panel.grid.minor = element_blank(), #remove minor gridlines
  legend.background = element_rect(color = NA,fill=NA), #transparent legend bg
  legend.box.background = element_rect(color = NA,fill=NA), #transparent legend panel
  
  axis.text.x =element_blank(),
  panel.grid = element_blank(),
  axis.title =element_blank(),
  axis.ticks.x =element_blank(),
  legend.title=element_blank(),
  plot.title = element_blank(),
  panel.border = element_blank(),
  legend.spacing = unit(0, 'cm'),)

code.dir = '~/Documents/scripts/github/btc-ccs/'
data.dir = '~/Documents/scripts/github/btc-ccs-data/'

sigprofiler.meta <- fread(file.path(data.dir,'/source_data/sigprofiler.meta.txt'))

cohort.summary <- fread(file.path(data.dir,'/results/summary.txt'))

full.sv.cnv <- fread(file.path(data.dir,'/results/sv.cnv.txt'))

mutation.landscape <- fread( paste0(data.dir, "/results/mutation.landscape.txt"))

HYPERMUTATOR <- mutation.landscape$tumour[mutation.landscape$mutator_phenotype %in% c('HRD','MSI')]
HYPERMUTATOR <- c(c("BTC_0025_Sn_M_526", "BTC_0025_Lv_M_526"), HYPERMUTATOR)

driver.landscape <- fread(file.path(data.dir, "/results/driver.landscape.txt"))
driver.landscape <-  driver.landscape %>% filter(gene %in% c("BAP1",'TP53')) %>% dcast(donor ~ gene)

predicted.w.score <- fread(file.path(data.dir, '/results/all.eCCS.txt') )

eCCSA <- unlist(predicted.w.score %>% filter(rna_class == 'CCS-A') %>% dplyr::select(Sample))
eCCSB <- unlist(predicted.w.score %>% filter(rna_class == 'CCS-B') %>% dplyr::select(Sample))

BAP1_tumours <- driver.landscape$donor[driver.landscape$BAP1 > 0 & driver.landscape$TP53 == 0 ]
tp53_tumours <- driver.landscape$donor[driver.landscape$BAP1 == 0 & driver.landscape$TP53 > 0 ]

```

## mutation load differences
Start with simply comparing TMB, accounting for hypermutator cases (MSI + HRD).

```{r}

mutation.landscape$driver[mutation.landscape$tumour %in% BAP1_tumours] <- 'BAP1'
mutation.landscape$driver[mutation.landscape$tumour %in% tp53_tumours] <- 'TP53'

mutation.landscape$rna_class[mutation.landscape$tumour %in% eCCSA] <- 'eCCS-A'
mutation.landscape$rna_class[mutation.landscape$tumour %in% eCCSB] <- 'eCCS-B'

mutation.landscape %>% filter(!is.na(rna_class)) %>% group_by( mutator_phenotype, rna_class) %>% tally()
tmb <- mutation.landscape %>% dplyr::select(tumour, driver, mutator_phenotype, rna_class, tmb, indel_count, snv_count, sv_count)

tmb$is.mutator <- F
tmb$is.mutator[tmb$tumour %ni% HYPERMUTATOR ] <- T

tmb.plot <- 
ggplot(
  tmb %>% filter(!is.na(rna_class) & is.mutator),
  aes(x = rna_class, y=log(tmb), color=rna_class)
) +

  ggbeeswarm::geom_quasirandom(
    aes(group = rna_class),
    width = 0.2
  ) +
  
  geom_boxplot(
    aes(
      
      group = interaction(rna_class)
    ),
    width = 0.45,
    alpha = 0.5,
    outlier.shape = NA,   # hide outliers since points are shown
  ) +
  
  # Color/fill for the two classes
  scale_color_manual(values = c('eCCS-A' = "#04597C", 'eCCS-B' = "#EDA71A"), na.value = 'grey85') +
  scale_fill_manual(values  = c('eCCS-A' = "#04597C", 'eCCS-B' = "#EDA71A"), na.value = 'grey85') +
  
  theme_bw(base_size=10) +
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "mm"),
    
    panel.grid        = element_blank(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    strip.background  = element_blank(),
    legend.position   = "top",
    # axis.text.x       = element_text(angle = 30, hjust = 1)
  ) +
  guides(color='none')+
  scale_x_discrete(limits = c("eCCS-B", "eCCS-A"),labels=c("B",'A')) +
  labs(x = 'CCS', y = "Tumor Mutation Burdern (log)", color = NULL, fill = NULL)

tmb.plot

ggsave(file.path(data.dir, "/results/fig/FIG5A.tmb.svg"),plot =
    tmb.plot     
   , device = "svg", width = 25, height = 40, units = 'mm', dpi= 72 )

wilcox.test(tmb$tmb[tmb$tumour %in% eCCSA  ],
            tmb$tmb[tmb$tumour %in% eCCSB  ])

median(tmb$tmb[tmb$tumour %in% eCCSB ]) / median(tmb$tmb[tmb$tumour %in% eCCSA  ])

wilcox.test(tmb$tmb[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ],
       tmb$tmb[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ])

median(tmb$tmb[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ]) / median(tmb$tmb[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ])


wilcox.test(tmb$snv_count[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ],
            tmb$snv_count[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ])

median(tmb$snv_count[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ]) / median(tmb$snv_count[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ])

wilcox.test(tmb$indel_count[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ],
            tmb$indel_count[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ])

median(tmb$indel_count[tmb$tumour %in% eCCSB & tmb$tumour %ni% HYPERMUTATOR ]) / median(tmb$indel_count[tmb$tumour %in% eCCSA & tmb$tumour %ni% HYPERMUTATOR ])

```


## HRD and MSI
Compare the rates of HRD and MSI in the CCS

```{r}

mutator.groups <- tmb %>%
  dplyr::filter(!is.na(rna_class)) %>%
  dplyr::count(rna_class, mutator_phenotype) %>%
  complete(rna_class, mutator_phenotype, fill = list(n = 0)) %>%
  group_by(rna_class) %>%
  mutate(total = sum(n),
         relative_freq = n / total) %>%
  rowwise() %>%
  mutate(ci = list(binom.test(n, total)$conf.int)) %>%
  unnest_wider(ci, names_sep = "_") %>%
  ungroup() %>%
  filter(mutator_phenotype != 'none')


this.phenotype = 'HRD'
these.counts <- mutator.groups %>% filter(mutator_phenotype == this.phenotype)
this.res <- these.counts[,c(3:4)] %>% fisher.test(conf.int = T)

hrd.fish <- cbind.data.frame(
  'mut'=  this.phenotype,
  'p'=this.res$p.value,
  'ci_low'=this.res$conf.int[1],
  'ci_hi'=this.res$conf.int[2]
)

this.phenotype = 'MSI'
these.counts <- mutator.groups %>% filter(mutator_phenotype == this.phenotype)
this.res <- these.counts[,c(3:4)] %>% fisher.test(conf.int = T)

mut.fish<- rbind.data.frame(
  hrd.fish,
  cbind.data.frame(
    'mut'=  this.phenotype,
    'p'=this.res$p.value,
    'ci_low'=this.res$conf.int[1],
    'ci_hi'=this.res$conf.int[2]
  )
)

mutator.tally.plot <- 
      ggplot(mutator.groups , aes(x=rna_class, y=relative_freq*100, fill=rna_class)) + 
          facet_grid(.~mutator_phenotype, space='free',scales='free')+
         geom_hline(yintercept = 0, color='grey')+
        geom_errorbar(aes(ymin = ci_1*100, 
                          ymax = ci_2*100), width = 0, color='black') +
        
          geom_point(aes(color=rna_class),stat='identity', position = position_dodge()) +
        
          theme_bw(base_size=10)+
          scale_color_manual(values = c('eCCS-A'="#04597C",'eCCS-B'= "#EDA71A"), na.value = 'grey90') +
          scale_x_discrete(limits = c("eCCS-B", "eCCS-A"),labels=c("B",'A')) +
         
          theme(panel.grid = element_blank(), 
               legend.position = 'bottom',
               panel.background = element_blank(),
               plot.background = element_blank(),
               plot.margin = unit(c(0, 0, 0, 0), "mm"),
               
               strip.background = element_blank()

            )+
          labs(y='% of samples', fill='',x='CCS') +


          guides(color = 'none',fill='none')  
       
mutator.tally.plot

ggsave(file.path(data.dir, "/results/fig/FIGS4g.mutator.tally.svg"),plot =
       mutator.tally.plot  
       
       , device = "svg", width = 35, height = 60, units = 'mm', dpi=72 )

mut.fish

```


## variance model
Here we compare variance in TMB explained by CCS vs primary tumor anatomical location

```{r}

cohort.info <- fread(file.path(data.dir,"/results/samples.info.txt")) %>%
  dplyr::select(Sample, location)

mutation.landscape <- full_join(mutation.landscape, cohort.info, by=c('tumour'='Sample'))
mutation.landscape$tmb.log <- log(mutation.landscape$tmb)

vars <- c("tmb.log", "location", "rna_class")
d <- na.omit(mutation.landscape[ , ..vars])

full <- lm(tmb.log ~ location + rna_class, data = d)

Anova(full, type = 2)


```

## Mutational Signatures

```{r}

sbs.sigs <- fread(file.path(data.dir,'/results/SBS.sig.txt'))
ncol(sbs.sigs) - 1
dbs.sigs <- fread(file.path(data.dir,'/results/DBS.sig.txt'))
ncol(dbs.sigs) - 1
id.sigs <- fread(file.path(data.dir,'/results/indel.sig.txt'))
ncol(id.sigs) - 1

all.sigs <- full_join(sbs.sigs, dbs.sigs, by=c('Samples'='Samples')) %>% full_join(id.sigs, by=c('Samples'='Samples'))

all.sigs <- all.sigs %>% filter(Samples %ni% HYPERMUTATOR)
all.sigs[is.na(all.sigs)] <- 0
all.sigs.frac.raw <- as.data.frame(t(apply(all.sigs[,-1], 1, function(x) x / sum(x))))

all.sigs.frac <- cbind.data.frame('Samples'=all.sigs$Samples, all.sigs.frac.raw)
all.sigs.frac.melt <- all.sigs.frac %>% reshape2::melt(id.vars='Samples')

common.sigs <- all.sigs.frac.melt %>% group_by(variable) %>% summarise(mean_perc = mean(value)) 
rare.sigs <- all.sigs.frac.melt %>% filter(value > 0) %>% group_by(variable) %>% tally %>% arrange(desc(n))

sig.distance <- as.matrix(dist(all.sigs.frac.raw))
rownames(sig.distance) = colnames(sig.distance) = all.sigs$Samples

cluster.sigs = F
if(cluster.sigs){
  sig.pvclust.result <- pvclust(sig.distance, method.hclust = 'complete' ) 
  plot(sig.pvclust.result )
  all.sigs.frac.melt$donor <- factor(all.sigs.frac.melt$Samples, levels = sig.pvclust.result$hclust$labels[sig.pvclust.result$hclust$order])
  
}

all.sigs.frac.melt <- left_join(all.sigs.frac.melt, predicted.w.score, by=c('Samples'='Sample'))

all.sigs.frac.melt <- left_join(all.sigs.frac.melt, sigprofiler.meta, by=c('variable'='sig'))
all.sigs.frac.melt$sig <- factor(all.sigs.frac.melt$variable, levels = common.sigs$variable[order(common.sigs$mean_perc)])
all.sigs.frac.melt$Samples.f <- factor(all.sigs.frac.melt$Samples, levels = predicted.w.score$Sample[order(predicted.w.score$confidence.polarized)])

all.sigs.frac.melt$aetiology_short[all.sigs.frac.melt$aetiology_short %in% c('','Unknown')] <- NA

sig_oncoplot <- 
          plot_grid(
            
            ggplot(all.sigs.frac.melt %>% filter(value > 0 & sig %in% rare.sigs$variable[rare.sigs$n >1] & !is.na(Samples.f)), aes(x=Samples.f, y='eCCS', fill=rna_class)) +
              geom_tile() + theme_minimal(base_size=10) + 
              
              scale_fill_manual(values = c('CCS-A'="#04597C",'CCS-B'= "#EDA71A"), na.value = 'grey') +
              guides(fill='none',size='none') +
              theme(panel.grid = element_blank(), 
                    plot.background = element_blank(), 
                    panel.background = element_blank(),
                    axis.text.x = element_blank(), 
                    axis.title = element_blank()) 
            
            ,NULL,
            
            ggplot(all.sigs.frac.melt %>% filter(value > 0 & sig %in% rare.sigs$variable[rare.sigs$n >1] & !is.na(Samples.f)), aes(x=Samples.f, y=sig, size=value*100)) +
              geom_point(aes(color=aetiology_short)) + theme_minimal(base_size=10) + 
              scale_size(range = c(0.5,2))  +
              scale_color_manual(values =  pnw_palette(name="Spring", n=length(unique(all.sigs.frac.melt$aetiology_short))), 
                                 na.value = 'lightgrey') +
              theme(panel.grid = element_blank(), 
                    axis.text.y = element_text(size = 6),
                    plot.background = element_blank(), 
                    axis.title = element_blank(),
                    panel.background = element_blank(),axis.text.x = element_blank())+
              labs(size='% of SSMs',color='Etiology')
            
            ,rel_heights = c(0.1,-0.02,0.9 ), ncol = 1,align = 'vh', axis = "rlbt")
     
sig_oncoplot   
     
ggsave( file.path(data.dir, "/results/fig/FIGS5.sig_oncoplot.svg"), plot =
          sig_oncoplot
        , device = "svg",width = 210, height = 90, units='mm', dpi=72 )

```

## enrichment analysis
Find significant biases in appeared of mutational signatures in each subtype, based on methods from [Diaz-Gay 2025](https://doi.org/10.1038/s41586-025-09025-8). A signature has to be in greater than 30% of samples to be considered for bias.

```{r}

enrich.sigs <- all.sigs %>% filter(Samples %ni% HYPERMUTATOR & Samples %in% c(eCCSA, eCCSB))

threshold <- 0.3

appearance.sigs <- 
  enrich.sigs %>%
  select_if(~ mean(. == 0) > threshold)
appearance.sigs[appearance.sigs > 0] <- 1

appearance.sigs <- appearance.sigs %>%
  select_if(~ sum(.) != 0)

load.sigs <- 
  enrich.sigs %>%
  select_if(~ mean(. == 0) < threshold)


load.sigs.transformed <- load.sigs[,-1] %>%
  mutate(across(everything(), ~ ifelse(. < median(.), 0, 1)))

load.sigs.transformed <- load.sigs.transformed %>%
  select_if(~ !all(. == 1))


enrich.binarized <- cbind.data.frame('Samples'=enrich.sigs$Samples, appearance.sigs, load.sigs.transformed)

write.table(
  enrich.binarized,
  file = file.path(data.dir, "/results/sig.enrich.binarized.txt"),
  row.names = FALSE, quote = FALSE, sep = "\t"
)

enrich.binarized <- fread(file.path(data.dir, "/results/sig.enrich.binarized.txt"))

enrich.binarized$rna_class <- NA
enrich.binarized$rna_class[enrich.binarized$Samples %in% eCCSA] <- 0
enrich.binarized$rna_class[enrich.binarized$Samples %in% eCCSB] <- 1

enrich.binarized.classed <- enrich.binarized %>% filter(!is.na(rna_class))
enrich.binarized.classed.melt <- enrich.binarized.classed %>% reshape2::melt(id.vars=c('Samples', 'rna_class'))

refit.sbs.models = F
if(refit.sbs.models){
  formula <- as.formula(paste("sig ~ rna_class"))
  
  first.model = T
  for (this.sig in unique(enrich.binarized.classed.melt$variable)) {
    
    enrich.binarized.classed.tmp <- enrich.binarized.classed %>% dplyr::select(all_of(c(this.sig,'rna_class')))  
    names(enrich.binarized.classed.tmp)[1] <- 'sig'
    
    driver.tally <- enrich.binarized.classed.tmp %>%
      group_by(rna_class) %>%
      summarise(non_zero_count = sum(sig != 0)) %>%
      filter(non_zero_count > 1)
    
    if(nrow(driver.tally) > 1){
      model <- glm(formula, data = enrich.binarized.classed.tmp, family = binomial)
      
      tidy_model <- tidy(model, exponentiate = TRUE, conf.int = TRUE)
      tidy_model <- tidy_model[tidy_model$term == 'rna_class',]
      #tidy_model$term <- this.sig
      
      model.data <- cbind.data.frame(
        'sig'=this.sig,
        'coefficients'=log(tidy_model$estimate),
        'ci.lower'=tidy_model$conf.low,
        'ci.upper'=tidy_model$conf.high,
        'p'=tidy_model$p.value,
        'model'='classic'
      )
      
      if(first.model){enrich.models <- model.data; first.model=F}else{
        enrich.models <- rbind.data.frame(enrich.models, model.data)
      }
    } else {
      firths.model <- logistf(formula, data = enrich.binarized.classed.tmp, family = binomial)
      firths.model.summary <- summary(firths.model)
      
      model.data <- cbind.data.frame(
        'sig'=this.sig,
        'coefficients'=firths.model.summary$coefficients[2],
        'ci.lower'=firths.model.summary$ci.lower[2],
        'ci.upper'=firths.model.summary$ci.upper[2],
        'p'=firths.model.summary$prob[2],
        'model'='firth'
      )
      
      if(first.model){enrich.models <- model.data; first.model=F}else{
        enrich.models <- rbind.data.frame(enrich.models, model.data)
      }
    }
    
    
  }
  
  write.table(
    enrich.models,
    file = paste0(data.dir, "/results/sig.bias_models.txt"),
    row.names = FALSE, quote = FALSE, sep = "\t"
  )
}

enrich.models <- fread(paste0(data.dir, "/results/sig.bias_models.txt"))


enrich.models$significant <- 'F'
enrich.models$significant[enrich.models$p < (0.05)] <- 'T'
enrich.models$significant[enrich.models$p < (0.05/nrow(enrich.models))] <- 'FDR'

enrich.models$mut_type <- NA
enrich.models$mut_type[enrich.models$sig %in% names(sbs.sigs)] <- 'SBS'
enrich.models$mut_type[enrich.models$sig %in% names(dbs.sigs)] <- 'DBS'
enrich.models$mut_type[enrich.models$sig %in% names(id.sigs)] <- 'ID'
enrich.models %>% group_by(mut_type) %>% tally()
enrich.models %>% group_by(mut_type, significant) %>% tally()

enrich.models <- left_join(enrich.models, sigprofiler.meta, by=c('sig'='sig'))

enrich.models.sub <- enrich.models %>% filter(mut_type != "DBS")
enrich.models.sub$sig <- gsub('SBS','',enrich.models.sub$sig)
enrich.models.sub$sig <- gsub('ID','',enrich.models.sub$sig)
enrich.models.sub$sig <- gsub('DBS','',enrich.models.sub$sig)


sig.enrichment.dot.plot <-
   ggplot(enrich.models.sub , aes(x=-coefficients, y=-log(p))) + 
         geom_vline(xintercept = 0, color='grey') + 
         #  geom_hline(yintercept = -log(0.05), color='grey50',linetype='dashed') +
         geom_hline(yintercept = -log((0.05/nrow(enrich.models.sub))), color='grey80',linetype='dotted') +
         
         geom_point(size=2, color='grey80', aes(shape=mut_type)) + 
         # geom_point(data = ~ subset(., p < (0.05/nrow(enrich.models))), aes(color=aetiology_short), size=3) + 
           geom_point(data = ~ subset(., p < 0.05/nrow(enrich.models.sub) ), aes(color=aetiology_short, shape=mut_type), size=2) +
         geom_point(data = ~ subset(., p < 0.05/nrow(enrich.models.sub) ), aes(color=aetiology_short, shape=mut_type), size=5) +
         
         geom_text(data = ~ subset(., p < 0.05/nrow(enrich.models.sub) ),
                   aes(label = sig)
         )+
         
         theme_bw(base_size=10) +
         scale_color_manual(values =  pnw_palette(name="Spring", n=length(unique(enrich.models.sub$aetiology_short[enrich.models.sub$p < 0.05/nrow(enrich.models.sub) ]))), 
                            na.value = 'lightgrey') +
         scale_shape_manual(values = c(1,16)) +
         guides(color='none',fill='none',shape='none')+
         theme(panel.grid = element_blank(), 
               plot.margin = unit(c(0, 0, 0, 0), "mm"),
               legend.position = 'top', 
               panel.background = element_blank(), 
               plot.background = element_blank()) + 
         
         labs(x='Log Odds Ratio', y='p-value (-log10)', color='',shape='') +  
  
  xlim(-max(abs(enrich.models.sub$coefficients)),max(abs(enrich.models.sub$coefficients)))
       
sig.enrichment.dot.plot       

ggsave(file.path(data.dir, "/results/fig/FIG5b.sig.enrichment.dot.svg"),plot =
         sig.enrichment.dot.plot
       , device = "svg", width = 60, height = 50, units = 'mm' , dpi= 72)

enrich.models.sub %>% filter(significant == "FDR")

```

## PCA of mutational signatures
Having found signatures that show a bias between the subtypes, we ask whether these differences are (1) important and (2) more important than anatomical location. To find out, we turn to a principal component analysis.

```{r}

allOfit_pca <- PCA(all.sigs.frac.raw, graph = F)

allOfit_pca_labs <- 
  cbind.data.frame(
    Sample=all.sigs$Samples,
    allOfit_pca$ind$coord
       )

samples.info <- fread(file.path(data.dir,"/results/samples.info.txt")) %>%
  dplyr::select(Sample, Site_of_tissue, location, tissue_before_chemo)

allOfit_pca_labs <- full_join(allOfit_pca_labs, samples.info, by=c('Sample'='Sample'))
allOfit_pca_labs$tissue_before_chemo[allOfit_pca_labs$tissue_before_chemo == 'TRUE'] <- 'treatment naive'
allOfit_pca_labs$tissue_before_chemo[allOfit_pca_labs$tissue_before_chemo == 'FALSE'] <- 'post-treatment'

allOfit_pca$eig[1:2,]

wilcox.test(allOfit_pca_labs %>% filter(tissue_before_chemo == 'treatment naive' & !is.na(Dim.1)) %>% dplyr::select(Dim.1)  %>% unlist(),
            allOfit_pca_labs %>% filter(tissue_before_chemo == 'post-treatment' & !is.na(Dim.1)) %>% dplyr::select(Dim.1) %>% unlist()  )

allOfit_pca_labs$class[allOfit_pca_labs$Sample %in% eCCSA] <- 'eCCS-A'
allOfit_pca_labs$class[allOfit_pca_labs$Sample %in% eCCSB] <- 'eCCS-B'

wilcox.test(allOfit_pca_labs %>% filter(class == 'eCCS-B' & !is.na(Dim.2)) %>% dplyr::select(Dim.2)  %>% unlist(),
            allOfit_pca_labs %>% filter(class == 'eCCS-A' & !is.na(Dim.2)) %>% dplyr::select(Dim.2) %>% unlist()  )

SIGPCA.plot <- 
         plot_grid(

           ggplot(allOfit_pca_labs %>% filter(!is.na(class)),aes(x=Dim.2)) + 
             geom_histogram(position=position_dodge(), aes(y=..density.., fill=class)) +
             scale_color_manual(values = c('treatment naive'='lightgrey',
                                           'post-treatment'='cornflowerblue',
                                           'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                           'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                                na.value = 'white') +
             scale_fill_manual(values = c('treatment naive'='lightgrey',
                                          'post-treatment'='cornflowerblue',
                                          'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                          'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                               na.value = 'white') +
             theme_minimal(base_size=10) + 
             guides(color='none',fill='none')+
             theme(
               panel.grid = element_blank(),
               axis.title= element_blank(),
               axis.text= element_blank(),
               plot.background = element_blank(),
               panel.background = element_blank())+ 
             xlim(min(allOfit_pca_labs$Dim.2, na.rm=T)-0.1, max(allOfit_pca_labs$Dim.2, na.rm=T)+0.1 )
           
           ,NULL,NULL,
           NULL,NULL,NULL,
           
           ggplot(data=allOfit_pca_labs %>% filter(!is.na(class)),aes(y=Dim.1,x=Dim.2, color=class)) +
             
             geom_point(size=1.2,   shape=1, aes(color=class), stroke = 0.8) + 
             geom_point(size=0.5, shape=16, aes(color=tissue_before_chemo)) + 
             
             
             scale_shape_manual(values = c(3,4)) +
             
             labs(
               y= "Mutational Signature PC1",
               x= "Mutational Signature PC2",
               
               color="Subtype") +
             guides(color='none',fill='none',shape='none')+
             theme_bw(base_size=10) +
             
             scale_color_manual(values = c('treatment naive'='lightgrey',
                                           'post-treatment'='cornflowerblue',
                                           'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                           'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                                na.value = 'white') +
             scale_fill_manual(values = c('treatment naive'='lightgrey',
                                          'post-treatment'='cornflowerblue',
                                          'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                          'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                               na.value = 'white') +
             theme(plot.background = element_blank(),
                   panel.grid = element_blank(),
                   panel.background = element_blank())  + 
             xlim(min(allOfit_pca_labs$Dim.2, na.rm=T)-0.1, max(allOfit_pca_labs$Dim.2, na.rm=T)+0.1 )+
             ylim(min(allOfit_pca_labs$Dim.1, na.rm=T)-0.3, max(allOfit_pca_labs$Dim.1, na.rm=T)+0.1 )
           
           ,
           NULL,
           ggplot(allOfit_pca_labs %>% filter(!is.na(tissue_before_chemo)),aes(y=Dim.1)) + 
             #geom_density(aes(color=tissue_before_chemo), linewidth = 0.2) +
             geom_histogram(position=position_dodge(), aes(x=..density.., fill=tissue_before_chemo)) +
             
             scale_color_manual(values = c('treatment naive'='lightgrey',
                                           'post-treatment'='cornflowerblue',
                                           'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                           'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                                na.value = 'white') +
             scale_fill_manual(values = c('treatment naive'='lightgrey',
                                          'post-treatment'='cornflowerblue',
                                          'eCCS-A'= pnw_palette(name="Bay", n=30)[c(3)],
                                          'eCCS-B'=pnw_palette(name="Bay", n=30)[c(20)]), 
                               na.value = 'white') +
             
             theme_minimal(base_size=10)+
             labs(fill='Platinum\nChemotherapy')+
             theme(plot.background = element_blank(),
                   panel.grid = element_blank(),
                   axis.title= element_blank(),
                   axis.text= element_blank(),
                   panel.background = element_blank()) + 
             
             ylim(min(allOfit_pca_labs$Dim.1, na.rm=T) -0.3, max(allOfit_pca_labs$Dim.1, na.rm=T)+0.1 ) +
             guides(color='none', fill='none')
           
           ,   rel_heights = c(0.35,-0.22,0.65), rel_widths = c(0.65,-0.18,0.35),ncol = 3, align = 'hv', axis = "rlbt")
       
SIGPCA.plot 

ggsave(file.path(data.dir,"/results/fig/FIG5c.SIGPCA.svg"),plot =
         SIGPCA.plot
       , device = "svg", height = 40, width = 65 , units='mm')


```

## RDA

```{r}

predicted.w.score <- fread(file.path(data.dir,'/results/all.eCCS.txt') )
predicted.w.score$rna_class[is.na(predicted.w.score$rna_class)] <- 'Other'

sample.info <- fread(file.path(data.dir,"/results/samples.info.txt"))

sample.typed <- left_join(sample.info, predicted.w.score)

sample.filt <- sample.typed %>% filter(!is.na(location) & 
                                          !is.na(rna_class) & 
                                          !is.na(Site_of_tissue))

location <- sample.filt$location
names(location)  <- sample.filt$Sample
rna_class <- sample.filt$rna_class
names(rna_class)  <- sample.filt$Sample
site <- sample.filt$Site_of_tissue
names(site)  <- sample.filt$Sample

these.samples <- intersect(all.sigs$Samples, sample.filt$Sample)
 
location <- location[these.samples]
rna_class <- rna_class[these.samples]
site <- site[these.samples]

# RDA identical ordination to PCA when centered/scaled
rownames(all.sigs.frac.raw) <- all.sigs$Samples
all.sigs.frac.filt <- all.sigs.frac.raw[these.samples,]

pca_rda <- rda(all.sigs.frac.filt, scale = TRUE)

eig.val <- (eigenvals(pca_rda) / pca_rda$tot.chi)*100

r.tab = NULL
for(i in c(1:5)){

  fit2_f1 <- envfit(pca_rda, location, permutations = 999, choices = i:i)
  fit2_f2 <- envfit(pca_rda, rna_class, permutations = 999, choices = i:i)
  fit3_f3 <- envfit(pca_rda, site, permutations = 999, choices = i:i)

  this.r.tab <- rbind.data.frame ( 
    cbind('i'=as.numeric(i),
               'f'='anatomy',
               'e'=as.numeric(eig.val[i]),
               'r'=unlist(fit2_f1$factors$r),
               'p'=unlist(fit2_f1$factors$pvals)
    ),
    cbind('i'=as.numeric(i),
        'f'='CCS',
        'e'=as.numeric(eig.val[i]),
        'r'=unlist(fit2_f2$factors$r),
        'p'=unlist(fit2_f2$factors$pvals)
    ),
    cbind('i'=as.numeric(i),
        'f'='site',
        'e'=as.numeric(eig.val[i]),
        'r'=unlist(fit3_f3$factors$r),
        'p'=unlist(fit3_f3$factors$pvals)
    )
  )
  
  if(is.null(r.tab)){
    r.tab <- this.r.tab
  } else{
    r.tab <- rbind.data.frame(r.tab,
                              this.r.tab)
  }
  
}

r.tab$p <- as.numeric(r.tab$p)
r.tab$r <- as.numeric(r.tab$r)
r.tab$e <- as.numeric(r.tab$e)

r.tab$p.sig <- NA
r.tab$p.sig[r.tab$p < 0.05] <- r.tab$p[r.tab$p < 0.05]

r.tab$pc.i <- paste0('PC',r.tab$i)
r.tab$pc.i <- factor(r.tab$pc.i, levels=rev(paste0('PC',c(1:5)) ))
r.tab$factor <- factor(r.tab$f, levels=rev(c('CCS','anatomy','site')))


r.tab$var = r.tab$r * r.tab$e
var.sum <- r.tab %>% filter(!is.na(p.sig)) %>% group_by(f) %>% summarise(var_sum = sum(var))
var.sum[1,2] / var.sum[2,2]

r.tab

exp.rda.stack.plot <- 
ggplot(
  r.tab %>% filter(p < 0.05 & factor != "site"),
  aes(y = factor, x = r * e)
) +
  geom_bar(aes(fill = pc.i), stat = "identity") +
  geom_text(
    aes(label = pc.i), color='white',
    position = position_stack(vjust = 0.5), size=2,
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.margin = unit(c(0, 2, 0, 2), "mm"),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) + labs(x='Variance explained (%)')+
  scale_fill_manual(
    values = c(
      PC1 = "black",
      PC2 = "grey20",
      PC3 = "grey40",
      PC4 = "grey60",
      PC5 = "grey80"
    ),
    na.value = "cornflowerblue"
  ) +
  expand_limits(x = max(r.tab$r * r.tab$e, na.rm = TRUE) * 1.1) + 
  guides(fill='none')

exp.rda.stack.plot

ggsave(file.path(data.dir,"/results/fig/FIG5d.sig_rda.stack.svg"),plot =
  exp.rda.stack.plot       
, device = "svg", height = 15, width = 55 , units='mm')



```
